---
title: "Getting started: which workflow do I need?"
author: "Tom Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: biomasslmb.json
vignette: >
  %\VignetteIndexEntry{Getting started: which workflow do I need?}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`biomasslmb` covers the path from search engine output to a list of differentially abundant proteins. This article is a map of that path: it explains what the other vignettes assume, which of them applies to your experiment, and which decisions along the way are worth thinking about rather than accepting a default. It contains almost no analysis itself — each stage links to the vignette that works through it in full.

# What the package assumes


**You have search engine output, not raw spectra.** The starting point is a PSM, peptide or precursor level table from Proteome Discoverer, MaxQuant, Spectronaut or DIA-NN. Everything upstream of that — the search itself, FDR control at the spectrum level, the choice of sequence database — happens elsewhere.

**Everything lives in a `QFeatures` object.** Each processing step adds a *new* assay rather than overwriting the previous one, and assay links record which features contributed to which. A finished object therefore holds the whole history of the analysis, from the raw features to the protein-level quantification, and you can always step back to ask where a protein-level value came from. This is why the vignettes accumulate assays with names like `psms_filtered_sn` and `peptides_for_summarisation` rather than reassigning one matrix.

**The destination is protein-level quantification and statistical testing.** Peptide- and site-level analyses are possible and are covered separately, but the default path summarises to protein level before testing.

# How to read the code in these vignettes

To clarify which functionality is provided by which package, the vignettes write calls as `package::function`. For your own code, there is no need to specify the package unless you want to maintain this clarity.

# Which material to refer to?

Which vignette you need depends on two independent things: how the data were acquired, and what kind of experiment it was.

**Acquisition type** determines the QC and filtering you need, because the artefacts differ. It selects one of three vignettes:

| Acquisition | Input | Vignette |
|---|---|---|
| TMT | PSM-level (PD or MaxQuant) | [TMT PSM QC and summarisation](TMT_PSM_QC_Summarisation.html) |
| LFQ-DDA | Peptide groups (PD or MaxQuant) | [LFQ-DDA peptide QC and summarisation](LFQ_DDA_Peptide_QC_Summarisation.html) |
| LFQ-DIA | Precursors (DIA-NN or Spectronaut) | [LFQ-DIA precursor QC and summarisation](LFQ_DIA_Precursor_QC_Summarisation.html) |

Each of those works through one straightforward experiment end to end. A TMT design too large for a single plex needs the extra step covered in [multi-plex TMT](TMT_multiplex.html).

**Experimental design** determines how the resulting quantification should be normalised and tested. A whole-proteome comparison, an enrichment experiment (IP, BioID, TurboID) and a PTM experiment ask different things of the same protein-level matrix:

- **Whole proteome.** Most proteins are unchanged between conditions, missing values usually mean the protein was present but not detected, and a two-sided test against the other condition is what you want. The [data exploration and statistical testing](exploration_and_statistical_testing.html) vignette works through this case.
- **Enrichment.** Most of the quantified proteome may be bait-dependent, so the assumption that underpins median normalisation fails. Absence from the control is frequently real absence rather than failed detection, which changes what imputation means. Testing is often one-sided. See [enrichment designs](interactome_designs.html).
- **PTM.** Site-level changes are confounded with changes in the abundance of the underlying protein, so the enriched data need normalising against a matched measurement of the total proteome before the site-level result means anything. See [PTM site quantification](PTM_site_quantification.html).

The two axes are independent: a TMT IP and an LFQ-DIA whole-proteome comparison use different QC vignettes and different statistical treatments, and each combination occurs in practice.

All three of those designs ask the same underlying question — which proteins, or sites, differ in abundance between conditions — and that is the question every vignette here works through to a result. Many experiments ask something else of the same measurements: how abundance changes over time, how it responds to a dose or a temperature, how quickly a protein is made or destroyed. The stages below are not specific to differential abundance, and everything up to and including protein-level quantification applies unchanged to those designs: the annotation, the QC and filtering, the summarisation, the normalisation, and the reasoning about missing values are all the same work. What differs is the model fitted at the end. If your experiment is one of those, read the vignette for your acquisition type for stages 1 and 2, and treat the testing stage as the point where this material stops applying.

Both axes concern comparing the same protein between samples, which is what the quantification supports well. Comparing *different* proteins with each other — asking which of two proteins is present in greater amount — is a separate question, and the intensities do not answer it directly, because a peptide's signal depends at least as much on how readily it ionises as on how much of it there is. [Absolute quantification and iBAQ](absolute_quantification_iBAQ.html) covers the correction usually applied and what it does and does not achieve.

# The stages, in order

Analyses typically follow the same sequence regardless of which combination above applies.

**1. Protein annotation.** Retrieve UniProt annotations for the accessions in your data before anything else, and cache them, so that every later step can use gene names, protein descriptions and GO terms. `get_uniprot_details()` retrieves the annotations, `collapse_uniprot_details_multi_accession()` handles protein groups with several accessions, and `get_go_terms(expand_terms = TRUE)` retrieves GO terms with ancestor terms included. Retired accessions that no longer resolve are recovered with `get_uniparc_fallback()`. See [protein annotation](protein_annotation.html).

**2. QC, filtering and summarisation.** Remove contaminants and low-confidence identifications, log-transform, normalise, and summarise features to protein-level abundance. This is the acquisition-specific stage — use the vignette from the table above.

**3. Exploration and statistical testing.** Check that the experiment worked before testing it: confirm the bait or protein(s) of interest behaves as expected, look at sample correlations and `plot_pca()`, then apply a testability filter and test. The conventional filter requires at least two genuinely quantified replicates in at least one condition, and at least two values in both conditions of whichever assay is actually fitted — one rule when nothing is imputed, two when something is. See [data exploration and statistical testing](exploration_and_statistical_testing.html).

**4. Functional enrichment.** Test whether the differentially abundant proteins are enriched for particular GO terms, against the proteins that were tested rather than against the proteome. `get_enriched_go()` runs the test, correcting for the tendency of abundant proteins to be called significant more often via a goseq probability weighting function; `estimate_overrep()` adds an effect size, `remove_redundant_go()` collapses the parent and child terms that otherwise dominate the output, and `plot_go()` displays it. See [functional enrichment](functional_enrichment.html), and [ORA or GSEA?](ORA_vs_GSEA.html) for the threshold-free alternative.

# The decisions worth thinking about

Every analysis makes choices about summarisation, missing values, testing and normalisation. They are not independent of each other, and there is no single set of correct answers, so it is worth understanding what each one assumes.

## Summarisation, missing values and testing are one decision

The obvious reading is that you pick a summarisation method, then decide what to do about missing values, then choose a test. Some of the available approaches do work that way. Others do not, because they treat the missingness as something to be modelled during the test rather than resolved before it:

- **Separate steps.** Summarise with `robustSummary` or by summing, decide on imputation at the protein level, then test with `limma`. Each step is independent and inspectable. `limma` fits a linear model per protein, but it fits it to whatever values survive the earlier steps, and knows nothing about why the rest are absent.
- **limpa** couples all three. It takes peptide-level data, fits a detection probability curve relating detectability to abundance, produces protein-level estimates *with standard errors*, and carries that uncertainty into the linear model. It replaces summarisation, imputation and the model fit together.
- **missBayes** couples the last two. It takes unimputed protein-level quantification and models the missingness explicitly as part of the test. Note that it fits every protein, so a result can concern a protein that a `limma` analysis would never have tested at all.

Because a `QFeatures` object holds parallel assays, these are not mutually exclusive: you can summarise with `robustSummary` and with limpa into the same object and compare the outcomes.

## The conventional starting point depends on acquisition type

Missingness differs by acquisition type, and since summarisation choices are largely a response to missingness, so do the conventions:

| | Summarisation | Missing values | Testing |
|---|---|---|---|
| TMT | sum PSMs (`base::colSums`) | little within a plex; between plexes handled by bridge correction | `limma` |
| LFQ-DDA | `robustSummary` | the live decision | `limma`, or missingness-aware (limpa, missBayes) |
| LFQ-DIA | `robustSummary` | live, though less severe than DDA | `limma`, or missingness-aware (limpa, missBayes) |

Within a TMT plex, all samples are quantified in the same MS run, so after signal-to-noise filtering the data are close to complete, and reporter ion intensities are additive across the PSMs of a protein. Summing them is both simpler and better justified than a robust method designed to tolerate missing values. TMT's missing value problem appears *between* plexes, and bridge or pooled reference channels address it directly.

LFQ quantifies each sample in a separate run, so missing values are pervasive and are related to abundance. `robustSummary` and limpa's detection probability curve are both responses to that problem. Applying them to within-plex TMT data addresses a problem that is not there. The [summarisation methods](summarisation_methods.html) vignette compares `sum` and `robustSummary` directly, on the same data.

Three things should make you depart from the conventional path:

1. **Severity of missingness.** Diagnose it rather than assuming: `global_condition_miss_score()` separates missingness explained by abundance from missingness explained by condition, `condition_miss_score()` does the same per feature, and `plot_missing_upset()` and `plot_missing_SN()` show its structure. [Handling missing values](handling_missing_values.html) works through the diagnosis and every option that follows from it.
2. **Experimental design.** In an enrichment experiment, absence from the control is usually real rather than a detection failure. Discarding those features removes your strongest candidates outright; imputing them keeps the candidates but restates a presence/absence result as a fold change against an invented background, so its magnitude is set by the imputation parameters rather than by the data.
3. **Downstream statistical test.** `limma` tests contrasts between groups. It does not estimate a decay rate, a melting temperature, or an amplitude and phase, so it is the wrong tool for a degradation or turnover timecourse, thermal proteome profiling, a dose response, or a circadian experiment — these call for the package built for the question, `nls` for kinetics, a cosinor model or MetaCycle for rhythms, and so on, fitted directly to the protein-level quantification matrix. That is what should make you depart here: those tools take a matrix of abundances, and limpa and missBayes return a fitted model instead, whatever their merits for handling missing values.

## Normalisation: what do you assume is unchanged?

The normalisations used in these vignettes all work the same way: they take some set of proteins to be unchanged between samples, and remove whatever difference is observed in that set. They differ only in how explicit they are about which set that is.

- **Reference set.** `get_medians()` and `center_normalise_to_ref()` normalise to a subset you define: the total proteome for a phospho experiment, a spike-in, or a set of proteins independent of the bait in an enrichment experiment. The assumption is explicit, and you can check whether it holds. The [PTM site quantification](PTM_site_quantification.html) vignette works through the phospho case.
- **Bridge or pooled channel.** For multi-plex TMT, the reference is a channel present in every plex rather than a set of proteins. This is what removes plex effects, and the [multi-plex TMT](TMT_multiplex.html) vignette works through it.
- **All proteins.** `diff.median` normalisation assumes most proteins are unchanged. For a whole-proteome comparison between similar conditions this is reasonable. For an enrichment experiment or a strong perturbation it can be badly wrong, and it fails silently.

Not every normalisation is of this form. Quantile normalisation, for instance, nominates no set at all: it replaces each sample's ranked abundances with the average across samples at that rank, so every sample ends up with an identical distribution. The assumption is correspondingly stronger — that the true distribution of abundances is the same in every sample, and any difference in its shape is technical — and where that fails it fails as quietly as `diff.median` does.

# Example data

The package ships processed `QFeatures` objects for each acquisition type, each produced by the corresponding vignette, so you can experiment with a realistic object without running the QC first.

```{r, message=FALSE}
library(QFeatures)
library(biomasslmb)
```

The assay names record the processing each one has been through:

```{r}
lapply(list(tmt_qf = tmt_qf, lfq_qf = lfq_qf, dia_qf = dia_qf), names)
```

A `QFeatures` object also carries the experimental design in its `colData`, which is what the exploration and testing functions use:

```{r}
tmt_qf
```

To start from raw search engine output instead, the corresponding input files are in `extdata`:

```{r}
dir(system.file("extdata", package = "biomasslmb"))
```

`tmt_qf_mq` is the same TMT pipeline run from MaxQuant rather than Proteome Discoverer output, and `psm_tmt_clock`, `psm_tmt_per2_mq`, `psm_tmt_2plex` and their design tables provide PSM-level starting points with accompanying experimental designs. `psm_tmt_2plex` spans two TMT plexes, each with a pooled bridge channel.

# Where to go next

If you are starting an analysis, go to the vignette for your acquisition type and work forwards, then on to [data exploration and statistical testing](exploration_and_statistical_testing.html) and [functional enrichment](functional_enrichment.html). If you are trying to decide between two ways of doing something, the [summarisation methods](summarisation_methods.html) vignette compares summarisation approaches directly, the testing vignette covers the testing stage for both a whole-proteome and an enrichment design, and [comparing search engines](search_engine_comparison.html) asks how much of a result depends on the pipeline that produced it.

Four articles cover steps that look the same whether or not they are working, which is what makes them worth reading before you trust a result rather than after:

- [Peptides are not proteins](gotcha_peptide_to_protein.html) — the inference step between what the instrument measured and what the output file reports, and which proteins it is unreliable for.
- [Batch effects](gotcha_batch_effects.html) — how much of the variance in a multi-plex TMT experiment is the plex rather than the biology, and what an uncorrected analysis costs you. Whether that cost is false positives or lost power depends on how the conditions are distributed across plexes.
- [Contaminants and protein FDR](gotcha_contaminants_and_FDR.html) — three overlapping defences against contaminants, which of them is actually doing the work, and what `filter_by_protein_fdr()` removes besides low-confidence proteins.
- [Technical replicates are not biological replicates](gotcha_technical_replicates.html) — what happens to a result when repeat acquisitions of one sample are tested as though they were independent samples.

# Getting help

These vignettes describe the common cases, and many experiments are not common cases. If you are unsure which route applies to your data, or you have a design the vignettes do not cover, please get in touch with Tom Smith (tsmith@mrc-lmb.cam.ac.uk) rather than guessing — it is far easier to help at the design or analysis stage than to unpick a decision afterwards.

This applies in particular to the designs that end in a model other than a comparison between conditions: thermal proteome profiling, protein degradation and turnover measured by pulse labelling, circadian and other timecourse experiments, dose responses, and clustering of temporal profiles. The quantification these need is produced by the vignettes here, but the analysis that follows is specific enough to each experiment that it is worth discussing directly rather than adapting the testing vignettes to it. Please get in touch, ideally before the samples are run, since these designs constrain the number of conditions, the labelling scheme and the choice of acquisition in ways that are impossible to fix afterwards.
