---
title: "Data exploration and statistical testing"
author: "Tom Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: biomasslmb.json
vignette: >
  %\VignetteIndexEntry{Data exploration and statistical testing}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Once protein-level abundances have been QCed, filtered and summarised, the next step is to explore the data with respect to the experimental design, then test for differential abundance between conditions. Below, we work through that process twice, using datasets carried over from the QC vignettes: first with `dia_qf` (Part A), an LFQ-DIA whole-proteome comparison of blood plasma from MPXV-infected and healthy individuals; then with `tmt_qf_mq` (Part B), a TMT immunoprecipitation comparing a bait pulldown (`IP`) against a `Control` pulldown. The two designs call for different choices at the statistical testing stage, which we highlight in Part B.

# Load required packages

```{r, message=FALSE}
library(QFeatures)
library(biomasslmb)
library(ggplot2)
library(tidyr)
library(dplyr)
library(limma)
```

# Part A: LFQ-DIA whole-proteome comparison

`dia_qf` is the blood plasma dataset from the [LFQ-DIA QC](LFQ_DIA_Precursor_QC_Summarisation.html) vignette [@Wang2022]: a case series comparing MPXV infection, COVID-19 and healthy controls, summarised to protein level with `robustSummary`. The clinical group of each sample is in the `group` column. Here we focus on the MPXV-vs-control comparison, so we restrict to those samples up front.

```{r}
names(dia_qf)
knitr::kable(as.data.frame(colData(dia_qf))[, c('group', 'sex', 'age.group')])
```

```{r}
dia_prot <- dia_qf[, colData(dia_qf)$group %in% c('control', 'MPXV')]
```

## Sample correlation

A pairwise correlation heatmap is a quick way to check that replicates agree with each other more than they agree with samples from the other condition. `plot_cor_samples` computes the Spearman correlation between samples for a given assay and plots it as a heatmap.

```{r, fig.height=5, fig.width=5}
plot_cor_samples(dia_prot, 'protein')
```

## Principal component analysis

`plot_pca` performs a PCA on a given assay and plots two of the components, optionally colouring points by a `colData` column. `robustSummary` doesn't require complete peptides, so `dia_qf[['protein']]` retains some missing values (`r round(100*mean(is.na(assay(dia_qf[['protein']]))), 1)` % of the matrix). `stats::prcomp` can't handle missing values, so instead of the default `allowing_missing=FALSE` (which would silently drop every protein with any missing value via `filterNA`), we set `allowing_missing=TRUE`, which uses `pcaMethods::pca` instead. Note that `scale`/`center` are passed on to whichever underlying PCA function is used, so with `pcaMethods::pca` `scale` takes a scaling method name (`'uv'`, unit-variance, is the equivalent of `scale=TRUE` in `stats::prcomp`) rather than `TRUE`/`FALSE`.

```{r, fig.height=4, fig.width=5, out.width='60%'}
plot_pca(dia_prot, 'protein', colour_by='group', scale='uv', center=TRUE, allowing_missing=TRUE) +
  theme_biomasslmb(base_size=10)
```

MPXV and control plasma separate along the first principal component, consistent with the correlation heatmap above.

## Missingness and experimental condition

We're about to test protein-level abundances that contain missing values, so it's worth checking whether that missingness is random or structured by group. Missingness concentrated in one group — a protein consistently absent from the controls, say — both distorts the linear model and can be biologically meaningful in its own right, so it shouldn't be ignored.

`condition_miss_score` fits a per-protein logistic regression of missingness against group and returns Tjur's R² per protein: values near 1 mean a protein's missingness is well explained by group, values near 0 mean it looks unrelated to group. `condition_miss_index` summarises this into a single dataset-level value.

```{r}
miss_res <- condition_miss_score(dia_qf, i = 'protein', group_cols = 'group')
condition_miss_index(miss_res$summary)$index
```

The index is close to zero: for this whole-proteome comparison, protein-level missingness looks unrelated to group — in contrast to the enrichment experiment in Part B, where absence from the control pulldown is exactly the signal of interest. We therefore test the protein-level abundances directly, letting `limma` fit each protein on whichever samples it was quantified in, rather than imputing.

## Which proteins can be tested?

`limma` fits a linear model per protein, so each protein needs enough genuinely quantified values for its group means and its within-group variance to be estimable. A protein that fails this is not "not significant" — it is untested, and the two must not be reported as if they were the same thing.

We therefore require at least 2 quantified values in *both* conditions, and count the quantified values per condition so we can apply that rule and report it alongside the results.

```{r}
group <- factor(dia_prot[['protein']]$group, levels = c('control', 'MPXV'))

n_quant <- sapply(levels(group), function(g){
  rowSums(!is.na(assay(dia_prot[['protein']])[, group == g, drop = FALSE]))
})

testable <- rownames(n_quant)[apply(n_quant, 1, min) >= 2]

length(testable)
n_quant[apply(n_quant, 1, min) < 2, ]
```

Three of the excluded proteins have no quantified values at all in one condition. `lmFit` cannot estimate a group coefficient from an empty group, so it returns `NA` for those rows with a `Partial NA coefficients` warning. `NA` adjusted p-values are then dropped silently by `table()`, `sum(..., na.rm=TRUE)` and most other summaries, so an unfiltered run reports a plausible-looking number of significant proteins without ever indicating that some proteins could not be fitted. Filtering first makes the exclusion explicit and countable.

The remaining excluded proteins were quantified in one or both conditions, but only once in one of them, so their fold-change rests on a single measurement with no way to judge its reproducibility. This filter is not free: `P02741` (CRP) is quantified in all 6 MPXV samples and just 1 of 15 controls, and is almost certainly a genuine, large increase — it is missing from the controls precisely *because* it is low there. Excluding it is the conservative choice, not the correct one in any absolute sense; recovering proteins like it requires modelling the missingness or imputing, which is what `restrict_imputation()` and the `mnar_score()` family are for.

When imputed values are being tested, this becomes two filters rather than one, because "quantified" and "has a value" stop being the same thing: at least 2 *genuinely quantified* values in at least one condition, checked against the unimputed assay, so that no result rests entirely on imputed values; and at least 2 values in *both* conditions, checked against the imputed assay that is actually being fitted, so that the model can be fitted. Here nothing is imputed, so both reduce to the single rule applied above.

For reporting, we keep the per-condition counts as columns so that a reader of the results table can see how much of each fold-change is supported by measurements.

```{r}
n_quant_cols <- n_quant %>%
  as.data.frame() %>%
  setNames(paste0('n_quant_', levels(group))) %>%
  tibble::rownames_to_column('Protein')
```

## Differential abundance testing with limma

We use `limma` for differential abundance testing between `control` and `MPXV`. `limma` fits a linear model per protein, then applies an empirical Bayes step which borrows information across all proteins to stabilise the variance estimate — important here since the 6 MPXV replicates give fairly noisy per-protein variance estimates on their own.

We use `treat` rather than plain `eBayes`. `treat` tests the null hypothesis that the absolute fold-change is *below* a threshold (here 1.2-fold), rather than the standard null of exactly no difference — this avoids flagging changes that are statistically detectable but too small to be biologically interesting. `trend=TRUE` lets the variance prior depend on mean abundance; `robust=TRUE` down-weights proteins with unusually extreme variance so they don't distort the shared prior.

We attach gene symbols from the `protein` assay's `rowData` to the results, since they read more clearly than accessions on the volcano plot.

```{r, fig.show='hide'}
design <- model.matrix(~group)

fit <- lmFit(assay(dia_prot[['protein']])[testable, ], design)
final_model <- treat(fit, fc = 1.2, trend = TRUE, robust = TRUE)

gene_names <- as.data.frame(rowData(dia_qf[['protein']])[, c('Protein.Group', 'Genes')])

limma_results <- topTreat(final_model, coef = 'groupMPXV', number = Inf) %>%
  tibble::rownames_to_column('Protein') %>%
  left_join(gene_names, by = c('Protein' = 'Protein.Group')) %>%
  left_join(n_quant_cols, by = 'Protein') %>%
  arrange(P.Value)

table(direction = ifelse(limma_results$logFC < 0, 'Dec.', 'Inc.'),
      sig = limma_results$adj.P.Val < 0.05)
```

The FDR adjustment is over the `r nrow(limma_results)` testable proteins, itself a subset of those that passed the filtering thresholds in the QC vignette; a stricter or looser peptide/protein filter, or a stricter testability rule, would change the multiple-testing burden and hence the adjusted p-values.

## Volcano plot

`plot_volcano` plots log-fold-change against -log10(p-value) directly from a statistical testing results `data.frame`; by default it expects `logFC` and `adj.P.Val` columns, which is exactly what `topTreat` returns.

```{r, fig.width=6, fig.height=6}
sig_hits <- limma_results %>% filter(adj.P.Val < 0.05) %>% pull(Protein)

plot_volcano(limma_results, sig_col = NULL) +
  geom_point(data = filter(limma_results, Protein %in% sig_hits),
             aes(fill = adj.P.Val < 0.05), pch = 21, size = 3) +
  scale_fill_manual(values = get_cat_palette(2)[2], guide = 'none') +
  ggrepel::geom_text_repel(
    data = filter(limma_results, Protein %in% sig_hits),
    aes(label = Genes), min.segment.length = 0, size = 3)
```

The significant proteins are dominated by the acute-phase response: positive acute-phase reactants such as `SAA2`, `HP`, `LBP` and `C9` are increased in MPXV plasma, while negative acute-phase reactants such as `APOA1`, `APOA2`, `TTR` and `RBP4` are decreased.

## Inspecting proteins of interest

Before trusting a hit, it's worth looking at the abundance values behind it across the processing pipeline — from precursor level through to the final protein-level estimate — to confirm the result isn't being driven by a single outlying precursor. `plot_protein_assays` plots one or more proteins across any set of assays in the `QFeatures` object.

```{r, fig.height=6, fig.width=8, warning=FALSE, message=FALSE}
top_hit <- head(sig_hits, 1)

plot_protein_assays(
  dia_prot, top_hit,
  experiments_to_plot = c('precursors', 'peptides_filtered_norm', 'protein'),
  protein_id_col = 'Protein.Group',
  label_col = 'Genes',
  log2transform_cols = c('precursors')) +
  aes(colour = colData(dia_prot)[colname, 'group']) +
  scale_colour_manual(values = get_cat_palette(2), name = 'group') +
  theme(aspect.ratio = 1/3)
```

The precursor-level trend is consistent across the processing steps and across the individual precursors contributing to the protein, giving confidence that this is a genuine change rather than an artefact of summarisation.

# Part B: MaxQuant (MQ) input

`tmt_qf_mq` is the immunoprecipitation dataset from the [enrichment designs](interactome_designs.html) vignette: a bait pulldown (`IP`) against a control pulldown (`Control`), 6 replicates each, summarised to protein level with `robustSummary`.

```{r}
names(tmt_qf_mq)
knitr::kable(data.frame(colData(tmt_qf_mq)))
```

## Sample correlation

```{r, fig.height=5, fig.width=5}
plot_cor_samples(tmt_qf_mq, 'protein')
```

## Principal component analysis

As in Part A, `robustSummary` doesn't require complete PSMs, so `tmt_qf_mq[['protein']]` retains a small number of missing values (`r round(100*mean(is.na(assay(tmt_qf_mq[['protein']]))), 2)` % of the matrix), and we set `allowing_missing=TRUE` so the PCA runs via `pcaMethods::pca`.

```{r, fig.height=4, fig.width=5, out.width='60%'}
plot_pca(tmt_qf_mq, 'protein', colour_by='Condition', scale='uv', center=TRUE, allowing_missing=TRUE) +
  theme_biomasslmb(base_size=10)
```

IP and Control samples separate clearly, as expected for a specific enrichment.

## Which proteins can be tested?

The same testability rule applies here, but TMT quantification within a plex is close to complete once the S:N and interference filters of the QC vignette have been applied, so it excludes nothing.

```{r}
condition_mq <- factor(tmt_qf_mq[['protein']]$Condition, levels = c('Control', 'IP'))

n_quant_mq <- sapply(levels(condition_mq), function(g){
  rowSums(!is.na(assay(tmt_qf_mq[['protein']])[, condition_mq == g, drop = FALSE]))
})

table(testable = apply(n_quant_mq, 1, min) >= 2)
```

That is the usual position for a single-plex TMT experiment, and it is worth checking rather than assuming: the filter matters where missingness is common, which for TMT means comparisons that span plexes rather than comparisons within one.

## Differential abundance testing with limma

The statistical model itself is the same as Part A, but the biological question is different, and that changes what fold-change threshold is appropriate. In Part A, we were looking for shifts in a whole-proteome comparison, so a modest 1.2-fold `treat` threshold made sense. Here, we're trying to identify bona fide interactors of the bait protein: a real interactor should be substantially enriched in the IP relative to the control, not just detectably different. We therefore use a stricter 2-fold threshold.

```{r, fig.show='hide'}
design_mq <- model.matrix(~condition_mq)

fit_mq <- lmFit(assay(tmt_qf_mq[['protein']]), design_mq)
final_model_mq <- treat(fit_mq, fc = 2, trend = TRUE, robust = TRUE)

limma_results_mq <- topTreat(final_model_mq, coef = 'condition_mqIP', number = Inf) %>%
  tibble::rownames_to_column('Protein') %>%
  arrange(P.Value)

table(direction = ifelse(limma_results_mq$logFC < 0, 'Dec.', 'Inc.'),
      sig = limma_results_mq$adj.P.Val < 0.05)
```

Both enriched and depleted proteins can be statistically significant here, but for identifying interactors we care specifically about enrichment in the IP: a depleted protein doesn't indicate an interaction with the bait, so below we restrict to the significantly *enriched* proteins as our candidate interactors.

```{r}
sig_hits_mq <- limma_results_mq %>%
  filter(adj.P.Val < 0.05, logFC > 0) %>%
  pull(Protein)

length(sig_hits_mq)
```

## Volcano plot

With `r length(sig_hits_mq)` candidate interactors, labelling every point would make the plot unreadable, so we label only the 10 most significant.

```{r, fig.width=6, fig.height=6}
top_hits_mq <- head(sig_hits_mq, 10)

plot_volcano(limma_results_mq, sig_col = NULL) +
  geom_point(data = filter(limma_results_mq, Protein %in% sig_hits_mq),
             aes(fill = adj.P.Val < 0.05), pch = 21, size = 3) +
  scale_fill_manual(values = get_cat_palette(2)[2], guide = 'none') +
  ggrepel::geom_text_repel(
    data = filter(limma_results_mq, Protein %in% top_hits_mq),
    aes(label = Protein), min.segment.length = 0, size = 3)
```

## Inspecting proteins of interest

As in Part A, we inspect the feature-to-protein trend behind the top candidate interactor, here working from PSMs rather than precursors: `psms_raw` and `psms_filtered_interference` in place of `precursors` and `peptides_filtered_norm`, and `Leading.razor.protein` in place of `Protein.Group`.

```{r, fig.height=6, fig.width=8, warning=FALSE, message=FALSE}
top_hit_mq <- head(sig_hits_mq, 1)

plot_protein_assays(
  tmt_qf_mq, top_hit_mq,
  experiments_to_plot = c('psms_raw', 'psms_filtered_interference', 'protein'),
  protein_id_col = 'Leading.razor.protein',
  label_col = 'Leading.razor.protein',
  log2transform_cols = c('psms_raw')) +
  aes(colour = colData(tmt_qf_mq)[colname, 'Condition']) +
  scale_colour_manual(values = get_cat_palette(2), name = 'Condition') +
  theme(aspect.ratio = 1/3)
```

The PSMs consistently show higher abundance in the IP samples than the Control samples, supporting this as a genuine interactor rather than an artefact of a single PSM or a missing-value pattern.

# Summary

Here, we have, for both a whole-proteome comparison (Part A) and an immunoprecipitation (Part B):

-   Checked that replicate samples correlate more strongly with each other than with the other condition, and that PCA separates the conditions — using `allowing_missing=TRUE` in both parts, since `robustSummary` retains some missing values
-   Checked whether protein-level missingness was structured by experimental condition: unrelated to group in the whole-proteome comparison (Part A), but strongly condition-informative in the immunoprecipitation (see the QC vignette), where a protein absent from the control pulldown is itself the signal of interest
-   Restricted testing to proteins with at least 2 quantified values in both conditions, and kept the per-condition counts in the results so that untested proteins are distinguishable from proteins tested and found unchanged
-   Tested for differential protein abundance with `limma::treat`, using a fold-change threshold appropriate to the biological question: a modest threshold for whole-proteome changes in Part A, and a strict threshold to identify substantially-enriched candidate interactors in Part B
-   Visualised the results with a volcano plot and inspected the feature-to-protein trend for a top hit

# Where to go next

-   [Functional enrichment](functional_enrichment.html) asks whether the proteins that changed share a function, location or complex, rather than reading them one at a time. It picks up from the Part A results.
-   [Choosing a summarisation method](summarisation_methods.html) covers how the protein-level values tested here were arrived at, and what a different choice would have changed.
-   [Enrichment designs](interactome_designs.html) covers the QC and normalisation decisions behind the Part B dataset, which differ from those a whole-proteome experiment calls for.

```{r}
sessionInfo()
```
