---
title: "LFQ-DDA workflow: peptide QC and protein summarisation"
author: "Tom Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: biomasslmb.json
vignette: >
  %\VignetteIndexEntry{LFQ-DDA workflow: peptide QC and protein summarisation}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


Label-Free Quantification (LFQ) is the simplest form of quantitative bottom-up proteomics, in which different samples are quantified in separate MS runs. Quantification is either performed by Data-Dependent Acquisition (DDA), where the Mass Spectrometer triggers fragmentation of ions within a given m/z range with the aim being to focus attention of individual peptides separately, or Data-Independent Acquisition (DIA), where a much wider m/z range is used and a mix of peptides are co-fragmented and quantified simultaneously by deconvoluting the resultant complex spectra. This article covers LFQ-DDA; DIA is a different enough problem to have its own, in [LFQ-DIA precursor QC](LFQ_DIA_Precursor_QC_Summarisation.html).

Since each sample is run separately, different peptides will be quantified in each sample and the peptide intensities may not be directly comparable between samples. The common solution to the higher burden of missing values is to use the 'match-between-runs' [@http://zotero.org/users/5634351/items/87YIDK69], or the functionally equivalent 'Minora' algorithm employed by Proteome Discoverer (PD). These algorithms use the observed retention times of MS1 ions which were successfully spectrum matched in one sample to identify the likely peptide sequence of MS1 ions that could not be spectrum matched in another sample. However, even with these algorithms enabled, DDA LFQ will still typically have many more missing values than labelled proteomics, e.g TMT.

The analysis is straightforward in outline. Three of its steps carry most of the risk — normalisation, the missing value threshold, and the protein inference that summarisation rests on — and each is flagged where it arrives below.

This vignette works through one typical experiment from end to end: a whole-proteome comparison between two conditions, three replicates each. That is the common case, and the choices made below are the conventional ones for it. An enrichment experiment — an IP, BioID or TurboID pulldown — needs different reasoning about normalisation and about missing values, and is covered in [enrichment designs](interactome_designs.html).

If this is your first analysis with the package, [getting started](biomasslmb.html) explains which route applies to your experiment and shows a complete one in forty lines, and [working with QFeatures objects](qfeatures_objects.html) covers the object every step below operates on. Annotations are worth retrieving before any of this — see [protein annotation](protein_annotation.html).

# Load required packages

```{r, message=FALSE}
library(QFeatures)
library(biomasslmb)
library(ggplot2)
library(tidyr)
library(dplyr)
```

# The experimental design

`lfq_dda_pd_PeptideGroups.txt` is a file available from the `biomasslmb` package containing the peptide-level output from Proteome Discoverer (PD) for a real DDA LFQ experiment: a whole-proteome comparison between a wild type human cell line and a point mutant of the same line, 3 replicates each. It is truncated to a subset of proteins for a manageable vignette.

```{r}
pep_inf <- system.file(
  "extdata", "lfq_dda_pd_PeptideGroups.txt",
  package = "biomasslmb"
)
```

# Defining the contaminant proteins

Contaminant proteins have to be removed. This search was performed against the '0602_Universal Contaminants' database [@Frankenfield2022], so parsing that same fasta gives the accessions to match against.

```{r}

contaminant_fasta_inf <- system.file(
  "extdata", "0602_Universal_Contaminants.fasta.gz",
  package = "biomasslmb"
)

# Extract the protein IDs associated with each contaminant protein
contaminant_accessions <- biomasslmb::get_contaminant_fasta_accessions(contaminant_fasta_inf)

print(head(contaminant_accessions))
```

This PD export also includes PD's own `Contaminant` column, populated from the same search against the '0602_Universal Contaminants' database. `filter_features_pd_dda` uses this column automatically, in addition to the `contaminant_accessions` extracted above.

These are separate defences that do not always agree with one another, and when one of them silently fails the filtering still appears to work. [Contaminants and protein FDR](gotcha_contaminants_and_FDR.html) measures what each one catches and what a mismatch between the accession list and the search database leaves behind.

# Read in input data

The data go into a `QFeatures` object, the standard Bioconductor container for quantitative proteomics data. See [here](https://www.bioconductor.org/packages/release/bioc/html/QFeatures.html) for documentation about the `QFeatures` object.

The abundance column names are not useful as they stand — they carry the PD file ID (`F7`, `F8`, ...) rather than which sample is which.
```{r}
infdf <- read.delim(pep_inf)

abundance_cols_ix <- grep('^Abundance', colnames(infdf))

colnames(infdf)[abundance_cols_ix]
```

Which file ID corresponds to which sample comes from the design table supplied with the experiment, so that is built first. The three wild type replicates were acquired before the three mutant replicates.

```{r}
exp_design <- data.frame(
  quantCols = colnames(infdf)[abundance_cols_ix],
  Condition = rep(c('WT', 'Mutant'), each = 3),
  Replicate = rep(1:3, times = 2))

exp_design$Sample <- paste(exp_design$Condition, exp_design$Replicate, sep = '_')

knitr::kable(exp_design)
```

Relabelling the abundance columns with the sample names means everything downstream refers to samples rather than file IDs.

```{r}
colnames(infdf)[abundance_cols_ix] <- exp_design$Sample
exp_design$quantCols <- exp_design$Sample

colnames(infdf)[abundance_cols_ix]
```

The data can now be read in.
```{r}
# Read in peptide-level quantification from the DDA LFQ experiment (using QFeatures function)
lfq_qf <- readQFeatures(assayData = infdf,
                        quantCols = abundance_cols_ix,
                        colData = exp_design,
                        name = "peptides_raw")

```

`readQFeatures()` attaches the design at the object level only, so `sync_coldata()` copies it onto the assay as well, which is what the plotting and modelling functions read. See [working with QFeatures objects](qfeatures_objects.html).

```{r}
lfq_qf <- sync_coldata(lfq_qf, 'peptides_raw')
```

The retention time (RT) distribution is worth assessing before anything is filtered, because a chromatography problem is cheaper to find now than after summarisation. `plot_rt_dist` defaults to the column name Proteome Discoverer uses for a Sequest HT search; this search was run with CHIMERYS, so the column is named explicitly.

**The peptide count holds up across most of the gradient**, tailing off over the last 20 minutes, which is what an even elution looks like. A large early or late spike, or a gap in the middle, would point at a chromatography problem worth resolving before going further.

```{r, warning=FALSE, fig.width=5, fig.height=4, out.width="60%"}
rt_col <- 'PSM.RT.in.min.by.Search.Engine.CHIMERYS'

plot_rt_dist(lfq_qf[['peptides_raw']], rt_col = rt_col)
```

The companion check, `plot_rt_vs_delta`, plots the precursor mass error against RT and would reveal a mass calibration that drifts across the run. It needs a reported mass error to work with, and this export gives a delta of exactly zero for every peptide, so there is nothing to plot here. A Sequest HT export reports the measured error and the check is worth making.


# Filter peptides

Routine filtering removes peptides that:

-   Could originate from contaminants. See `?filter_features_pd_dda` for
    further details, including the removal of 'associated' contaminants.
-   Don't have any quantification values

`unique_master = FALSE` keeps the peptides whose master protein is a group of several accessions, such as `Q16875; P16118` — `r sum(grepl(';', rowData(lfq_qf[['peptides_raw']])$Master.Protein.Accessions))` of the `r nrow(lfq_qf[['peptides_raw']])` peptides here, in `r length(unique(grep(';', rowData(lfq_qf[['peptides_raw']])$Master.Protein.Accessions, value = TRUE)))` distinct groups. Each group summarises to its own protein row later in this vignette, separate from the rows for its individual members; `unique_master = TRUE` drops those peptides instead. The `proteotypic` argument asks a different question — whether the peptide sequence occurs in only one protein at all — and is likewise not applied here. See [peptides are not proteins](gotcha_peptide_to_protein.html).

```{r}
# Perform routine raw data filtering.
# - Remove peptides from contaminant proteins
# - Remove peptides with no master protein

lfq_qf[['peptides_filtered']] <- filter_features_pd_dda(lfq_qf[['peptides_raw']],
                                                 contaminant_proteins=contaminant_accessions,
                                                 filter_contaminant=TRUE,
                                                 filter_associated_contaminant=TRUE,
                                                 unique_master=FALSE,
                                                 remove_no_quant = TRUE)

lfq_qf <- sync_coldata(lfq_qf, 'peptides_filtered')
```


Peptides that are not rank 1 according to the search engine are removed next. Every peptide in this export is rank 1, so nothing goes at this step — but it belongs in the pipeline, because an export that reports lower-ranked matches will contain some.

```{r}

lfq_qf <- lfq_qf %>%
filterFeatures(~ PSM.Search.Engine.Rank.by.Search.Engine.CHIMERYS == 1,
               i = "peptides_filtered")

biomasslmb:::message_parse(rowData(lfq_qf[['peptides_filtered']]),
                           'Master.Protein.Accessions',
                           "Removing peptides that are not rank 1")
```

Summarisation needs the quantification values on a log scale, so they are transformed
here.

```{r}
lfq_qf[['peptides_filtered']] <- QFeatures::logTransform(
  lfq_qf[['peptides_filtered']], base=2)
```

The peptide intensity distributions, coloured by `Condition`:

```{r, warning=FALSE, fig.show='hold', fig.show="hold", out.width="50%"}

# Plot the peptide-level quantification distributions per sample
plot_quant(lfq_qf[['peptides_filtered']], log2transform=FALSE, method='density') +
  theme_biomasslmb() +
  aes(colour=Condition, group=sample) +
  xlab('Peptide abundance (log2)')
```

## Normalise

The same amount of material was injected for every sample, and this is a whole-proteome comparison in which most proteins are expected to be unchanged between the two conditions, so any systematic offset between the distributions above is technical rather than biological. `diff.median` normalisation shifts each sample to a common median and removes it.

```{r, warning=FALSE, fig.show='hold', fig.show="hold", out.width="50%"}
lfq_qf[['peptides_filtered_norm']] <- QFeatures::normalize(
  lfq_qf[['peptides_filtered']], method = 'diff.median')

lfq_qf <- sync_coldata(lfq_qf, 'peptides_filtered_norm')

plot_quant(lfq_qf[['peptides_filtered_norm']], log2transform=FALSE, method='density') +
  theme_biomasslmb() +
  aes(colour=Condition, group=sample) +
  xlab('Peptide abundance (log2)')
```

That assumption is what makes this step safe, and it is worth stating explicitly because it is not always true: an [enrichment design](interactome_designs.html) expects the conditions to differ in overall composition, and normalising to a common median there erodes the signal being measured.

## Summarising to protein-level abundances

Summarisation takes the master protein assignment at face value: every peptide contributes to exactly one protein, the one the search engine assigned it to. That assignment is an inference rather than a measurement, and for proteins whose peptides are largely shared with their homologues it is not a reliable one. [Peptides are not proteins](gotcha_peptide_to_protein.html) covers how the assignment is made and how to tell which of your proteins it can support.

Before deciding how to handle missing values, it's worth checking how much missingness there is,
and whether it's structured by `Condition`. `condition_miss_score` fits, for each peptide, a
logistic regression of missingness against `Condition`, and returns Tjur's R² as a per-feature
score: values near 1 mean a peptide's missingness pattern is well explained by `Condition`.
`condition_miss_index` aggregates these into a single dataset-level value between 0 and 1.

```{r}
miss_res <- condition_miss_score(lfq_qf, i = 'peptides_filtered_norm', group_cols = 'Condition')
condition_miss_index(miss_res$summary)$index
```

**Missing values are common, but they are not structured by condition.** `r miss_res$global['n_features_informative']` of the `r miss_res$global['n_features_total']` peptides have at least one, and yet the index is low. That is the expected picture for a whole-proteome comparison: a peptide near the limit of detection drops out of whichever runs happened to miss it, rather than out of one condition specifically. `plot_missing_upset` shows the most common patterns.

```{r, fig.height=6, fig.width=8, warning=FALSE, message=FALSE}
plot_missing_upset(lfq_qf, i='peptides_filtered_norm')
```

Unstructured or not, there are far too many missing values to discard every peptide that has one, which is the usual state of affairs for DDA LFQ. That rules out summing, which needs complete features, and makes `MsCoreUtils::robustSummary` the appropriate estimator: it summarises accurately without requiring complete data [@http://zotero.org/users/5634351/items/FZN3QTTZ]. Peptides missing in almost every sample still contribute little beyond noise, so a threshold is applied first — at most 4/6 missing values here. Set it tighter and more of the proteome falls below the two-peptide rule imposed next; set it looser and protein estimates rest on peptides measured twice. [Comparing protein summarisation approaches](summarisation_methods.html) puts `sum` and `robustSummary` side by side on this same dataset.

```{r}

lfq_qf[['peptides_filtered_missing']] <- QFeatures::filterNA(
  lfq_qf[['peptides_filtered_norm']], 4/6)

biomasslmb:::message_parse(rowData(lfq_qf[['peptides_filtered_missing']]),
                           'Master.Protein.Accessions',
                           "Removing peptides with more than 4/6 missing values")
```

Peptides belonging to proteins with fewer than 2 peptides are then removed.

```{r}
min_peps <- 2
lfq_qf[['peptides_for_summarisation']] <- filter_features_per_protein(
  lfq_qf[['peptides_filtered_missing']], min_features = min_peps)

biomasslmb:::message_parse(rowData(lfq_qf[['peptides_for_summarisation']]),
                           'Master.Protein.Accessions',
                           "Removing 'one-hit' wonders")
```

Summarising with `robustSummary`:

```{r, warning=FALSE}
set.seed(42)

# Aggregate to protein-level abundances (using QFeatures function)
lfq_qf <- aggregateFeatures(lfq_qf,
                            i = "peptides_for_summarisation",
                            fcol = "Master.Protein.Accessions",
                            name = "protein",
                            fun = MsCoreUtils::robustSummary,
                            maxit=10000)

lfq_qf <- sync_coldata(lfq_qf, 'protein')
```

The two-peptide filter above was applied per protein, not per sample, and peptides with missing values were deliberately kept — so in any given sample a protein can still end up summarised from a single peptide. Nothing in the output marks these values as weaker than the rest. `get_protein_no_quant_mask()` finds where a protein abundance rests on fewer than `n` peptides, and `mask_protein_level_quant()` replaces those values with `NA`.

```{r, warning=FALSE, fig.show='hold', out.width="50%", fig.width=4, fig.height=4}
# plot = TRUE means we will also get a plot of the number of proteins quantified in each sample
protein_retain_mask <- biomasslmb::get_protein_no_quant_mask(
  lfq_qf[['peptides_for_summarisation']], min_features=min_peps, plot=TRUE)

lfq_qf[['protein']] <- biomasslmb::mask_protein_level_quant(
  lfq_qf[['protein']], protein_retain_mask)

```

## Re-inspecting missing values at protein-level

Masking changes the missingness picture, so it is worth re-reading at protein level. `r round(nNA(lfq_qf[['protein']])$nNA$pNA*100, 3)`% of protein values are missing overall, and at most `r round(max(nNA(lfq_qf[['protein']])$nNAcols$pNA)*100, 3)`% in any one sample. Summarising has reduced the missingness considerably relative to the peptide level, since a protein only loses a sample when every one of its peptides does. What remains is spread across samples rather than concentrated in one condition, consistent with the peptide-level picture above.

```{r, fig.height=6, fig.width=8, warning=FALSE, message=FALSE}

plot_missing_upset(lfq_qf, i='protein' )
```


## Inspecting the number of peptides and proteins through the processing steps

Each filtering step above removed something, and the totals are easier to judge together than one message at a time. `get_samples_present()` and `plot_samples_present()` show how many peptides and proteins survived each stage, and in how many samples. Both take a named character vector selecting the assays to plot, since the `QFeatures` names are not self-explanatory.

The assay names to choose from:

```{r}
names(lfq_qf)
```

### Samples per peptide

Peptides first. The row variables are `Annotated.Sequence` and `Modifications`, so that the count is of unique modified peptides rather than of rows.

```{r, fig.height=7, , fig.width=7, fig.fullwidth=TRUE, fig.cap="Samples quantified for each peptide at each level of processing", out.width='75%'}


rename_cols <- c('All peptides' = 'peptides_raw' ,
                 'Quantified, contaminants removed' = 'peptides_filtered',
                 'At most 4/6 missing values' = 'peptides_filtered_missing',
                 '>1 peptide per protein' = 'peptides_for_summarisation')

rowvars <- c('Annotated.Sequence', 'Modifications')

samples_present <- get_samples_present(lfq_qf[,,unname(rename_cols)], rowvars, rename_cols)
plot_samples_present(samples_present, rowvars, breaks=seq(2,6,2)) + ylab('Peptide')



```

### Samples per Protein

Then proteins, from the same functions: the named vector gains the `protein` assay, and the row variable becomes `Master.Protein.Accessions` alone.

```{r, fig.height=7, , fig.width=7, fig.fullwidth=TRUE, fig.cap="Samples quantified for each protein at each level of processing", out.width='75%'}

rename_cols_prot <- c(rename_cols, 'Protein'='protein')

rowvars_prot <- c('Master.Protein.Accessions')

samples_present <- get_samples_present(lfq_qf, rowvars_prot, rename_cols_prot)
plot_samples_present(samples_present, rowvars_prot, breaks=seq(2,6,2))
```

**The two-peptide rule is cheap in peptides and expensive in proteins.** It removes few peptides, because most peptides belong to proteins that have several; it removes many more proteins, because a protein identified by one peptide loses its only evidence. Those proteins are the ones whose quantification would have been least reliable, so the trade is usually worth making — but it is a trade, and an experiment targeting a low-abundance protein family may not want it.

The same six runs were also processed with MaxQuant, and [comparing processing pipelines](processing_pipeline_comparison.html) puts that output through this pipeline to show how much of the protein list and the significant hits depend on the choice of pipeline rather than on the biology.


```{r, eval = FALSE}
# Save file to package as data so it can be read back in in other vignettes
usethis::use_data(lfq_qf, overwrite=TRUE)
```

```{r}
sessionInfo()
```
