---
title: "LFQ-DIA workflow: precursor QC and protein summarisation"
author: "Tom Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: biomasslmb.json
vignette: >
  %\VignetteIndexEntry{LFQ-DIA workflow: precursor QC and protein summarisation}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Label-Free Quantification (LFQ) is the simplest form of quantitative proteomics, in which different samples are quantified in separate MS runs. Quantification is either performed by Data-Dependent Acquisition (DDA), where the Mass Spectrometer triggers fragmentation of ions within a given m/z range with the aim being to focus attention of individual peptides separately, or Data-Independent Acquisition (DIA), where a much wider m/z range is used and a mix of peptides are co-fragmented and quantified simultaneously by deconvoluting the resultant complex spectra. Here, we will focus on just LFQ-DIA.

DIA data is commonly searched with [DIA-NN](https://github.com/vdemichev/DiaNN), which reports results at multiple levels. Here, we will use the data from the precursor level (a specific peptide sequence, with modifications, at a specific charge state). Since DIA co-fragments many precursors at once, DIA-NN infers identifications either from a project-specific spectral library or via a library predicted in-silico from the sequence database, then rescores them, optionally using match-between-runs (MBR) to borrow confidence across runs. As with DDA, we still need to filter to confident identifications, remove contaminants, handle missing values, and summarise to protein-level abundance before any statistical analysis.

This vignette works through one typical experiment from end to end: a whole-proteome comparison between groups, with no enrichment step and a single acquisition batch. That is the common case, and the choices made below are the conventional ones for it. An enrichment experiment — an IP, BioID or TurboID pulldown — needs different reasoning about normalisation and about missing values, and is covered in [enrichment designs](interactome_designs.html).

The data used in this vignette is a real DIA-NN `report.parquet` output for a plasma proteomics case series comparing Mpox (MPXV) infection, COVID-19, and healthy controls [@Wang2022]. It is also used, with accompanying teaching material, in the [Proteomics Data Analysis Using R](https://github.com/bioinformatics-core-shared-training/Proteomics_stats_R) course. Unlike the truncated demonstration files used elsewhere in `biomasslmb`, this dataset is used here in full - it is small enough not to need subsetting

Spectronaut is the other search engine in common use for DIA data. Only the reading step differs: [reading Spectronaut output](#reading-spectronaut-output) at the end of this vignette shows how to reach the same `precursors` assay, after which every step below applies unchanged.

# Load required packages

```{r, message=FALSE}
library(QFeatures)
library(biomasslmb)
library(ggplot2)
library(tidyr)
library(dplyr)
library(arrow)
library(readxl)
```

# Defining the contaminant proteins

We need to remove contaminant proteins. This search was performed against a version of the Hao lab's '0602 Universal Contaminants' database [@Frankenfield2022] with the Fetal Bovine Serum-derived entries removed (not relevant for a human plasma sample). We use the standard '0602 Universal Contaminants' database bundled with `biomasslmb` below; the FBS-specific entries it additionally contains are simply never observed in this dataset, so filtering behaves identically either way.

```{r}
contaminant_fasta_inf <- system.file(
  "extdata", "0602_Universal_Contaminants.fasta.gz",
  package = "biomasslmb"
)

# Extract the protein IDs associated with each contaminant protein
contaminant_accessions <- biomasslmb::get_contaminant_fasta_accessions(contaminant_fasta_inf)

print(head(contaminant_accessions))
```

DIA-NN was run with this contaminants database included directly in the search FASTA, so contaminant entries are named with a `Cont_` prefix that is carried straight through into the `Protein.Ids` and `Protein.Group` columns of the DIA-NN output. `filter_features_diann` uses this prefix automatically, in addition to the `contaminant_accessions` we've just extracted.

The prefix and the accession list are separate defences, and a search database whose contaminant entries were never renamed provides neither. [Contaminants and protein FDR](gotcha_contaminants_and_FDR.html) measures what each one catches and what is left behind when they do not match.

# Read in input data

We start by reading in quantitative proteomics data into a `QFeatures` object, which is the standard Bioconductor object for holding quantitative proteomics data. See [here](https://www.bioconductor.org/packages/release/bioc/html/QFeatures.html) for documentation about the `QFeatures` object.

Unlike PD's peptide-level export, DIA-NN's `report.parquet` is a **long-format** table: each row is a single precursor-sample combination, rather than one row per precursor across all samples. We read it in with `arrow::read_parquet`.

```{r}
diann_report_inf <- system.file(
  "extdata", "monkeypox_plasma_proteomes.parquet",
  package = "biomasslmb"
)

diann_df <- arrow::read_parquet(diann_report_inf)

dim(diann_df)
```

Sample metadata (which group, sex and age band each run belongs to) is provided separately, in `monkeypox_metadata.xlsx`.

```{r}
metadata_inf <- system.file(
  "extdata", "monkeypox_metadata.xlsx",
  package = "biomasslmb"
)

sample_metadata <- readxl::read_excel(metadata_inf)

print(sample_metadata)
```

The `Run` column in the DIA-NN output inherits the full raw file path, pasting together the `.wiff2` folder and `.wiff.scan` file names. We clean this up so it matches the `sample_id` field in our metadata, then add the concise `runCol` label from the metadata, which we'll use throughout as the sample identifier.

```{r}
diann_df[, "Run"] %>% unique() %>% head(2)

diann_df <- diann_df %>%
  mutate(Run = sub('\\..*', '', Run)) %>%
  merge(sample_metadata[, c("sample_id", "runCol")], by.x = "Run", by.y = "sample_id")

diann_df[, "Run"] %>% unique() %>% head(2)
```

Now we can read the data into a `QFeatures` object, using `readQFeaturesFromDIANN` from `QFeatures`. This creates one `SummarizedExperiment` per run, since the per-run quality metrics in `rowData` (e.g. `Q.Value`, `FWHM`) are run-specific and would be lost if all runs were combined immediately. We filter on these metrics below before joining the runs into a single combined set.

```{r}
dia_qf <- readQFeaturesFromDIANN(
  diann_df,
  quantCols = "Precursor.Quantity",
  fnames = "Precursor.Id",
  runCol = "runCol",
  colData = sample_metadata
)

dia_qf
```

It's helpful to assess the retention time (RT) distribution for the precursors and its relationship with the chromatographic peak width (FWHM, the Full-Width at Half Maximum) --- narrow, symmetrical peaks are preferred, and we don't expect FWHM to increase systematically with RT.

```{r, warning=FALSE, fig.show='hold', fig.width=5, fig.height=4}
rbindRowData(dia_qf, i = names(dia_qf)) %>%
  data.frame() %>%
  ggplot(aes(RT, FWHM)) +
  geom_hex(bins = 60) +
  theme_biomasslmb(aspect_square = FALSE) +
  labs(x = 'Retention time (min)', y = 'Peak width, FWHM (min)')
```

# Filter precursors

DIA-NN assigns a Q-value to each precursor and protein group identification, representing the local false discovery rate (FDR) at that score threshold. Since this data was processed with match-between-runs enabled, we filter on both the run-level and library-level Q-values (`Q.Value`/`PG.Q.Value` and `Lib.Q.Value`/`Lib.PG.Q.Value` respectively) at a standard 1% FDR threshold. We don't pass an `i` argument to `filterFeatures`, so the filter is applied to every per-run set simultaneously.

```{r}
dia_qf <- dia_qf %>%
  filterFeatures(~ Q.Value <= 0.01) %>%
  filterFeatures(~ PG.Q.Value <= 0.01) %>%
  filterFeatures(~ Lib.Q.Value <= 0.01) %>%
  filterFeatures(~ Lib.PG.Q.Value <= 0.01)
```

Next, we combine the per-run sets into a single `"precursors"` set with `joinAssays`, aligning features by `Precursor.Id`. A precursor not detected in a given run becomes `NA` for that sample. Only `rowData` columns that are both shared across runs and have consistent values are retained in the combined set; run-specific quality metrics we've already filtered on (e.g. `Q.Value`, `RT`) are dropped, since they're no longer needed.

```{r}
dia_qf <- joinAssays(
  x = dia_qf,
  i = names(dia_qf),
  name = "precursors",
  fcol = "Precursor.Id"
)

# The per-run sets are no longer needed now they're joined; removing them
# keeps the QFeatures object focused on the combined set for the rest of
# this vignette.
dia_qf <- removeAssay(dia_qf, i = which(names(dia_qf) != "precursors"))

colData(dia_qf[['precursors']]) <- colData(dia_qf)
```

We first perform routine filtering to remove precursors that:

-   Could originate from contaminants. See `?filter_features_diann` for further details, including the removal of contaminant-associated proteins.
-   Don't have a unique master protein
-   Don't have any quantification values

```{r}
dia_qf[['peptides_filtered']] <- biomasslmb::filter_features_diann(
  dia_qf[['precursors']],
  contaminant_proteins = contaminant_accessions
)

colData(dia_qf[['peptides_filtered']]) <- colData(dia_qf)
```

For the summarisation, we need our quantification values to be on a log-scale, so we log-transform below

```{r}
dia_qf[['peptides_filtered']] <- QFeatures::logTransform(
  dia_qf[['peptides_filtered']], base = 2)
```

Next, we plot the precursor intensities, coloured by `group`.

```{r, warning=FALSE, fig.show='hold', fig.width=7, fig.height=5, out.width="75%"}
biomasslmb::plot_quant(dia_qf[['peptides_filtered']], log2transform = FALSE, method = 'density') +
  theme_biomasslmb(aspect_square = FALSE) +
  aes(colour = group, group = sample) +
  xlab('Precursor abundance (log2)')
```

## Normalise

This is a whole-plasma-proteome comparison between clinical groups, where we don't expect a systematic difference in overall protein content between groups. Standard normalisation is therefore appropriate here, and we apply median normalisation with `QFeatures::normalize`.

```{r}
dia_qf <- QFeatures::normalize(
  dia_qf, i = 'peptides_filtered', name = 'peptides_filtered_norm', method = 'diff.median')

colData(dia_qf[['peptides_filtered_norm']]) <- colData(dia_qf)
```

```{r, warning=FALSE, fig.show='hold', fig.width=7, fig.height=5, out.width="75%"}
biomasslmb::plot_quant(dia_qf[['peptides_filtered_norm']], log2transform = FALSE, method = 'density') +
  theme_biomasslmb(aspect_square = FALSE) +
  aes(colour = group, group = sample) +
  xlab('Precursor abundance (log2, normalised)')
```

## Summarising to protein-level abundances

For the summarisation to protein, we will use `robustSummary`. We do not need to remove
peptides with any missing values. However, we still don't want to retain
peptides with too many missing values, since these will not be very
informative in estimating the protein-level quantification.

Summarisation takes the protein group assignment at face value: every precursor contributes to exactly one protein group. That assignment is an inference rather than a measurement, and for proteins whose peptides are largely shared with their homologues it is not a reliable one. [Peptides are not proteins](gotcha_peptide_to_protein.html) covers how the assignment is made and how to tell which of your proteins it can support.

Before deciding how to handle missing values, it's worth checking how much missingness there is,
and whether it's structured by `group`. `condition_miss_score` fits, for each peptide, a
logistic regression of missingness against `group`, and returns Tjur's R² as a per-feature
score: values near 1 mean a peptide's missingness pattern is well explained by `group`.
`condition_miss_index` aggregates these into a single dataset-level value between 0 and 1.

```{r}
miss_res <- condition_miss_score(dia_qf, i = 'peptides_filtered_norm', group_cols = 'group')
condition_miss_index(miss_res$summary)$index
```

With `r miss_res$global['n_features_informative']` out of `r miss_res$global['n_features_total']` precursors having any missing values, this data has the substantial missingness typical of DIA-NN precursor-level output, but the condition missingness index is close to zero: unlike a pulldown/enrichment experiment, missingness here is not strongly explained by clinical group. `plot_missing_upset` shows the most common patterns.

```{r, fig.height=6, fig.width=8, warning=FALSE, message=FALSE}
plot_missing_upset(dia_qf, i = 'peptides_filtered_norm')
```

Since DIA-NN typically has too many missing values to exclude peptides with any missing values, the `MsCoreUtils::robustSummary` method should be preferred for summarisation to protein-level abundance. With `robustSummary`, we do not need to remove all peptides with missing values since the summarisation algorithm deals with them appropriately [@http://zotero.org/users/5634351/items/FZN3QTTZ]. However, we still don't want to retain peptides with too many missing values, since these will not be very informative in estimating the protein-level quantification. Here, we will retain peptides observed in at least 25% of the 31 samples (`pNA` of at most 0.75), matching the threshold applied to this dataset in the [Proteomics Data Analysis Using R](https://github.com/bioinformatics-core-shared-training/Proteomics_stats_R) course.

```{r}
dia_qf[['peptides_filtered_missing']] <- QFeatures::filterNA(
  dia_qf[['peptides_filtered_norm']], pNA = 0.75)

biomasslmb:::message_parse(rowData(dia_qf[['peptides_filtered_missing']]),
                           'Protein.Group',
                           "Removing peptides with > 75% missing values")
```

Next, we remove peptides for proteins with fewer than 2 peptides. This is routine practice and worth being explicit about, because it is doing more than tidying.

A protein quantified from a single precursor has no internal replication: there is no second measurement to disagree with the first, so nothing about it can be checked. Any error in that one precursor — a misassigned peptide, interference, a bad integration — passes straight through to the protein-level value with nothing to average it out. Its missingness pattern is also the missingness pattern of one precursor rather than of the protein, which matters for everything downstream that reasons about missing values.

The cost is real, and asymmetric: single-precursor proteins are disproportionately the low-abundance ones, so this filter removes preferentially from the part of the proteome that was hardest to measure. `min_features` is the dial. Raising it buys confidence and loses coverage; there is no threshold that is correct independent of what the result will be used for.

```{r}
min_peps <- 2
dia_qf[['peptides_for_summarisation']] <- filter_features_per_protein(
  dia_qf[['peptides_filtered_missing']], min_features = min_peps, master_protein_col = 'Protein.Group')

biomasslmb:::message_parse(rowData(dia_qf[['peptides_for_summarisation']]),
                           'Protein.Group',
                           "Removing 'one-hit' wonders")
```

Now we can summarise with `robustSummary`

```{r, warning=FALSE}
set.seed(42)

dia_qf <- aggregateFeatures(dia_qf,
                            i = "peptides_for_summarisation",
                            fcol = "Protein.Group",
                            name = "protein",
                            fun = MsCoreUtils::robustSummary,
                            maxit = 10000)

colData(dia_qf[['protein']]) <- colData(dia_qf)
```

Prior to summarisation, we removed peptides from proteins with fewer than 2
peptides. However, since we left in peptides with missing values, it's
possible for some protein-level abundances to be derived from just a
single peptide. We can use `get_protein_no_quant_mask` from
`biomasslmb` to identify where the protein abundances will be derived
from fewer than `n` features (peptides). We can then give this mask to
`mask_protein_level_quant` to replace these quantification values with
NA.

```{r, warning=FALSE, fig.show='hold', out.width="50%", fig.width=8, fig.height=4}
protein_retain_mask <- biomasslmb::get_protein_no_quant_mask(
  dia_qf[['peptides_for_summarisation']], min_features = min_peps,
  master_protein_col = 'Protein.Group')

dia_qf[['protein']] <- biomasslmb::mask_protein_level_quant(
  dia_qf[['protein']], protein_retain_mask)
```

## Re-inspecting missing values at protein-level

Now that we have protein-level abundances, we would like to re-inspect the missing values. Overall, we have `r round(nNA(dia_qf[['protein']])$nNA$pNA*100, 3)` % missing values, with at most `r round(max(nNA(dia_qf[['protein']])$nNAcols$pNA)*100, 3)` % missing values in any given sample.

```{r, fig.height=6, fig.width=8, warning=FALSE, message=FALSE}
plot_missing_upset(dia_qf, i = 'protein')
```

## Inspecting the number of peptides and proteins through the processing steps

Now that we have performed all the filtering steps and summarisation to protein-level abundances, it's helpful to visualise how many peptides/proteins were retained at each level of the processing. We can use the `biomasslmb::get_samples_present` and `biomasslmb::plot_samples_present` functions for this. First though, we need to decide which 'experiments' we want to plot and define a named character vector since the QFeatures names are not sufficiently clear by themselves

Below, we inspect the experiment names.

```{r}
names(dia_qf)
```

### Samples per peptide

We'll start by inspecting the number of peptides in each experiment. We therefore define a named character vector of all the peptide-level experiments. We set the row variable to be `Precursor.Id` so that we count the number of unique precursors.

```{r, fig.height=7, fig.width=7, fig.fullwidth=TRUE, fig.cap="Samples quantified for each precursor at each level of processing", out.width='75%'}
rename_cols <- c('All precursors' = 'precursors',
                 'Quantified, contaminants removed' = 'peptides_filtered',
                 'At most 75% missing values' = 'peptides_filtered_missing',
                 '>1 peptide per protein' = 'peptides_for_summarisation')

rowvars <- c('Precursor.Id')

samples_present <- get_samples_present(dia_qf[,,unname(rename_cols)], rowvars, rename_cols)
plot_samples_present(samples_present, rowvars, breaks = seq(5, 30, 5)) + ylab('Precursor')
```

### Samples per Protein

Next, we'll use the same functions to inspect the number of proteins at each level of processing. We need to supply an updated named character vector to include the `protein` experiment and set the row variable to be just `Protein.Group`.

```{r, fig.height=7, fig.width=7, fig.fullwidth=TRUE, fig.cap="Samples quantified for each protein at each level of processing", out.width='75%'}
rename_cols_prot <- c(rename_cols, 'Protein' = 'protein')

rowvars_prot <- c('Protein.Group')

samples_present <- get_samples_present(dia_qf, rowvars_prot, rename_cols_prot)
plot_samples_present(samples_present, rowvars_prot, breaks = seq(5, 30, 5))
```

From these two plots, we can see that the filtering to remove precursors with more than 75% missing values and to require \>1 precursor per protein removed relatively few precursors, but many proteins. Whether this is appropriate will depend on your data in hand.

```{r, eval = FALSE}
# Save file to package as data so it can be read back in in other vignettes
usethis::use_data(dia_qf, overwrite = TRUE)
```

# Reading Spectronaut output {#reading-spectronaut-output}

Everything above the `precursors` assay is DIA-NN specific: the file format, the column names and the Q-value fields. Everything below it is not. This section covers the difference, using a Spectronaut export from a total-lysate comparison between a vehicle control and an inhibitor treatment, three replicates each.

Spectronaut's 'BGS Factory Report' is long-format like DIA-NN's report, one row per precursor per run, but a plain TSV rather than parquet. The export shipped here has been reduced to the columns needed downstream, which is worth doing to your own exports too — the full report has well over a hundred columns and is correspondingly slow to read.

```{r}
spectronaut_inf <- system.file(
  "extdata", "spectronaut_report.tsv.gz", package = "biomasslmb")

spectronaut_df <- read.delim(spectronaut_inf)

colnames(spectronaut_df)
```

The correspondence with the DIA-NN fields used above is direct:

| Quantity | DIA-NN | Spectronaut |
|---|---|---|
| Run | `Run` | `R.FileName` |
| Precursor quantity | `Precursor.Quantity` | `FG.Quantity` |
| Precursor identifier | `Precursor.Id` | `EG.ModifiedSequence` + `FG.Charge` |
| Protein group | `Protein.Group` | `PG.ProteinGroups` |
| Precursor confidence | `Q.Value` | `EG.Qvalue` |
| Protein group confidence | `PG.Q.Value` | `PG.Qvalue` |

Spectronaut has no single precursor identifier column, so we build one, since `readQFeatures` needs a value that is unique within a run.

```{r}
spectronaut_df$Precursor.Id <- paste(
  spectronaut_df$EG.ModifiedSequence, spectronaut_df$FG.Charge, sep = '.')
```

The sample metadata is constructed the same way as before. Here the run names already encode the design, so we derive the columns from them rather than reading a separate file.

```{r}
spectronaut_design <- data.frame(runCol = sort(unique(spectronaut_df$R.FileName)))

spectronaut_design <- spectronaut_design %>%
  mutate(quantCols = 'FG.Quantity',
         Condition = sub('_.*', '', runCol),
         Replicate = sub('.*_', '', runCol))

knitr::kable(spectronaut_design)
```

```{r}
sn_qf <- readQFeatures(
  spectronaut_df,
  quantCols = "FG.Quantity",
  fnames = "Precursor.Id",
  runCol = "R.FileName",
  colData = spectronaut_design
)
```

Filtering on the two Q-value columns and joining the runs is the same operation as above, with the Spectronaut column names substituted. Note that there are no library Q-value fields to filter on: match-between-runs is a DIA-NN concept and Spectronaut's equivalent is handled within its own scoring.

```{r}
sn_qf <- sn_qf %>%
  filterFeatures(~ EG.Qvalue <= 0.01) %>%
  filterFeatures(~ PG.Qvalue <= 0.01)

sn_qf <- joinAssays(
  x = sn_qf, i = names(sn_qf), name = "precursors", fcol = "Precursor.Id")

colData(sn_qf[['precursors']]) <- colData(sn_qf)

dim(sn_qf[['precursors']])
```

From here the vignette applies unchanged, with `Protein.Group` replaced by `PG.ProteinGroups` wherever it appears as a `master_protein_col` or `fcol`. The one function that is DIA-NN specific is `filter_features_diann()`, which relies on DIA-NN's `Cont_` contaminant prefix and its column names; for Spectronaut, filter contaminants by matching `PG.ProteinAccessions` against the accessions from your contaminants FASTA.

```{r}
c(precursors = nrow(sn_qf[['precursors']]),
  protein_groups = length(unique(rowData(sn_qf[['precursors']])$PG.ProteinGroups)),
  percent_missing = round(100 * mean(is.na(assay(sn_qf[['precursors']]))), 1))
```

At the equivalent stage the DIA-NN dataset above is `r round(100*mean(is.na(assay(dia_qf[['precursors']]))), 1)`% missing against this one's `r round(100*mean(is.na(assay(sn_qf[['precursors']]))), 1)`%, though with `r ncol(dia_qf[['precursors']])` samples rather than `r ncol(sn_qf[['precursors']])`, so the two are not strictly comparable. The point is only that they are the same order of magnitude: pervasive precursor-level missingness is a property of label-free acquisition rather than of the search engine, and the reasoning in [handling missing values](handling_missing_values.html) applies to both.

```{r}
sessionInfo()
```
