---
title: "LFQ-DDA QC Peptide-level quantification and summarisation to protein-level abundance"
author: "Tom Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: biomasslmb.json
vignette: >
  %\VignetteIndexEntry{LFQ-DDA QC Peptide-level quantification and summarisation to protein-level abundance}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


Label-Free Quantification (LFQ) is the simplest form of quantitative bottom-up proteomics, in which different samples are quantified in separate MS runs. Quantification is either performed by Data-Dependent Aquisition (DDA), where the Mass Spectrometer triggers fragmentation of ions within a given m/z range with the aim being to focus attention of individual peptides separately, or Data-Independent Aquisition (DIA), where a much wider m/z range is used and a mix of peptides are co-fragmented and quantified simultaneously by deconvoluting the resultant complex spectra. Here, we will focus on just LFQ-DDA.

Since each sample is run separately, different peptides will be quantified in each sample and the peptide intensities may not be directly comparable between samples. The common solution to the higher burden of missing values is to use the 'match-between-runs' [@http://zotero.org/users/5634351/items/87YIDK69], or the functionally equivalent 'Minora' algorithm employed by Proteome Discoverer (PD). These algorithms use the observed retention times of MS1 ions which were successfully spectrum matched in one sample to identify the likely peptide sequence of MS1 ions that could not be spectrum matched in another sample. However, even with these algorithms enabled, DDA LFQ will still typically have many more missing values than labelled proteomics, e.g TMT.

Despite the pitfalls of LFQ, the data analysis is still relatively straightforward, though there are steps that need some careful consideration and quality control assessment.

# Load required packages

```{r, message=FALSE}
library(QFeatures)
library(biomasslmb)
library(ggplot2)
library(tidyr)
library(dplyr)
```

## The experimental design

`lfq_dda_pd_PeptideGroups.txt` is a file available from the `biomasslmb` package containing the peptide-level output from Proteome Discoverer (PD), using the Sequest HT search engine, for a real DDA LFQ TurboID (proximity biotinylation) experiment in mouse embryonic stem cells: a bait fusion protein is expressed and, when biotin is supplied, biotinylates nearby proteins, which are then enriched by streptavidin pulldown. We compare the biotin-labelled samples against unlabelled control pulldowns from the same bait (3 replicates each). It is truncated to a subset of 500 proteins for a manageable vignette.

```{r}
pep_inf <- system.file(
  "extdata", "lfq_dda_pd_PeptideGroups.txt",
  package = "biomasslmb"
)
```

# Defining the contaminant proteins

We need to remove contaminant proteins. This search was performed against the '0602_Universal Contaminants' database [@Frankenfield2022]. Below, we parse the contaminants fasta to extract the protein IDs.

```{r}

contaminant_fasta_inf <- system.file(
  "extdata", "0602_Universal_Contaminants.fasta.gz",
  package = "biomasslmb"
)

# Extract the protein IDs associated with each contaminant protein
contaminant_accessions <- biomasslmb::get_contaminant_fasta_accessions(contaminant_fasta_inf)

print(head(contaminant_accessions))
```

This PD export also includes PD's own `Contaminant` column, populated from the same search against the '0602_Universal Contaminants' database. `filter_features_pd_dda` uses this column automatically, in addition to the `contaminant_accessions` we've just extracted.

# Read in input data

We start by reading in quantitative proteomics data into a `QFeatures` object, which is the standard Bioconductor object for holding quantitative proteomics data. See [here](https://www.bioconductor.org/packages/release/bioc/html/QFeatures.html) for documentation about the `QFeatures` object.

Inspecting the column names for the abundances, we see the names are not particularly useful: they include the PD file ID (`F1`, `F2`, ...) rather than which sample is which.
```{r}
infdf <- read.delim(pep_inf)

abundance_cols_ix <- 13:18

colnames(infdf)[abundance_cols_ix]
```

Below, we update the column names for the abundance columns to reflect whether the sample was biotin-labelled or an unlabelled control, and the replicate number.
```{r}
# Clean up sample names
condition <- ifelse(grepl('biotin$', colnames(infdf)[abundance_cols_ix]), 'Biotin', 'Control')
colnames(infdf)[abundance_cols_ix] <- paste(condition, rep(1:3, times=2), sep='_')

colnames(infdf)[abundance_cols_ix]
```

Before we read in the data, we also want to define the experimental conditions for each sample.

```{r}
exp_design <- data.frame(quantCols=colnames(infdf)[abundance_cols_ix]) %>%
  separate(quantCols, sep='_', into=c('Condition', 'Replicate'), remove = FALSE)

knitr::kable(exp_design)
```

Now we can read the data into a `QFeatures` object
```{r}
# Read in peptide-level quantification from the DDA LFQ experiment (using QFeatures function)
lfq_qf <- readQFeatures(assayData = infdf,
                        quantCols = abundance_cols_ix,
                        colData = exp_design,
                        name = "peptides_raw")

```

Adding the `colData` to the peptide-level data too.

```{r}
colData(lfq_qf[['peptides_raw']]) <- colData(lfq_qf)


```

It's helpful to assess the retention time (RT) for the peptides and the
relationship with the delta mass for the precursor ion. The RT distribution declines steadily
after the first ~20 minutes, with a transient dip in peptide counts around 32-34 minutes, and
the vast majority of peptides fall within +/-5 ppm delta mass regardless of RT, with just a
modest number of higher-error outliers concentrated in the middle of the gradient.

```{r, warning=FALSE, fig.show='hold', fig.show="hold", out.width="50%"}
plot_rt_dist(lfq_qf[['peptides_raw']])
plot_rt_vs_delta(lfq_qf[['peptides_raw']])
```


# Filter peptides

We first perform routine filtering to remove PSMs that:

-   Could originate from contaminants. See `?filter_features_pd_dda` for
    further details, including the removal of 'associated cRAP'.
-   Don't have a unique master protein
-   Don't have any quantification values

```{r}
# Perform routine raw data filtering.
# - Remove PSMs from contaminant proteins
# - Remove PSMs where protein ID is empty or not unique

lfq_qf[['peptides_filtered']] <- filter_features_pd_dda(lfq_qf[['peptides_raw']],
                                                 contaminant_proteins=contaminant_accessions,
                                                 filter_contaminant=TRUE,
                                                 filter_associated_contaminant=TRUE,
                                                 unique_master=FALSE,
                                                 remove_no_quant = TRUE)

colData(lfq_qf[['peptides_filtered']]) <- colData(lfq_qf)
```


We will also remove peptides that are not rank 1 according to the search
engine

```{r}

lfq_qf <- lfq_qf %>%
filterFeatures(~ Search.Engine.Rank.by.Search.Engine.Sequest.HT == 1,
               i = "peptides_filtered")

biomasslmb:::message_parse(rowData(lfq_qf[['peptides_filtered']]),
                           'Master.Protein.Accessions',
                           "Removing peptides that are not rank 1")
```

For the summarisation, we need our quantification
values to be on a log-scale, so we log-transform below

```{r}
lfq_qf[['peptides_filtered']] <- QFeatures::logTransform(
  lfq_qf[['peptides_filtered']], base=2)
```

Next, we plot the peptide intensities, coloured by `Condition`.

```{r, warning=FALSE, fig.show='hold', fig.show="hold", out.width="50%"}

# Plot the peptide-level quantification distributions per sample
plot_quant(lfq_qf[['peptides_filtered']], log2transform=FALSE, method='density') +
  theme_biomasslmb() +
  aes(colour=Condition, group=sample) +
  xlab('Peptide abundance (log2)')
```

## Normalise (or not)

Since this is a proximity-labelling experiment, the biotin samples are expected to have a
genuinely different protein composition to the controls: the bait and its proximal partners
are specifically enriched, while the unlabelled controls capture mostly non-specific background
binding to the streptavidin beads. Standard normalisation assumes overall protein content is
equivalent between samples and would partially erode this real biological difference, so we do
not normalise here (see the [TMT QC](TMT_PSM_QC_Summarisation.html) vignette's MaxQuant
immunoprecipitation example for the same reasoning applied to an antibody pulldown).

## Summarising to protein-level abundances

For the summarisation to protein, we will use `robustSummary`. We do not need to remove
peptides with any missing values. However, we still don't want to retain
peptides with too many missing values, since these will not be very
informative in estimating the protein-level quantification.

Before deciding how to handle missing values, it's worth checking how much missingness there is,
and whether it's structured by `Condition`. `condition_miss_score` fits, for each peptide, a
logistic regression of missingness against `Condition`, and returns Tjur's R² as a per-feature
score: values near 1 mean a peptide's missingness pattern is well explained by `Condition`.
`condition_miss_index` aggregates these into a single dataset-level value between 0 and 1.

```{r}
miss_res <- condition_miss_score(lfq_qf, i = 'peptides_filtered', group_cols = 'Condition')
condition_miss_index(miss_res$summary)$index
```

A substantial fraction of peptides have missing values here (`r miss_res$global['n_features_informative']` out of `r miss_res$global['n_features_total']`), and the condition missingness index is well above zero, consistent with this being a proximity-labelling pulldown rather than a whole-proteome comparison: a genuine proximity partner is expected to be reliably detected in every biotin replicate but frequently missing from the (mostly background) control replicates. `plot_missing_upset` shows the most common patterns.

```{r, fig.height=6, fig.width=8, warning=FALSE, message=FALSE}
plot_missing_upset(lfq_qf, i='peptides_filtered')
```

Since a meaningful fraction of the missingness here is genuinely informative (a real proximity partner is absent from the control) rather than simply missing at random, we don't want to remove every peptide with any missing value. Since DDA LFQ typically has too many missing values to exclude peptides with missing values, the `MsCoreUtils::robustSummary` method should be preferred for summarisation to protein-level abundance. With `robustSummary`, we do not need to remove all peptides with missing values since the summarisation algorithm deals with them appropriately [@http://zotero.org/users/5634351/items/FZN3QTTZ]. However, we still don't want to retain peptides with too many missing values, since these will not be very informative in estimating the protein-level quantification. Here, we will retain peptides with at most 4/6 missing values (see the [Comparing protein summarisation approaches](summarisation_methods.html) vignette for a direct comparison with `sum`, using this same dataset).

```{r}

lfq_qf[['peptides_filtered_missing']] <- QFeatures::filterNA(
  lfq_qf[['peptides_filtered']], 4/6)

biomasslmb:::message_parse(rowData(lfq_qf[['peptides_filtered_missing']]),
                           'Master.Protein.Accessions',
                           "Removing peptides with more than 4/6 missing values")
```

Next, we remove peptides for proteins with fewer than 2 peptides.

```{r}
min_peps <- 2
lfq_qf[['peptides_for_summarisation']] <- filter_features_per_protein(
  lfq_qf[['peptides_filtered_missing']], min_features = min_peps)

biomasslmb:::message_parse(rowData(lfq_qf[['peptides_for_summarisation']]),
                           'Master.Protein.Accessions',
                           "Removing 'one-hit' wonders")
```

Now we can summarise with `robustSummary`

```{r, warning=FALSE}
set.seed(42)

# Aggregate to protein-level abundances (using QFeatures function)
lfq_qf <- aggregateFeatures(lfq_qf,
                            i = "peptides_for_summarisation",
                            fcol = "Master.Protein.Accessions",
                            name = "protein",
                            fun = MsCoreUtils::robustSummary,
                            maxit=10000)

colData(lfq_qf[['protein']]) <- colData(lfq_qf)
```

Prior to summaristaion, we removed PSMs from proteins with fewer than 2
peptides. However, since we left in peptides with missing values, it's
possible for some protein-level abundances to be derived from just a
single peptide. We can use the `get_protein_no_quant_mask` from
`biomasslmb` to identify where the protein abundances will be derived
from fewer than `n` features (peptides). We can then give this mask to
`mask_protein_level_quant` to replace these quantification values with
NA.

```{r, warning=FALSE, fig.show='hold', out.width="50%", fig.width=4, fig.height=4}
# plot = TRUE means we will also get a plot of the number of proteins quantified in each sample
protein_retain_mask <- biomasslmb::get_protein_no_quant_mask(
  lfq_qf[['peptides_for_summarisation']], min_features=min_peps, plot=TRUE)

lfq_qf[['protein']] <- biomasslmb::mask_protein_level_quant(
  lfq_qf[['protein']], protein_retain_mask)

```

## Re-inspecting missing values at protein-level

Now that we have protein-level abundances, we would like to re-inspect the missing values. Overall, we have `r round(nNA(lfq_qf[['protein']])$nNA$pNA*100, 3)` % missing values, with at most `r round(max(nNA(lfq_qf[['protein']])$nNAcols$pNA)*100, 3)` % missing values in any given sample. As at the peptide level, missingness is concentrated in the control samples: a real proximity partner of the bait is expected to be consistently quantified in the biotin samples but frequently missing from the (background-only) controls.

```{r, fig.height=6, fig.width=8, warning=FALSE, message=FALSE}

plot_missing_upset(lfq_qf, i='protein' )
```


## Inspecting the number of peptides and proteins through the processing steps

Now that we have performed all the filtering steps and summarisation to protein-level abundances, it's helpful to visualise how many peptides/proteins were retained at each level of the processing. We can use the `biomasslmb::get_samples_present` and `biomasslmb::plot_samples_present` functions for this. First though, we need to decide which 'experiments' we want to plot and define a named character vector since the QFeatures names are not sufficiently clear by themselves

Below, we inspect the experiment names.

```{r}
names(lfq_qf)
```

### Samples per peptide

We'll start by inspecting the number of peptides in each experiment. We therefore define a named character vector all the peptides-level experiments. We set the row variables to be `Annotated.Sequence` and `Modifications` so that we count the number of unique peptides.

```{r, fig.height=7, , fig.width=7, fig.fullwidth=TRUE, fig.cap="Samples quantified for each PSM at each level of processing", out.width='75%'}


rename_cols <- c('All peptides' = 'peptides_raw' ,
                 'Quantified, contaminants removed' = 'peptides_filtered',
                 'No missing values' = 'peptides_filtered_missing',
                 '>1 PSM per protein' = 'peptides_for_summarisation')

rowvars <- c('Annotated.Sequence', 'Modifications')

samples_present <- get_samples_present(lfq_qf[,,unname(rename_cols)], rowvars, rename_cols)
plot_samples_present(samples_present, rowvars, breaks=seq(2,10,2)) + ylab('PSM')



```

### Samples per Protein

Next, we'll use the same functions to inspect the number of proteins at each level of processing. We need to supply an updated named character vector to include the `protein` experiment and set the row variables to be just the `Master.Protein.Accesions` column.

```{r, fig.height=7, , fig.width=7, fig.fullwidth=TRUE, fig.cap="Samples quantified for each protein at each level of processing", out.width='75%'}

rename_cols_prot <- c(rename_cols, 'Protein'='protein')

rowvars_prot <- c('Master.Protein.Accessions')

samples_present <- get_samples_present(lfq_qf, rowvars_prot, rename_cols_prot)
plot_samples_present(samples_present, rowvars_prot, breaks=seq(2,10,2))
```

From these two plots, we can see that the filtering to ensure that all proteins have \>1 peptides removed only a few peptides, but more proteins. Whether this is appropriate will depend on your data in hand.


```{r, eval = FALSE}
# Save file to package as data so it can be read back in in other vignettes
usethis::use_data(lfq_qf, overwrite=TRUE)
```

```{r}
sessionInfo()
```
