---
title: "LFQ-DIA workflow: precursor QC and protein summarisation"
author: "Tom Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: biomasslmb.json
vignette: >
  %\VignetteIndexEntry{LFQ-DIA workflow: precursor QC and protein summarisation}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Label-Free Quantification (LFQ) is the simplest form of quantitative proteomics, in which different samples are quantified in separate MS runs. Quantification is either performed by Data-Dependent Acquisition (DDA), where the Mass Spectrometer triggers fragmentation of ions within a given m/z range with the aim being to focus attention of individual peptides separately, or Data-Independent Acquisition (DIA), where a much wider m/z range is used and a mix of peptides are co-fragmented and quantified simultaneously by deconvoluting the resultant complex spectra. This article covers LFQ-DIA; DDA is a different enough problem to have its own, in [LFQ-DDA peptide QC](LFQ_DDA_Peptide_QC_Summarisation.html).

DIA data is commonly searched with [DIA-NN](https://github.com/vdemichev/DiaNN), which reports results at multiple levels. This article starts from the precursor level — a specific peptide sequence, with modifications, at a specific charge state — because it is the level at which the QC decisions below can still be made. Since DIA co-fragments many precursors at once, DIA-NN infers identifications either from a project-specific spectral library or from a library predicted in-silico from the sequence database, then rescores them, optionally using match-between-runs (MBR) to borrow confidence across runs. None of that removes the need to filter to confident identifications, remove contaminants, handle missing values and summarise to protein-level abundance before any statistical analysis.

This vignette works through one typical experiment from end to end: a whole-proteome comparison between groups, with no enrichment step and a single acquisition batch. That is the common case, and the choices made below are the conventional ones for it. An enrichment experiment — an IP, BioID or TurboID pulldown — needs different reasoning about normalisation and about missing values, and is covered in [enrichment designs](interactome_designs.html).

If this is your first analysis with the package, [getting started](biomasslmb.html) explains which route applies to your experiment and shows a complete one in forty lines, and [working with QFeatures objects](qfeatures_objects.html) covers the object every step below operates on. Annotations are worth retrieving before any of this — see [protein annotation](protein_annotation.html).

The data used in this vignette is a real DIA-NN `report.parquet` output for a plasma proteomics case series comparing Mpox (MPXV) infection, COVID-19, and healthy controls [@Wang2022]. It is also used, with accompanying teaching material, in the [Proteomics Data Analysis Using R](https://github.com/bioinformatics-core-shared-training/Proteomics_stats_R) course. Unlike the truncated demonstration files used elsewhere in `biomasslmb`, this dataset is used here in full - it is small enough not to need subsetting

Spectronaut is the other search engine in common use for DIA data. Only the reading step differs: [reading Spectronaut output](#reading-spectronaut-output) at the end of this vignette shows how to reach the same `precursors` assay, after which every step below applies unchanged.

# Load required packages

```{r, message=FALSE}
library(QFeatures)
library(biomasslmb)
library(ggplot2)
library(tidyr)
library(dplyr)
library(arrow)
library(readxl)
```

# Defining the contaminant proteins

Contaminant proteins have to be removed. This search was performed against a version of the Hao lab's '0602 Universal Contaminants' database [@Frankenfield2022] with the Fetal Bovine Serum-derived entries removed, since they are not relevant to a human plasma sample. The standard database bundled with `biomasslmb` is used below instead — the FBS-specific entries it additionally contains are never observed in this dataset, so filtering behaves identically either way.

```{r}
contaminant_fasta_inf <- system.file(
  "extdata", "0602_Universal_Contaminants.fasta.gz",
  package = "biomasslmb"
)

# Extract the protein IDs associated with each contaminant protein
contaminant_accessions <- biomasslmb::get_contaminant_fasta_accessions(contaminant_fasta_inf)

print(head(contaminant_accessions))
```

DIA-NN was run with this contaminants database included directly in the search FASTA, so contaminant entries are named with a `Cont_` prefix that is carried straight through into the `Protein.Ids` and `Protein.Group` columns of the DIA-NN output. `filter_features_diann` uses this prefix automatically, in addition to the `contaminant_accessions` extracted above.

The prefix and the accession list are separate defences, and a search database whose contaminant entries were never renamed provides neither. [Contaminants and protein FDR](gotcha_contaminants_and_FDR.html) measures what each one catches and what is left behind when they do not match.

# Read in input data

The data go into a `QFeatures` object, the standard Bioconductor container for quantitative proteomics data. See [here](https://www.bioconductor.org/packages/release/bioc/html/QFeatures.html) for documentation about the `QFeatures` object.

Unlike PD's peptide-level export, DIA-NN's `report.parquet` is a **long-format** table: each row is a single precursor-sample combination, rather than one row per precursor across all samples. `arrow::read_parquet` reads it.

```{r}
diann_report_inf <- system.file(
  "extdata", "monkeypox_plasma_proteomes.parquet",
  package = "biomasslmb"
)

diann_df <- arrow::read_parquet(diann_report_inf)

dim(diann_df)
```

Sample metadata (which group, sex and age band each run belongs to) is provided separately, in `monkeypox_metadata.xlsx`.

```{r}
metadata_inf <- system.file(
  "extdata", "monkeypox_metadata.xlsx",
  package = "biomasslmb"
)

sample_metadata <- readxl::read_excel(metadata_inf)

print(sample_metadata)
```

The `Run` column in the DIA-NN output inherits the full raw file path, pasting together the `.wiff2` folder and `.wiff.scan` file names. Stripping that back to match the `sample_id` field in the metadata allows the concise `runCol` label to be merged in, which serves as the sample identifier from here on.

```{r}
diann_df[, "Run"] %>% unique() %>% head(2)

diann_df <- diann_df %>%
  mutate(Run = sub('\\..*', '', Run)) %>%
  merge(sample_metadata[, c("sample_id", "runCol")], by.x = "Run", by.y = "sample_id")

diann_df[, "Run"] %>% unique() %>% head(2)
```

`readQFeaturesFromDIANN` reads the long-format table in, creating one `SummarizedExperiment` per run. The split matters: the per-run quality metrics in `rowData` (`Q.Value`, `FWHM` and the rest) are run-specific, and combining the runs immediately would discard them before they can be filtered on. They are used below, and the runs joined afterwards.

```{r}
dia_qf <- readQFeaturesFromDIANN(
  diann_df,
  quantCols = "Precursor.Quantity",
  fnames = "Precursor.Id",
  runCol = "runCol",
  colData = sample_metadata
)

dia_qf
```

The retention time (RT) distribution and its relationship with chromatographic peak width (FWHM, the Full-Width at Half Maximum) are worth assessing before anything is filtered, because a chromatography problem is cheaper to find now than after summarisation. Narrow, symmetrical peaks are what a good separation looks like, and FWHM rising systematically with RT would indicate the later part of the gradient is resolving poorly.

```{r, warning=FALSE, fig.show='hold', fig.width=5, fig.height=4}
rbindRowData(dia_qf, i = names(dia_qf)) %>%
  data.frame() %>%
  ggplot(aes(RT, FWHM)) +
  geom_hex(bins = 60) +
  theme_biomasslmb(aspect_square = FALSE) +
  labs(x = 'Retention time (min)', y = 'Peak width, FWHM (min)')
```

# Filter precursors

DIA-NN assigns a Q-value to each precursor and protein group identification, representing the local false discovery rate (FDR) at that score threshold. This data was processed with match-between-runs enabled, which is why both the run-level and library-level Q-values are filtered (`Q.Value`/`PG.Q.Value` and `Lib.Q.Value`/`Lib.PG.Q.Value` respectively), at a standard 1% FDR threshold. Filtering only the run-level pair would accept identifications that MBR transferred on weak library-level evidence. Passing no `i` argument to `filterFeatures` applies the filter to every per-run set at once.

```{r}
dia_qf <- dia_qf %>%
  filterFeatures(~ Q.Value <= 0.01) %>%
  filterFeatures(~ PG.Q.Value <= 0.01) %>%
  filterFeatures(~ Lib.Q.Value <= 0.01) %>%
  filterFeatures(~ Lib.PG.Q.Value <= 0.01)
```

The per-run sets are then combined into a single `"precursors"` set with `joinAssays`, aligning features by `Precursor.Id`. A precursor not detected in a given run becomes `NA` for that sample. Only `rowData` columns that are shared across runs and have consistent values survive the join; the run-specific quality metrics already filtered on (`Q.Value`, `RT`) are dropped, having served their purpose.

```{r}
dia_qf <- joinAssays(
  x = dia_qf,
  i = names(dia_qf),
  name = "precursors",
  fcol = "Precursor.Id"
)

# The per-run sets are no longer needed now they're joined; removing them
# keeps the QFeatures object focused on the combined set for the rest of
# this vignette.
dia_qf <- removeAssay(dia_qf, i = which(names(dia_qf) != "precursors"))

dia_qf <- sync_coldata(dia_qf, 'precursors')
```

Routine filtering removes precursors that:

-   Could originate from contaminants. See `?filter_features_diann` for further details, including the removal of contaminant-associated proteins.
-   Have more than one accession in `Protein.Group`, meaning DIA-NN could not resolve which protein the precursor came from.
-   Don't have any quantification values

`filter_features_diann()` applies all three by default, so the call below passes none of them. `unique_master` asks whether the search engine settled on a single protein identity. The separate `proteotypic` argument asks whether the peptide sequence occurs in only one protein at all — a feature can have one unambiguous master protein and still be shared with the other members of that protein group. See [peptides are not proteins](gotcha_peptide_to_protein.html).

```{r}
dia_qf[['peptides_filtered']] <- biomasslmb::filter_features_diann(
  dia_qf[['precursors']],
  contaminant_proteins = contaminant_accessions
)

dia_qf <- sync_coldata(dia_qf, 'peptides_filtered')
```

Summarisation needs the quantification values on a log scale, so they are transformed here.

```{r}
dia_qf[['peptides_filtered']] <- QFeatures::logTransform(
  dia_qf[['peptides_filtered']], base = 2)
```

The precursor intensity distributions, coloured by `group`:

```{r, warning=FALSE, fig.show='hold', fig.width=7, fig.height=5, out.width="75%"}
biomasslmb::plot_quant(dia_qf[['peptides_filtered']], log2transform = FALSE, method = 'density') +
  theme_biomasslmb(aspect_square = FALSE) +
  aes(colour = group, group = sample) +
  xlab('Precursor abundance (log2)')
```

## Normalise

This is a whole-plasma-proteome comparison between clinical groups, in which overall protein content is not expected to differ systematically between groups, so any offset between the distributions above is technical. `diff.median` normalisation with `QFeatures::normalize` shifts each sample to a common median and removes it. The assumption is worth stating, because an [enrichment design](interactome_designs.html) breaks it — there the conditions are expected to differ in composition, and normalising to a common median erodes the signal being measured.

```{r}
dia_qf <- QFeatures::normalize(
  dia_qf, i = 'peptides_filtered', name = 'peptides_filtered_norm', method = 'diff.median')

dia_qf <- sync_coldata(dia_qf, 'peptides_filtered_norm')
```

```{r, warning=FALSE, fig.show='hold', fig.width=7, fig.height=5, out.width="75%"}
biomasslmb::plot_quant(dia_qf[['peptides_filtered_norm']], log2transform = FALSE, method = 'density') +
  theme_biomasslmb(aspect_square = FALSE) +
  aes(colour = group, group = sample) +
  xlab('Precursor abundance (log2, normalised)')
```

## Summarising to protein-level abundances

Summarisation takes the protein group assignment at face value: every precursor contributes to exactly one protein group. That assignment is an inference rather than a measurement, and for proteins whose peptides are largely shared with their homologues it is not a reliable one. [Peptides are not proteins](gotcha_peptide_to_protein.html) covers how the assignment is made and how to tell which of your proteins it can support.

Before deciding how to handle missing values, it's worth checking how much missingness there is,
and whether it's structured by `group`. `condition_miss_score` fits, for each peptide, a
logistic regression of missingness against `group`, and returns Tjur's R² as a per-feature
score: values near 1 mean a peptide's missingness pattern is well explained by `group`.
`condition_miss_index` aggregates these into a single dataset-level value between 0 and 1.

```{r}
miss_res <- condition_miss_score(dia_qf, i = 'peptides_filtered_norm', group_cols = 'group')
condition_miss_index(miss_res$summary)$index
```

**Missingness is substantial but unstructured by group.** `r miss_res$global['n_features_informative']` of `r miss_res$global['n_features_total']` precursors have at least one missing value, which is typical of DIA-NN precursor-level output, and yet the condition missingness index is close to zero. That combination is what distinguishes this from a pulldown, where missingness is explained by condition by construction and discarding it would discard the finding. `plot_missing_upset` shows the most common patterns.

```{r, fig.height=6, fig.width=8, warning=FALSE, message=FALSE}
plot_missing_upset(dia_qf, i = 'peptides_filtered_norm')
```

At this level of missingness, requiring complete precursors would discard most of the data, which rules out summing and makes `MsCoreUtils::robustSummary` the appropriate estimator: it summarises accurately without complete features [@http://zotero.org/users/5634351/items/FZN3QTTZ]. Precursors missing from almost every sample still contribute little beyond noise, so a threshold is applied first — retaining those observed in at least 25% of the 31 samples (`pNA` of at most 0.75), matching the threshold applied to this dataset in the [Proteomics Data Analysis Using R](https://github.com/bioinformatics-core-shared-training/Proteomics_stats_R) course. Set it tighter and more of the plasma proteome falls below the two-precursor rule imposed next; set it looser and protein estimates rest on precursors seen a handful of times.

```{r}
dia_qf[['peptides_filtered_missing']] <- QFeatures::filterNA(
  dia_qf[['peptides_filtered_norm']], pNA = 0.75)

message_parse(rowData(dia_qf[['peptides_filtered_missing']]),
                           'Protein.Group',
                           "Removing peptides with > 75% missing values")
```

Precursors belonging to proteins with fewer than 2 precursors are removed next. This is routine practice and worth being explicit about, because it is doing more than tidying.

A protein quantified from a single precursor has no internal replication: there is no second measurement to disagree with the first, so nothing about it can be checked. Any error in that one precursor — a misassigned peptide, interference, a bad integration — passes straight through to the protein-level value with nothing to average it out. Its missingness pattern is also the missingness pattern of one precursor rather than of the protein, which matters for everything downstream that reasons about missing values.

The cost is real, and asymmetric: single-precursor proteins are disproportionately the low-abundance ones, so this filter removes preferentially from the part of the proteome that was hardest to measure. `min_features` is the dial. Raising it buys confidence and loses coverage; there is no threshold that is correct independent of what the result will be used for.

```{r}
min_peps <- 2
dia_qf[['peptides_for_summarisation']] <- filter_features_per_protein(
  dia_qf[['peptides_filtered_missing']], min_features = min_peps, master_protein_col = 'Protein.Group')

message_parse(rowData(dia_qf[['peptides_for_summarisation']]),
                           'Protein.Group',
                           "Removing 'one-hit' wonders")
```

Summarising with `robustSummary`:

```{r, warning=FALSE}
set.seed(42)

dia_qf <- aggregateFeatures(dia_qf,
                            i = "peptides_for_summarisation",
                            fcol = "Protein.Group",
                            name = "protein",
                            fun = MsCoreUtils::robustSummary,
                            maxit = 10000)

dia_qf <- sync_coldata(dia_qf, 'protein')
```

The two-precursor filter above was applied per protein, not per sample, and precursors with missing values were deliberately kept — so in any given sample a protein can still end up summarised from a single precursor. Nothing in the output marks these values as weaker than the rest. `get_protein_no_quant_mask()` finds where a protein abundance rests on fewer than `n` precursors, and `mask_protein_level_quant()` replaces those values with `NA`.

```{r, warning=FALSE, fig.show='hold', out.width="50%", fig.width=8, fig.height=4}
protein_retain_mask <- biomasslmb::get_protein_no_quant_mask(
  dia_qf[['peptides_for_summarisation']], min_features = min_peps,
  master_protein_col = 'Protein.Group')

dia_qf[['protein']] <- biomasslmb::mask_protein_level_quant(
  dia_qf[['protein']], protein_retain_mask)
```

## Re-inspecting missing values at protein-level

Masking changes the missingness picture, so it is worth re-reading at protein level. `r round(nNA(dia_qf[['protein']])$nNA$pNA*100, 3)`% of protein values are missing overall, and at most `r round(max(nNA(dia_qf[['protein']])$nNAcols$pNA)*100, 3)`% in any one sample — far less than at precursor level, because a protein only loses a sample when every one of its precursors does.

```{r, fig.height=6, fig.width=8, warning=FALSE, message=FALSE}
plot_missing_upset(dia_qf, i = 'protein')
```

## Inspecting the number of peptides and proteins through the processing steps

Each filtering step above removed something, and the totals are easier to judge together than one message at a time. `get_samples_present()` and `plot_samples_present()` show how many precursors and proteins survived each stage, and in how many samples. Both take a named character vector selecting the assays to plot, since the `QFeatures` names are not self-explanatory.

The assay names to choose from:

```{r}
names(dia_qf)
```

### Samples per peptide

Precursors first. The row variable is `Precursor.Id`, so that the count is of unique precursors rather than of rows.

```{r, fig.height=7, fig.width=7, fig.fullwidth=TRUE, fig.cap="Samples quantified for each precursor at each level of processing", out.width='75%'}
rename_cols <- c('All precursors' = 'precursors',
                 'Quantified, contaminants removed' = 'peptides_filtered',
                 'At most 75% missing values' = 'peptides_filtered_missing',
                 '>1 peptide per protein' = 'peptides_for_summarisation')

rowvars <- c('Precursor.Id')

samples_present <- get_samples_present(dia_qf[,,unname(rename_cols)], rowvars, rename_cols)
plot_samples_present(samples_present, rowvars, breaks = seq(5, 30, 5)) + ylab('Precursor')
```

### Samples per Protein

Then proteins, from the same functions: the named vector gains the `protein` assay, and the row variable becomes `Protein.Group` alone.

```{r, fig.height=7, fig.width=7, fig.fullwidth=TRUE, fig.cap="Samples quantified for each protein at each level of processing", out.width='75%'}
rename_cols_prot <- c(rename_cols, 'Protein' = 'protein')

rowvars_prot <- c('Protein.Group')

samples_present <- get_samples_present(dia_qf, rowvars_prot, rename_cols_prot)
plot_samples_present(samples_present, rowvars_prot, breaks = seq(5, 30, 5))
```

**The missingness and two-precursor filters are cheap in precursors and expensive in proteins.** Few precursors go, because most belong to proteins that have several; many more proteins go, because a protein resting on one sparsely-observed precursor loses its only evidence. Those are the proteins whose quantification would have been least reliable, so the trade is usually worth making — but in plasma it falls hardest on the low-abundance fraction, which is often exactly where the interesting biology is.

```{r, eval = FALSE}
# Save file to package as data so it can be read back in in other vignettes
usethis::use_data(dia_qf, overwrite = TRUE)
```

# Reading Spectronaut output {#reading-spectronaut-output}

Everything above the `precursors` assay is DIA-NN specific: the file format, the column names and the Q-value fields. Everything below it is not. This section covers the difference, using a Spectronaut export from a total-lysate comparison between a vehicle control and an inhibitor treatment, three replicates each.

Spectronaut's 'BGS Factory Report' is long-format like DIA-NN's report, one row per precursor per run, but a plain TSV rather than parquet. The export shipped here has been reduced to the columns needed downstream, which is worth doing to your own exports too — the full report has well over a hundred columns and is correspondingly slow to read.

```{r}
spectronaut_inf <- system.file(
  "extdata", "spectronaut_report.tsv.gz", package = "biomasslmb")

spectronaut_df <- read.delim(spectronaut_inf)

colnames(spectronaut_df)
```

The correspondence with the DIA-NN fields used above is direct:

| Quantity | DIA-NN | Spectronaut |
|---|---|---|
| Run | `Run` | `R.FileName` |
| Precursor quantity | `Precursor.Quantity` | `FG.Quantity` |
| Precursor identifier | `Precursor.Id` | `EG.ModifiedSequence` + `FG.Charge` |
| Protein group | `Protein.Group` | `PG.ProteinGroups` |
| Precursor confidence | `Q.Value` | `EG.Qvalue` |
| Protein group confidence | `PG.Q.Value` | `PG.Qvalue` |

Spectronaut has no single precursor identifier column, so one has to be built: `readQFeatures` needs a value that is unique within a run.

```{r}
spectronaut_df$Precursor.Id <- paste(
  spectronaut_df$EG.ModifiedSequence, spectronaut_df$FG.Charge, sep = '.')
```

The sample metadata is constructed the same way as before. Here the run names already encode the design, so the columns are derived from them rather than from a separate file.

```{r}
spectronaut_design <- data.frame(runCol = sort(unique(spectronaut_df$R.FileName)))

spectronaut_design <- spectronaut_design %>%
  mutate(quantCols = 'FG.Quantity',
         Condition = sub('_.*', '', runCol),
         Replicate = sub('.*_', '', runCol))

knitr::kable(spectronaut_design)
```

```{r}
sn_qf <- readQFeatures(
  spectronaut_df,
  quantCols = "FG.Quantity",
  fnames = "Precursor.Id",
  runCol = "R.FileName",
  colData = spectronaut_design
)
```

Filtering on the two Q-value columns and joining the runs is the same operation as above, with the Spectronaut column names substituted. Note that there are no library Q-value fields to filter on: match-between-runs is a DIA-NN concept and Spectronaut's equivalent is handled within its own scoring.

```{r}
sn_qf <- sn_qf %>%
  filterFeatures(~ EG.Qvalue <= 0.01) %>%
  filterFeatures(~ PG.Qvalue <= 0.01)

sn_qf <- joinAssays(
  x = sn_qf, i = names(sn_qf), name = "precursors", fcol = "Precursor.Id")

sn_qf <- sync_coldata(sn_qf, 'precursors')

dim(sn_qf[['precursors']])
```

From here the vignette applies unchanged, with `Protein.Group` replaced by `PG.ProteinGroups` wherever it appears as a `master_protein_col` or `fcol`. The one function that is DIA-NN specific is `filter_features_diann()`, which relies on DIA-NN's `Cont_` contaminant prefix and its column names; for Spectronaut, filter contaminants by matching `PG.ProteinAccessions` against the accessions from your contaminants FASTA.

```{r}
c(precursors = nrow(sn_qf[['precursors']]),
  protein_groups = length(unique(rowData(sn_qf[['precursors']])$PG.ProteinGroups)),
  percent_missing = round(100 * mean(is.na(assay(sn_qf[['precursors']]))), 1))
```

At the equivalent stage the DIA-NN dataset above is `r round(100*mean(is.na(assay(dia_qf[['precursors']]))), 1)`% missing against this one's `r round(100*mean(is.na(assay(sn_qf[['precursors']]))), 1)`%, though with `r ncol(dia_qf[['precursors']])` samples rather than `r ncol(sn_qf[['precursors']])`, so the two are not strictly comparable. The point is only that they are the same order of magnitude: pervasive precursor-level missingness is a property of label-free acquisition rather than of the search engine, and the reasoning in [handling missing values](handling_missing_values.html) applies to both.

# Where to go next

- [Data exploration and statistical testing](exploration_and_statistical_testing.html)
  picks up from the protein-level abundances produced here, and works through this
  same dataset: checking the experiment worked, deciding which proteins can be
  tested, and testing them.
- [Handling missing values](handling_missing_values.html) covers the `filterNA`
  threshold applied above in full. DIA carries enough missingness for that choice
  to matter, and for the missingness-aware alternatives to testing to be worth
  considering.
- [Choosing a summarisation method](summarisation_methods.html) explains why
  `robustSummary` rather than summing, and measures the difference.
- If this was an enrichment experiment rather than a whole-proteome comparison,
  [enrichment designs](interactome_designs.html) covers the normalisation and
  missing-value reasoning that changes.

# Getting help

If your experiment does not fit what this article assumes, or a result looks
wrong, please get in touch with Tom Smith (tsmith@mrclmb.ac.uk) rather than
guessing — it is far easier to help while an analysis is in progress than to
unpick a decision afterwards, and easier still before the samples are run.
[Getting started](biomasslmb.html) sets out which article applies to which
experiment.

```{r}
sessionInfo()
```
