---
title: "PTM designs: site-level quantification and normalisation"
author: "Tom Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: biomasslmb.json
vignette: >
  %\VignetteIndexEntry{PTM designs: site-level quantification and normalisation}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

A post-translational modification (PTM) experiment asks a different question from a whole-proteome one. The quantity of interest is not how much of a protein there is, but how much of it carries a modification at a particular residue. Everything below follows from that difference.

Because modified peptides are a small fraction of the peptides in a digest, they have to be enriched before they can be quantified with any depth. A common design is to TMT label every sample once, pool the labelled material, and then split the pool: one part is enriched for the modification, the other is left alone. The two parts are acquired as separate runs, giving two datasets over the same samples — an **enriched** one that reports modified peptides, and a **total** one that reports protein abundance.

The alternative is to enrich each sample separately and quantify label-free, and the reason not to is that the enrichment is the most variable step in the workflow. Recovery from an immobilised metal affinity or titanium dioxide column depends on binding capacity, wash stringency and elution, and none of those repeat exactly between preparations. Enriching separately imprints that variability on the samples individually, where nothing in the data distinguishes it from biology. Labelling first and pooling moves the enrichment downstream of the point at which the samples become one tube, so a single enrichment is applied to all of them and its idiosyncrasies land on every channel alike.

Two further things follow from pooling. Modified peptides sit low in the dynamic range and are identified less reproducibly than unmodified ones, so a label-free DDA experiment loses many sites to missing values in a subset of samples; within a plex, a site identified once is quantified in every channel from the same spectrum. And the total fraction is split from the pool after labelling, which is what later makes it usable as a normalisation reference: it is the same labelled material, so it carries the same loading differences as the enriched fraction.

The costs are the usual isobaric ones. The plex sets a ceiling on how many samples can share an enrichment, and a design that exceeds it needs bridge channels and the treatment in [multi-plex TMT](TMT_multiplex.html). Co-isolation compresses fold changes, which SPS MS3 [@http://zotero.org/users/5634351/items/ZK69WYZ2] and the filtering below limit, but cannot completely resolve. Label-free is a reasonable choice where the enrichment can be tightly controlled, and DIA acquisition answers the missing value argument, but neither recovers the shared enrichment.

With TMT-label, pool then enrich experiment design, there are three important considerations which we cover here:

- **Which residue is modified matters**, so the search engine's localisation confidence has to be assessed and the site placed within the protein, not just within the peptide.
- **The enriched fraction cannot be normalised against itself.** A treatment that changes phosphorylation globally is the signal, and median-centring the enriched data would remove it. The total fraction supplies the reference instead.
- **A site can change because its protein changed.** Dividing the site out by its protein separates the two, and the answer is not always the same as the site abundance.

This vignette works through a phosphoproteomics experiment searched with Proteome Discoverer (PD). The principles apply to any enriched PTM; only the search-engine columns differ. MaxQuant's equivalents are `filter_maxquant_ptm()` and `add_filter_ptm_pos_rowdata_mq()`, which take the same role as `parse_PTM_scores_pd()` below.

Everything up to the localisation step is the routine PSM processing covered in [TMT QC and summarisation](TMT_PSM_QC_Summarisation.html), so it is run here with less commentary.

# Load required packages

```{r, message=FALSE}
library(QFeatures)
library(biomasslmb)
library(ggplot2)
library(dplyr)
library(limma)
```

# The experimental design

`psm_tmt_phospho`, `psm_tmt_phospho_total` and `tmt_phospho_design` are datasets available from the `biomasslmb` package, derived from a real TMTpro experiment in mouse fibroblasts. A drug treatment is compared against a vehicle control at two timepoints, with four replicates of each combination, all labelled in a single plex. `psm_tmt_phospho` is the PD PSM-level output for the phospho-enriched fraction and `psm_tmt_phospho_total` for the matched total fraction, both truncated to a subset of proteins for a manageable vignette.

```{r}
knitr::kable(tmt_phospho_design)
```

The two fractions share this design, because they are two acquisitions of the same labelled pool. The TMT tag that carried a given sample is the same in both.

# Defining the contaminant proteins

As for any experiment, we need the contaminant accessions to filter against. This search used the '0602_Universal Contaminants' database [@Frankenfield2022].

```{r}
contaminant_fasta_inf <- system.file(
  "extdata", "0602_Universal_Contaminants.fasta.gz",
  package = "biomasslmb"
)

contaminant_accessions <- get_contaminant_fasta_accessions(contaminant_fasta_inf)
contaminant_accessions <- c(contaminant_accessions,
                            sub('^Cont_', '', contaminant_accessions))
```

# Read in and filter both fractions

The two fractions are read into separate `QFeatures` objects. They cannot share one, because they hold different features: modified peptides in one, all peptides in the other. Keeping them apart also keeps it obvious which assay is which when they are combined at the end.

The filtering applied here is the same for both, and follows the PSM-level processing in the [TMT PSM QC and summarisation](TMT_PSM_QC_Summarisation.html) vignette: drop contaminants and PSMs without a unique master protein, then drop PSMs with high co-isolation, low signal:noise, or a search engine rank below 1.

```{r, message=FALSE, warning=FALSE}
read_and_filter <- function(infdf) {
  qf <- readQFeatures(assayData = infdf,
                      colData = tmt_phospho_design,
                      name = 'psm_raw')
  qf <- sync_coldata(qf, 'psm_raw')

  # A more accurate average S:N ratio value than PD reports
  qf[['psm_raw']] <- update_average_sn(qf[['psm_raw']])

  qf[['psm_filtered']] <- filter_features_pd_dda(
    qf[['psm_raw']],
    contaminant_proteins = contaminant_accessions,
    filter_contaminant = TRUE,
    filter_associated_contaminant = TRUE,
    unique_master = TRUE)

  qf[['psm_quality']] <- filter_TMT_PSMs(
    qf[['psm_filtered']], inter_thresh = 50, sn_thresh = 5)

  filterFeatures(qf, ~ Rank == 1, i = 'psm_quality')
}

phospho <- read_and_filter(psm_tmt_phospho)
total <- read_and_filter(psm_tmt_phospho_total)
```

```{r}
data.frame(
  fraction = c('Phospho-enriched', 'Total'),
  raw = c(nrow(phospho[['psm_raw']]), nrow(total[['psm_raw']])),
  filtered = c(nrow(phospho[['psm_quality']]), nrow(total[['psm_quality']])))
```

# Localising the modification

A search engine reports which peptide was identified and which residues carry the modification, but the second of those is much less certain than the first. A peptide with several serines and one phosphate can often be explained about equally well by a phosphate on any of them: the fragment ions that would distinguish the possibilities may not have been observed. PD's ptmRS node scores each candidate site, and `parse_PTM_scores_pd()` reads those scores out of the `ptmRS.Best.Site.Probabilities` column.

`threshold` sets the minimum score a site needs. If any site on a peptide falls below it, the modifications on that peptide are all discarded rather than partially kept — a peptide whose phosphate could be on either of two residues cannot be assigned to either. It is not silent about the region, though, and [peptides that cannot be localised](#peptides-that-cannot-be-localised) below recovers what it does say.

```{r}
phospho[['psm_localised']] <- parse_PTM_scores_pd(
  phospho[['psm_quality']], threshold = 75)
```

The log distinguishes features from sites, which matters when peptides carry more than one modification. `parse_PTM_scores_pd()` annotates rather than filters, writing empty values where nothing passed, so the failures still need removing.

```{r}
phospho[['psm_localised']] <- phospho[['psm_localised']][
  grepl('Phospho', rowData(phospho[['psm_localised']])$ptms), ]

nrow(phospho[['psm_localised']])
```

A threshold of 75 is a common choice and is the one used here; 95 is stricter and is worth considering when the question rests on a single site. The cost is one-sided — raising it discards peptides, it does not add any — so the appropriate value depends on whether you would rather lose sites or believe some that are misplaced.

# Placing sites within the protein

The positions reported so far are positions within the peptide. Two PSMs of the same site reached by different missed cleavages will disagree about it, and no two proteins can be compared on it. `add_ptm_positions()` converts them to positions within the protein by re-digesting the protein sequences and locating each peptide.

That needs the sequences. In a real analysis you would fetch them for your master proteins with `make_fasta()`; a FASTA covering this dataset's proteins is included with the package so that building the vignette does not depend on UniProt being reachable.

```{r, eval=FALSE}
proteome_fasta <- tempfile(fileext = '.fasta')
make_fasta(unique(rowData(phospho[['psm_localised']])$Master.Protein.Accessions),
           file = proteome_fasta)
```

```{r, message=FALSE, warning=FALSE}
proteome_fasta <- system.file(
  "extdata", "tmt_phospho_proteome.fasta.gz", package = "biomasslmb")

phospho[['psm_localised']] <- add_ptm_positions(
  phospho[['psm_localised']],
  proteome_fasta = proteome_fasta,
  master_protein_col = 'Master.Protein.Accessions')
```

This adds `start` and `end` for the peptide, `ptm_positions_prot` for the sites within the protein, and `ptm_name`, which combines residue and position into the conventional form.

```{r}
rowData(phospho[['psm_localised']]) %>%
  data.frame() %>%
  select(Sequence, Master.Protein.Accessions, ptm_positions,
         start, ptm_positions_prot, ptm_name) %>%
  head(4)
```

Two things stop a peptide from being placed. Its master protein may have no sequence in the FASTA, which happens when an accession has been withdrawn or merged since the search; and the peptide may occur at more than one position in its protein, in which case `start` holds several values and the site cannot be assigned to one of them. Neither is common, and in a subset this size either count can come out at zero, but both have to be checked for and removed: a site with no unambiguous position cannot be summarised or compared.

```{r}
site_positions <- rowData(phospho[['psm_localised']])

data.frame(
  reason = c('No sequence for the master protein',
             'Peptide occurs more than once in its protein',
             'Placed'),
  PSMs = c(sum(is.na(site_positions$start)),
           sum(grepl(';', site_positions$start)),
           sum(!is.na(site_positions$start) & !grepl(';', site_positions$start))))
```

```{r}
phospho[['psm_sites']] <- phospho[['psm_localised']][
  !is.na(site_positions$start) & !grepl(';', site_positions$start), ]
```

# Peptides that cannot be localised

The localisation filter is the only step in this workflow that discards data on the strength of a probability rather than a measurement, and it is the most expensive one. Applied to the quality-filtered PSMs of this dataset at a threshold of 75, it removes a little under a quarter of them.

```{r}
psm_all_ptms <- phospho[['psm_quality']]
candidates <- parse_ptm_candidates_pd(psm_all_ptms)

resolution <- candidates %>%
  group_by(row, n_ptms) %>%
  summarise(n_localised = sum(prob >= 0.75), .groups = 'drop') %>%
  mutate(resolved = n_localised == n_ptms)

c(PSMs = nrow(resolution),
  discarded = sum(!resolution$resolved),
  percent = round(100 * mean(!resolution$resolved), 1))
```

**A peptide whose phosphate is tied between two serines is not an uninformative peptide.** It establishes that the region is phosphorylated and by how much, and leaves open only which of the two residues carries the phosphate. The filter throws away the quantification along with the ambiguity, and what it throws away is not a random subset: serine-rich and threonine-rich regions are exactly the ones a spectrum struggles to resolve, so the sites that go missing are concentrated in the sequences most worth looking at.

The alternative is to keep those peptides by pooling them. Peptides whose sites all reach the threshold are untouched and keep their own single-residue identity. Peptides that do not resolve contribute every candidate residue that is still plausible, and candidate sets that overlap are merged into one group, which is quantified as a unit. The cost is stated plainly at the end of this section: a pooled group names a region rather than a residue, and several things you would want to do with a site cannot be done with one.

## Reading every candidate residue

`parse_ptm_candidates_pd()` reads `ptmRS.Phospho.Site.Probabilities`, **not** the `ptmRS.Best.Site.Probabilities` column that `parse_PTM_scores_pd()` uses. The distinction is easy to miss and fails silently. Best Site reports only the winning isoform's sites, so the alternatives a group would be built from are absent from it, and pointing the parser at that column yields groups containing only the sites ptmRS already preferred.

```{r}
example_row <- resolution$row[!resolution$resolved][1]

rowData(psm_all_ptms)[example_row, c('ptmRS.Best.Site.Probabilities',
                                     'ptmRS.Phospho.Site.Probabilities')] %>%
  data.frame() %>%
  t()
```

The parser returns one row per candidate residue, with the position within the peptide, the residue, the probability and the number of modifications on the peptide. `parse_ptm_candidates_mq()` returns the same table from MaxQuant's `Phospho..STY..Probabilities` column, and everything after this point is shared between the two.

```{r}
head(candidates, 4)
```

Probabilities are on a 0 to 1 scale in both parsers, so a threshold means the same thing whichever search engine produced the data. ptmRS reports percentages and `parse_PTM_scores_pd()` takes its `threshold` as one — 75 above, 0.75 below — which is the one place in this workflow where the same quantity is expressed two ways.

## The probabilities are a budget, not a score

Each modification has to sit on some residue, so the probabilities reported for a peptide sum to the number of modifications on it. That holds in this dataset to within the rounding ptmRS applies.

```{r}
candidates %>%
  group_by(row, n_ptms) %>%
  summarise(total = sum(prob), .groups = 'drop') %>%
  summarise(min = round(min(total / n_ptms), 3),
            max = round(max(total / n_ptms), 3))
```

**This is what makes the pooling quantitative rather than cosmetic.** A candidate's probability is its share of one modification, so the summed probability of a set of candidates is the chance that the set contains the true site, and the probability left outside the set is the chance that it does not. A group is therefore not a vague claim about a region; it is a statement with a coverage figure attached.

## Grouping the candidates

Grouping happens in protein coordinates, so the peptides need placing first, as in the previous section — this time before the localisation filter has been applied rather than after.

```{r, message=FALSE, warning=FALSE}
psm_all_ptms <- add_peptide_positions_from_cleavage(
  psm_all_ptms, proteome_fasta, master_protein_col = 'Master.Protein.Accessions')

psm_all_ptms <- add_ambiguous_ptm_group_rowdata(
  psm_all_ptms, candidates,
  master_protein_col = 'Master.Protein.Accessions',
  min_prob = 0.75)
```

Candidate residues are nodes, and two residues are joined when they are candidates on the same peptide. The groups are the connected components of that graph, so a "S32 or S33" peptide and a "S33 or S37" peptide share a node and become one group over all three residues. The graph is built separately for each protein and each modification count, which stops a singly phosphorylated peptide from merging with a doubly phosphorylated one covering the same residues: they overlap in sequence but are different molecular species.

```{r}
grouped <- data.frame(rowData(psm_all_ptms))

c(resolved_PSMs = sum(grouped$ptm_group_resolved, na.rm = TRUE),
  pooled_PSMs = sum(!grouped$ptm_group_resolved %in% TRUE &
                      !is.na(grouped$ptm_group_id)),
  unassigned = sum(is.na(grouped$ptm_group_id)))
```

The PSMs the filter would have kept are unchanged — each still reports its own residue — and the PSMs it would have discarded are now distributed over a much smaller number of quantifiable groups.

```{r}
grouped %>%
  filter(!is.na(ptm_group_id)) %>%
  group_by(pooled = !ptm_group_resolved) %>%
  summarise(PSMs = n(), features = n_distinct(ptm_group_id))
```

A pooled group's size is set by how much genuine ambiguity there is, not by how many PSMs fell into it.

```{r}
grouped %>%
  filter(!ptm_group_resolved %in% TRUE, !is.na(ptm_group_id)) %>%
  distinct(ptm_group_id, ptm_group_n_candidates) %>%
  pull(ptm_group_n_candidates) %>%
  table()
```

## Choosing min_candidate_prob

`min_candidate_prob` is the probability below which a residue on an unresolved peptide is treated as ruled out rather than kept as a candidate. Because the probabilities are a budget, it is a coverage guarantee: the mass it discards is the chance the resulting group excludes the real site. `summarise_ptm_groups()` reports both sides of that trade at a given value.

```{r, message=FALSE, warning=FALSE}
lapply(c(0.01, 0.02, 0.05, 0.1, 0.2), function(threshold) {
  summarise_ptm_groups(psm_all_ptms, candidates, threshold, min_prob = 0.75,
                       master_protein_col = 'Master.Protein.Accessions')
}) %>%
  bind_rows() %>%
  select(min_candidate_prob, pooled_groups, median_size, pc_size_5_plus,
         p95_span, pc_miss_over_5)
```

**Raising the threshold buys very little resolution and pays for it steeply in coverage.** The number of groups barely moves, because the same peptides are pooled either way and only the size of the group they land in changes; the median group holds three candidates at every value in the table. What does move is the miss rate, which is the argument for keeping the threshold low.

The default of `min_candidate_prob = 0.02` is not taken from this subset, which is too small to settle it — 410 proteins, and the sweep above rests on `r sum(!resolution$resolved)` pooled PSMs. It comes from applying the same criterion to two complete datasets: the largest threshold at which under 1% of pooled peptides have more than a 5% chance of excluding the true site. That gives 0.03 on a whole-proteome MaxQuant phosphoproteome and 0.02 on a complete Proteome Discoverer one, and the stricter of the two is the default. At 0.02, 0.10% and 0.62% of pooled peptides respectively exceed a 5% miss rate.

The two search engines put the threshold to different work, which is worth knowing before adjusting it. MaxQuant never writes a probability below 0.001, so its low values are a reporting floor and 0.02 sits an order of magnitude above it. ptmRS writes exact zeros, and does so freely — they were 31% of all candidates in the dataset the default was measured on — so on Proteome Discoverer data any threshold above zero removes a large block of candidates before the value of the threshold matters at all.

## The span guard

`max_group_span` caps how far apart the first and last candidate in a group may be. It is a distance in residues, not a count of sites, and it exists to stop a chain of overlapping missed-cleavage peptides fusing genuinely unrelated regions of a protein into one group.

```{r}
grouped %>%
  filter(!ptm_group_resolved %in% TRUE, !is.na(ptm_group_id)) %>%
  distinct(ptm_group_id, ptm_group_members) %>%
  pull(ptm_group_members) %>%
  strsplit(';') %>%
  sapply(function(pos) diff(range(as.numeric(pos)))) %>%
  max()
```

The default of 50 is insurance rather than a tuning parameter. The widest group that forms unaided covers 33 residues here, 39 in the complete MaxQuant dataset and 43 in the complete Proteome Discoverer one, so on all three it never acts. Lowering it to 30 or below starts to cut groups, and a peptide whose candidates end up on both sides of a cut has no single group and is left unassigned rather than being forced into one.

## What a pooled group cannot do

A pooled group is a feature you can quantify and test, and `ptm_group_id` is usable directly as an `fcol` for `aggregateFeatures()` in place of the `site` identifier built earlier. Four things do not carry over.

A pooled group has no single residue, so it has no surrounding sequence. `add_site_sequence()` and the `site_seq` column it produces are undefined for one, and kinase and motif enrichment cannot consume pooled groups as they stand.

A pooled group must never be labelled as though it were a residue. `ptm_group_id` keeps the candidates visible, and `ptm_group_n_candidates` is the column to carry through to any label drawn elsewhere — a group reported as `S32` when it is one of four candidates is a claim the data does not support.

```{r}
group_labels <- grouped %>%
  filter(!is.na(ptm_group_id)) %>%
  distinct(ptm_group_id, ptm_group_members, ptm_group_n_candidates)

rbind(head(filter(group_labels, ptm_group_n_candidates == 1), 1),
      head(filter(group_labels, ptm_group_n_candidates > 1), 1))
```

`ptm_group_n_candidates` counts possible assignments rather than residues, which is why the localised peptide above reports two positions and a count of 1: it carries two phosphates, both placed, and there is one way to assign them. A count above 1 marks a pooled group, and its members are alternatives to one another rather than sites that co-occur. The distinction is the whole content of the column, and it is what any downstream label has to preserve.

Pooling sums intensity across residues, so a group containing two sites that move in opposite directions reports their average, which may be no change at all. The filter's answer to this case is to report nothing; pooling's is to report something attenuated, and neither is the site-level answer.

And the residues in a pooled group may also be quantified as resolved sites in their own right, from peptides that did localise. The same modification event then contributes to two features, which matters for multiple testing and for any enrichment analysis that treats features as independent.

```{r}
resolved_sites <- grouped %>%
  filter(ptm_group_resolved %in% TRUE) %>%
  pull(ptm_group_members) %>%
  unique()

pooled_members <- grouped %>%
  filter(!ptm_group_resolved %in% TRUE, !is.na(ptm_group_id)) %>%
  distinct(ptm_group_id, ptm_group_members)

c(pooled_groups = nrow(pooled_members),
  overlapping_a_resolved_site = sum(sapply(
    strsplit(pooled_members$ptm_group_members, ';'),
    function(pos) any(pos %in% resolved_sites))))
```

**That overlap is the rule rather than the exception.** Nearly every pooled group in this dataset contains at least one residue that is also measured as a localised site, which is unsurprising — a region gets pooled because some spectra could not place the modification there, not because none ever could. The consequence is that pooled and localised features are not two disjoint sets of measurements, and the number of independent tests is smaller than the number of features. How much smaller is not quantified: the overlap is easy to count, as above, but its effect on FDR depends on the test applied downstream, and this vignette does not attempt a correction for it.

Whether to group at all follows from the question. An analysis that rests on individual residues — a motif, a kinase substrate, a site to mutate — needs the localised sites and nothing else. An analysis asking which proteins and which regions respond to a treatment is the one that pays most for the filter, and it is the one grouping is for.

Finally, the identifier the PSMs will be summarised over. A site is only meaningful with its protein attached — `S614` names a residue in some protein, not a measurable thing — so the accession and the site name are pasted together.

```{r}
rowData(phospho[['psm_sites']])$site <- paste(
  rowData(phospho[['psm_sites']])$Master.Protein.Accessions,
  gsub(' ', '', rowData(phospho[['psm_sites']])$ptm_name),
  sep = '_')

head(unique(rowData(phospho[['psm_sites']])$site))
```

# Summarising to sites and proteins

With an identifier per fraction, both are summarised the same way as any TMT experiment: drop the PSMs with missing values, sum the rest, and log-transform. The enriched fraction is summarised over the site identifier, the total fraction over the master protein.

```{r, message=FALSE, warning=FALSE}
phospho[['psm_complete']] <- filterNA(phospho[['psm_sites']], 0)
total[['psm_complete']] <- filterNA(total[['psm_quality']], 0)

phospho <- aggregateFeatures(phospho, i = 'psm_complete', fcol = 'site',
                             name = 'site_raw', fun = base::colSums)
total <- aggregateFeatures(total, i = 'psm_complete',
                           fcol = 'Master.Protein.Accessions',
                           name = 'protein_raw', fun = base::colSums)

phospho[['site_raw']] <- logTransform(phospho[['site_raw']], base = 2)
total[['protein_raw']] <- logTransform(total[['protein_raw']], base = 2)

c(sites = nrow(phospho[['site_raw']]), proteins = nrow(total[['protein_raw']]))
```

A peptide carrying two phosphates becomes one feature named for both, not two features of one site each, because that is what was measured: the quantification belongs to the doubly-modified form of that peptide, and there is no way to divide it between the two residues. Peptides carrying one of the sites alone, if they were observed, form their own separate feature.

```{r}
table(sites_per_feature = lengths(
  strsplit(rowData(phospho[['site_raw']])$ptm_name, '; ')))
```

This means the same residue can appear in more than one feature, and that a feature's abundance is not the abundance of any single site. Treating each feature as an independent observation of its residues would double-count them.

# Normalising the enriched fraction

Unequal loading between channels has to be corrected in a PTM experiment just as in any other. The usual correction — centring every sample on a common median, as `QFeatures::normalize(method = 'diff.median')` does — assumes that most features are unchanged between samples. In the total fraction that is reasonable. In the enriched fraction it is not: a treatment that changes phosphorylation globally will shift the median of the enriched data, and centring it away removes exactly the effect the experiment was run to detect.

The total fraction gives the reference instead. It comes from the same labelled pool, so it carries the same loading differences, and its median is a legitimate estimate of them. `get_medians()` extracts the per-sample medians, and `center_normalise_to_ref()` applies them to whichever assay you pass.

```{r}
total_medians <- get_medians(total[['protein_raw']])

total[['protein']] <- center_normalise_to_ref(
  total[['protein_raw']], total_medians, on_log_scale = TRUE)

phospho[['site']] <- center_normalise_to_ref(
  phospho[['site_raw']], total_medians, on_log_scale = TRUE)
```

`center_normalise_to_ref()` with the default `center_to_zero = FALSE` subtracts each median only after the medians have themselves been centred on their mean, so it applies the *relative* differences between samples and leaves the overall level of the data alone. That is why the same reference can be applied to two assays on quite different scales.

```{r, warning=FALSE, fig.show='hold', out.width="50%", fig.height=4, fig.width=5, fig.cap="Site-level abundance distributions, before and after normalisation"}
plot_quant(phospho[['site_raw']], log2transform = FALSE, method = 'density') +
  theme_biomasslmb(base_size = 9) +
  theme(legend.key.height = unit(3, 'mm')) +
  labs(x = 'Site abundance (log2)', title = 'Before normalisation')

plot_quant(phospho[['site']], log2transform = FALSE, method = 'density') +
  theme_biomasslmb(base_size = 9) +
  theme(legend.key.height = unit(3, 'mm')) +
  labs(x = 'Site abundance (log2)', title = 'After normalisation')
```

# Site abundance and occupancy

A site's abundance can change for two reasons: the fraction of the protein that is modified has changed, or the amount of the protein has changed. Only the first is a change in signalling; the second would show up just as well in the total fraction.

Subtracting the protein's log abundance from the site's separates them. The result is often called occupancy or stoichiometry, though neither name is quite right — an enrichment step does not preserve absolute proportions, so this is a relative measure, comparable between samples but not interpretable as "x% of the protein is modified".

```{r}
site_protein <- rowData(phospho[['site']])$Master.Protein.Accessions
has_protein <- site_protein %in% rownames(total[['protein']])

occupancy <- assay(phospho[['site']])[has_protein, ] -
  assay(total[['protein']])[site_protein[has_protein], ]

c(sites = nrow(phospho[['site']]), with_a_protein_estimate = sum(has_protein))
```

Not every site gets one. A protein can be quantified in the enriched fraction and not the total, which happens most for low-abundance proteins, where the enrichment is the only reason they were seen at all — so the sites that lose their protein estimate are not a random subset.

# Testing

Both matrices can go into `limma` in the usual way. The design here has a treatment and a timepoint, and the question is about the treatment, so the timepoint enters as a blocking factor.

```{r}
condition <- factor(phospho[['site']]$Condition,
                    levels = c('Control', 'Treated'))
timepoint <- factor(phospho[['site']]$Timepoint)
design <- model.matrix(~ timepoint + condition)

test_treatment <- function(quant) {
  fit <- eBayes(lmFit(quant, design))
  topTable(fit, coef = 'conditionTreated', number = Inf, sort.by = 'none')
}

site_res <- test_treatment(assay(phospho[['site']]))
protein_res <- test_treatment(assay(total[['protein']]))
occupancy_res <- test_treatment(occupancy)

data.frame(
  level = c('Site abundance', 'Occupancy'),
  tested = c(nrow(site_res), nrow(occupancy_res)),
  significant = c(sum(site_res$adj.P.Val < 0.05),
                  sum(occupancy_res$adj.P.Val < 0.05)))
```

Those are the two answers you might report. The protein-level test is not a third one — it is a different question over a different set of features — but it is what tells you how far to trust the agreement between these two. Only `r sum(protein_res$adj.P.Val < 0.05)` of the `r nrow(protein_res)` proteins quantified in the total fraction changes detectably with treatment, so for most sites there is nothing for the correction to remove, and the two site-level answers mostly agree.

```{r}
comparison <- data.frame(
  site = rownames(occupancy),
  protein = site_protein[has_protein],
  site_logFC = site_res$logFC[has_protein],
  site_padj = site_res$adj.P.Val[has_protein],
  occupancy_logFC = occupancy_res$logFC,
  occupancy_padj = occupancy_res$adj.P.Val) %>%
  mutate(
    protein_logFC = protein_res[protein, 'logFC'],
    protein_padj = protein_res[protein, 'adj.P.Val'],
    called = case_when(
      site_padj < 0.05 & occupancy_padj < 0.05 ~ 'both',
      site_padj < 0.05 ~ 'site only',
      occupancy_padj < 0.05 ~ 'occupancy only',
      TRUE ~ 'neither'))

table(comparison$called)
```

The interesting cases are the disagreements.

```{r}
comparison %>%
  filter(called != 'neither', called != 'both') %>%
  select(site, called, site_logFC, site_padj,
         occupancy_logFC, occupancy_padj, protein_logFC, protein_padj) %>%
  mutate(across(where(is.numeric), ~ signif(.x, 2)))
```

```{r, fig.height=4, fig.width=5.5, fig.fullwidth=TRUE, fig.cap="Site-level against occupancy-level fold change. Points off the diagonal are sites whose protein also moved."}
ggplot(comparison, aes(site_logFC, occupancy_logFC)) +
  geom_hline(yintercept = 0, colour = 'grey80') +
  geom_vline(xintercept = 0, colour = 'grey80') +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = 'grey50') +
  geom_point(aes(colour = called), size = 1.5, alpha = 0.7) +
  scale_colour_manual(values = c(get_cat_palette(3), 'grey70'),
                      breaks = c('both', 'occupancy only', 'site only', 'neither'),
                      name = 'Significant at') +
  theme_biomasslmb(base_size = 9) +
  labs(x = 'Site abundance (log2 fold change)',
       y = 'Occupancy (log2 fold change)')
```

A site called at one level and not the other is not a contradiction; the two are answering different questions. A site that moves only in the abundance test, while its protein moves by a similar amount, is a protein-level change showing through — the modified fraction is constant and more of the protein is present. A site that moves only in the occupancy test is one where protein-level variation was adding noise, and removing it made a modest change detectable. With so few protein-level changes in this subset, the first of those cases appears only once here; where a treatment moves protein abundances more widely, it is correspondingly commoner.

Which of the two to report depends on the question. Occupancy is the right answer for "did signalling through this site change", and it is what most phosphoproteomics is asking. Site abundance is the right answer for "how much of this modified form is present", which is the relevant quantity when the modified form is itself the effector. Reporting both, as above, costs nothing and makes the distinction visible; reporting site abundance alone and describing it as a change in phosphorylation is the mistake this section exists to prevent.

Note also that occupancy is only available for sites whose protein was quantified in the total fraction, so restricting to it discards sites — and, as noted above, not a random subset of them.

# Testing both levels in one model

Subtracting the protein from the site and testing the result treats the protein
estimate as if it were exact. It is not — it is a measurement with its own
uncertainty, made on the same sample — and the subtraction throws that
uncertainty away before `limma` ever sees it.

The alternative is to give `limma` both measurements and let it do the
subtraction as part of the model. Stack the protein and site abundances into
one matrix, one row per site, with a `type` factor saying which kind of
measurement each column holds.

```{r}
site_mat <- assay(phospho[['site']])[has_protein, ]
protein_mat <- assay(total[['protein']])[site_protein[has_protein], ]

joint <- cbind(protein_mat, site_mat)
colnames(joint) <- paste(rep(c('protein', 'site'), each = ncol(site_mat)),
                         colnames(site_mat), sep = '.')

joint_sample <- rep(colnames(site_mat), 2)
joint_type <- factor(rep(c('protein', 'site'), each = ncol(site_mat)),
                     levels = c('protein', 'site'))
joint_condition <- factor(rep(phospho[['site']]$Condition, 2),
                          levels = c('Control', 'Treated'))
joint_timepoint <- factor(rep(phospho[['site']]$Timepoint, 2))

dim(joint)
```

Every term that can act differently on the two measurement types is then
interacted with `type`. The treatment effect on the protein is
`conditionTreated`; the extra effect on the site, over and above the protein's,
is `conditionTreated:typesite`. That interaction is the quantity occupancy was
constructed to isolate.

```{r}
joint_design <- model.matrix(~ joint_timepoint + joint_condition * joint_type +
                               joint_timepoint:joint_type)

colnames(joint_design)
```

The same sample supplies one protein column and one site column, so those two
are not independent observations. `duplicateCorrelation` estimates a single
consensus correlation within blocks, and `lmFit` uses it to weight the fit.
[Technical replicates](gotcha_technical_replicates.html) covers the same
machinery applied to repeated acquisitions of one sample.

```{r}
corfit <- duplicateCorrelation(joint, joint_design, block = joint_sample)

round(corfit$consensus.correlation, 3)
```

```{r}
joint_fit <- eBayes(lmFit(joint, joint_design, block = joint_sample,
                          correlation = corfit$consensus.correlation))

joint_res <- topTable(joint_fit, coef = 'joint_conditionTreated:joint_typesite',
                      number = Inf, sort.by = 'none')

table(joint = joint_res$adj.P.Val < 0.05,
      occupancy = occupancy_res$adj.P.Val < 0.05)
```

**The two routes estimate the same fold changes and disagree only on the
uncertainty.** The interaction coefficient and the occupancy fold change are
the same number to machine precision, which they have to be — both are the
difference between the site's response and the protein's.

```{r}
max(abs(joint_res$logFC - occupancy_res$logFC))
```

What differs is the standard error each is divided by, and that is enough to
move a few sites across the threshold in both directions. The gain is modest
here because the pairing is weak in this dataset — a consensus correlation of
`r round(corfit$consensus.correlation, 3)` — but it is the block term that
delivers it, and dropping it gives back most of the difference.

```{r}
unblocked <- eBayes(lmFit(joint, joint_design))

c(blocked = sum(joint_res$adj.P.Val < 0.05),
  unblocked = sum(topTable(unblocked, coef = 'joint_conditionTreated:joint_typesite',
                           number = Inf)$adj.P.Val < 0.05),
  occupancy = sum(occupancy_res$adj.P.Val < 0.05))
```

The reason to reach for this rather than the subtraction is not the handful of
sites it moves. It is that the model can express designs the subtraction
cannot. A term interacted with `type` is a term allowed to act differently on
protein and site, so a covariate that affects protein abundance and site
occupancy in different ways can be adjusted for as such, and a design with more
than two groups yields one interaction coefficient per group from a single fit
rather than a subtraction repeated per contrast.

The cost is the assumption `duplicateCorrelation` makes: one correlation
describes every block. Where some samples pair their two fractions much more
tightly than others — different enrichment batches, say — that single number
describes none of them well.

# Summary

A PTM experiment adds three steps to the whole-proteome workflow, and each of them is a place to get the answer wrong:

- **Localisation.** The search engine's site assignment is a probabilistic claim, scored separately from the peptide identification. `parse_PTM_scores_pd()` reads the scores and discards peptides whose sites cannot be placed confidently. Positions then have to be converted from peptide coordinates to protein coordinates with `add_ptm_positions()`, dropping the peptides that occur at more than one position in their protein. Discarding is not the only option: `parse_ptm_candidates_pd()` and `add_ambiguous_ptm_group_rowdata()` keep the unlocalised peptides by pooling their candidate residues into a group with a stated chance of containing the true site, which buys back a quarter of the PSMs here at the price of a feature that names a region rather than a residue.
- **Normalisation.** The enriched fraction's own median is not a valid reference, because a global change in modification is signal rather than a loading artefact. `get_medians()` on the total fraction and `center_normalise_to_ref()` on the enriched one keeps the loading correction and leaves the biology.
- **Occupancy.** A change in site abundance is a change in modification only once the protein's own change has been divided out. Dividing it out by subtraction and testing the result is the direct route; fitting protein and site abundances jointly, with a `type` interaction and `duplicateCorrelation` on the sample, reaches the same fold changes with an uncertainty that accounts for the pairing, and extends to designs the subtraction cannot express.

Two of the steps in this vignette are shared with any TMT experiment and are covered in more detail elsewhere: PSM filtering and summarisation in [TMT QC and summarisation](TMT_PSM_QC_Summarisation.html), and testing in [Data exploration and statistical testing](exploration_and_statistical_testing.html). If the design spans more than one plex, both fractions need bringing onto a common scale before any of the above, which [multi-plex TMT](TMT_multiplex.html) covers — the bridge correction is applied to the enriched and total assays separately, using each one's own bridge channels.

# Getting help

If your experiment does not fit what this article assumes, or a result looks
wrong, please get in touch with Tom Smith (tsmith@mrclmb.ac.uk) rather than
guessing — it is far easier to help while an analysis is in progress than to
unpick a decision afterwards, and easier still before the samples are run.
[Getting started](biomasslmb.html) sets out which article applies to which
experiment.

```{r}
sessionInfo()
```
