---
title: "Choosing a summarisation method"
author: "Tom Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: biomasslmb.json
vignette: >
  %\VignetteIndexEntry{Choosing a summarisation method}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Summarising lower-level quantification (PSMs for TMT, peptides for LFQ) to protein-level abundance can be done with `sum`, `median` or `MsCoreUtils::robustSummary`, and the choice changes both which proteins are quantified and what their abundance profiles look like. (`mean` is not considered separately: it's just `sum` divided by the number of contributing features, so it produces the same abundance profile across samples as `sum`, only rescaled.) These are general considerations for bottom-up proteomics, independent of the quantification technology: the same weighting and missing-value behaviours apply whether the lower-level features are PSMs or peptides. The methods differ in two distinct ways, and the two are easy to run together, so they are separated here across three datasets — two from the [TMT QC](TMT_PSM_QC_Summarisation.html) vignette and one from the [LFQ-DDA QC](LFQ_DDA_Peptide_QC_Summarisation.html) vignette:

- **How the methods weight individual features.** Even with complete data, `sum` and `robustSummary`/`median` produce different protein-level estimates, because they do not treat the underlying features equally. Part 1 isolates this using `tmt_qf`, the whole-proteome TMT PD dataset, which has very little missingness for it to be confounded with.
- **How the methods handle missing values.** `sum` requires complete data, whereas `robustSummary` can use features with some missing values. Part 2 takes this up with `tmt_qf_mq`, the TMT MaxQuant IP dataset, where missingness is common and potentially informative; Part 3 repeats it with `lfq_qf`, an LFQ-DDA whole-proteome dataset, to establish that the conclusion is about missingness rather than about TMT.

# Load required packages

```{r, message=FALSE}
library(QFeatures)
library(biomasslmb)
library(ggplot2)
library(tidyr)
library(dplyr)

tmt_qf <- biomasslmb::tmt_qf
tmt_qf_mq <- biomasslmb::tmt_qf_mq
lfq_qf <- biomasslmb::lfq_qf
```

# Part 1: How the methods weight individual PSMs

`tmt_qf` already contains protein-level quantification from a simple `sum` summarisation (`protein`), with missing values and single-PSM proteins removed. `robustSummary` and `median` are computed below from the same starting point, so that any difference between the three is attributable to the estimator.

## Summarisation with robustSummary

`robustSummary` handles missing values within the algorithm, so PSMs carrying some do not have to be discarded. PSMs missing from most channels still contribute little beyond noise, so a threshold is applied anyway — here, PSMs missing from at most half the channels.

```{r}
tmt_qf[['psms_filtered_forRobust']] <- QFeatures::filterNA(
  tmt_qf[['psms_filtered_rank']], 5/10)

```

PSMs belonging to proteins with fewer than 2 PSMs are then removed.
```{r}

min_psms <- 2

tmt_qf[['psms_filtered_forRobust']] <- biomasslmb::filter_features_per_protein(
  tmt_qf[['psms_filtered_forRobust']], min_features = min_psms)

```

`robustSummary` assumes approximately Gaussian quantification values, so they are log transformed.

```{r}
tmt_qf[['psms_filtered_forRobust']] <- QFeatures::logTransform(
  tmt_qf[['psms_filtered_forRobust']], base=2)
```

Summarising with `robustSummary`:

```{r, warning=FALSE}
# Aggregate to protein-level abundances (using QFeatures function)
tmt_qf <- QFeatures::aggregateFeatures(tmt_qf,
                                       i = "psms_filtered_forRobust",
                                       fcol = "Master.Protein.Accessions",
                                       name = "protein_robust",
                                       fun = MsCoreUtils::robustSummary,
                                       maxit=1000)


```

## Summarisation with median

`median` is included because it is a typical way to summarise across observations, and because it makes the behaviour of the other two easier to read — not because it is recommended, for reasons the rest of Part 1 sets out. It is given the PSMs with missing values already filtered out. `median` would accept them with `na.rm=TRUE`, but silently taking the median of whichever PSMs happen to be present computes a different quantity in each sample.

```{r}

tmt_qf <- QFeatures::aggregateFeatures(tmt_qf,
                                       i = "psms_filtered_forSum",
                                       fcol = "Master.Protein.Accessions",
                                       name = "protein_median",
                                       fun = matrixStats::colMedians, na.rm=TRUE)

tmt_qf[['protein_median']] <- QFeatures::logTransform(
  tmt_qf[['protein_median']], base=2)
```

## Comparing the summarisation approaches

As established in the QC vignette, this dataset has very few missing values to begin with, so tolerating missing PSMs gains `robustSummary` no proteins here. That is what makes it the right dataset for the first question: with the missing-value advantage held at zero, any disagreement between the methods is about how they weight features. Part 2 takes the missing-value question to a dataset that has some.

```{r}

# Single object with protein inference from both methods
compare_protein_abundances <- qfeatures_long(
  tmt_qf[,,c('protein', 'protein_robust', 'protein_median')]) %>%
  data.frame() %>%
  mutate(method=recode_values(
    assay,
    'protein'~'Sum',
    'protein_robust'~'Robust',
    'protein_median'~'Median'))

```

The proteins where the methods disagree most are the informative ones to look at, and a disagreement between three methods has three shapes: any one of them can be the odd one out while the other two agree. One protein is taken for each shape — the protein where that method departs furthest from the other two, scored as the correlation between the other two minus the better of the departing method's own two correlations.

The comparison is restricted to proteins where all three methods summarise an identical set of PSMs. `robustSummary` accepts PSMs carrying missing values that `sum` and `median` never see, and where it does, a departure would be a difference in the features available to it rather than in the estimator — that is Part 2's subject, and it would confound this one.

```{r}
psms_per_protein <- function(i) {
  split(rownames(tmt_qf[[i]]),
        rowData(tmt_qf[[i]])$Master.Protein.Accessions)
}

sum_psms <- psms_per_protein('psms_filtered_forSum')
robust_psms <- psms_per_protein('psms_filtered_forRobust')

same_psms <- Filter(function(protein) setequal(sum_psms[[protein]], robust_psms[[protein]]),
                    intersect(names(sum_psms), names(robust_psms)))

method_agreement <- compare_protein_abundances %>%
  filter(rowname %in% same_psms) %>%
  select(method, value, rowname, colname) %>%
  pivot_wider(names_from = method, values_from = value) %>%
  group_by(rowname) %>%
  filter(!any(is.na(c(Sum, Robust, Median)))) %>%
  summarise(sum_robust = cor(Sum, Robust),
            sum_median = cor(Sum, Median),
            robust_median = cor(Robust, Median)) %>%
  mutate(Sum = robust_median - pmax(sum_robust, sum_median),
         Robust = sum_median - pmax(sum_robust, robust_median),
         Median = sum_robust - pmax(sum_median, robust_median))

proteins_of_interest <- sapply(
  c(Sum = 'Sum', Robust = 'Robust', Median = 'Median'),
  function(method) method_agreement$rowname[which.max(method_agreement[[method]])])

proteins_of_interest
```

A helper plots one protein twice: its PSM-level abundances on their original scale, and the protein-level abundance from each method normalised to mean abundance, so the three methods sit on a comparable scale.

```{r}
plot_pep_and_protein <- function(protein_of_interest) {

  to_plot_compare <- compare_protein_abundances %>%
    filter(rowname == protein_of_interest) %>%
    group_by(method) %>%
    mutate(value = value - mean(value)) %>%
    ungroup()

  pep_plot <- QFeatures::filterFeatures(
    tmt_qf,
    VariableFilter("Master.Protein.Accessions",
                   protein_of_interest,
                   condition = "=="))[['psms_filtered_sn']] %>%
    qfeatures_long() %>%
    ggplot(aes(x = colname, y = log2(value))) +
    geom_line(aes(group = rowname), colour = 'grey') +
    geom_point(colour = 'grey') +
    theme_biomasslmb(base_size = 15, border = FALSE, base_family = 'sans') +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    labs(
      title = protein_of_interest,
      x = '',
      y = 'PSM abundance (log2)'
    )

  protein_plot <- to_plot_compare %>%
    ggplot(aes(x = colname, y = value, colour = method, group = method)) +
    geom_line() +
    scale_colour_manual(values = get_cat_palette(3),
                        name = 'Protein summarisation method') +
    theme_biomasslmb(base_size = 15, border = FALSE, base_family = 'sans') +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    labs(
      title = protein_of_interest,
      x = '',
      y = 'Protein abundance (log2, mean-normalised)'
    ) +
    guides(color = guide_legend(override.aes = list(linewidth = 2) ) )

  list(peptide = pep_plot, protein = protein_plot)
}
```

In the peptide-level plots below, each grey line is a single PSM's abundance on its original scale. In the protein-level plots, each summarisation method's abundance profile is mean-normalised, so the methods sit on a comparable scale and can be compared directly. The important thing to focus on is the abundance profile across the tags for any one summarisation method or PSM.

```{r, plot.show='hold', out.width='45%', fig.width=7, fig.height=7, warning=FALSE, message=FALSE}
for(method in names(proteins_of_interest)) {
  plots <- plot_pep_and_protein(proteins_of_interest[[method]])
  print(plots$peptide + labs(subtitle = paste(method, 'departs')))
  print(plots$protein + labs(subtitle = paste(method, 'departs')))
}
```


Two of the three are shapes of genuine conflict: for `r proteins_of_interest[['Sum']]` and `r proteins_of_interest[['Median']]`, the departing method describes a different profile across most of the plex rather than a discrepancy at one or two tags.

**`robustSummary` is seldom the method that departs.** The worst case for `sum` scores `r round(max(method_agreement$Sum), 2)` on the measure above and the worst case for `median` `r round(max(method_agreement$Median), 2)`, while across the `r nrow(method_agreement)` proteins compared here the worst case for `robustSummary` reaches only `r round(max(method_agreement$Robust), 2)`. That worst case is milder than its correlations suggest: the `robustSummary` profile for `r proteins_of_interest[['Robust']]` varies across the tags with a standard deviation of `r round(sd(assay(tmt_qf[['protein_robust']])[proteins_of_interest[['Robust']], ]), 2)` against `r round(sd(assay(tmt_qf[['protein_median']])[proteins_of_interest[['Robust']], ]), 2)` for `median`, so the disagreement is between a nearly flat profile and a structured one rather than between two conflicting accounts of the protein.

Since `sum` simply adds together the abundance values across the PSMs for a given tag and the abundance values may span multiple orders of magnitude, the protein-level abundance pattern across the tags is weighted towards the most abundant PSMs. This is likely to be a positive attribute, since the most abundant PSMs will be more accurately quantified. However, the summarisation is also sensitive to high abundance outliers. `robustSummary` and `median`, by contrast, ignore absolute abundance: `robustSummary` weights each PSM by how far it sits from the consensus profile, while `median` keeps only whichever PSMs fall in the middle of the range.

This is visible directly for the protein where `median` departs. Correlating each PSM's abundance profile across the tags against the `sum`-summarised protein profile gives a low or negative value for any PSM that disagrees with the majority.

```{r}
poi <- proteins_of_interest[['Median']]

rd <- rowData(tmt_qf[['psms_filtered_sn']])
psm_mask <- rd$Master.Protein.Accessions == poi
psm_values <- log2(assay(tmt_qf[['psms_filtered_sn']])[psm_mask, , drop = FALSE])
sum_profile <- assay(tmt_qf[['protein']])[poi, ]

psm_cors <- apply(psm_values, 1, function(x) cor(x, sum_profile, use = "complete.obs"))

sort(psm_cors)
```

Colouring the same PSM traces by that correlation puts the disagreement and the abundance on one pair of axes.

```{r, fig.width=7, fig.height=5, warning=FALSE, message=FALSE}
psm_cor_long <- data.frame(psm_values, check.names = FALSE) %>%
  mutate(rowname = rownames(psm_values),
         correlation = psm_cors[rowname]) %>%
  pivot_longer(cols = all_of(colnames(psm_values)),
               names_to = 'colname', values_to = 'abundance')

psm_cor_long %>%
  ggplot(aes(x = colname, y = abundance, group = rowname, colour = correlation)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 1.5) +
  scale_colour_gradient2(low = get_cat_palette(2)[2], mid = 'grey80',
                         high = get_cat_palette(1), midpoint = 0,
                         limits = c(-1, 1),
                         name = 'Correlation with\nsum profile') +
  theme_biomasslmb(base_size = 15, border = FALSE, base_family = 'sans') +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(title = poi, x = '', y = 'PSM abundance (log2)')
```

**Disagreement concentrates in the faint PSMs, and the one that matters is the exception.** Agreement with the summed profile rises with abundance — Spearman `r round(cor(rowMeans(psm_values), psm_cors, method = 'spearman'), 2)` between the two here — which is what noise in the poorly quantified PSMs would look like, and it is the case for weighting by intensity, since `sum` discounts those automatically. The most strongly anticorrelated PSM is the one that breaks the pattern: it ranks `r rank(rowMeans(psm_values))[[names(which.min(psm_cors))]]` of `r nrow(psm_values)` by mean abundance and carries `r round(100 * mean(2^psm_values[names(which.min(psm_cors)), ] / colSums(2^psm_values)), 1)`% of the summed signal against the `r round(100 / nrow(psm_values), 1)`% an equal split would give it, so `sum` neither discounts it for being dim nor lets it dominate for being bright.

What a PSM in that position costs each method can be measured by dropping each PSM in turn and re-summarising from the rest.

```{r}
summarise_psms <- function(ids, method) {
  m <- 2^psm_values[ids, , drop = FALSE]
  profile <- switch(method,
                    Sum = log2(colSums(m)),
                    Robust = MsCoreUtils::robustSummary(log2(m), maxit = 1000),
                    Median = log2(matrixStats::colMedians(m)))
  profile - mean(profile)
}

loo_cor <- sapply(c(Sum = 'Sum', Robust = 'Robust', Median = 'Median'), function(method) {
  full <- summarise_psms(rownames(psm_values), method)
  sapply(rownames(psm_values), function(psm)
    cor(full, summarise_psms(setdiff(rownames(psm_values), psm), method)))
})

# the largest and second-largest effect any single PSM has on each method
round(data.frame(worst = apply(loo_cor, 2, min),
                 next_worst = apply(loo_cor, 2, function(x) sort(x)[2])), 3)
```

**One PSM out of `r nrow(psm_values)` decides the `median` profile, and none of them decides the other two.** No single PSM moves `sum` below a correlation of `r round(min(loo_cor[, 'Sum']), 3)` with its original profile, or `robustSummary` below `r round(min(loo_cor[, 'Robust']), 3)`. `median` survives the loss of any PSM equally well — down to `r round(sort(loo_cor[, 'Median'], decreasing = TRUE)[[nrow(psm_values) - 1]], 2)` — except one, whose removal takes it to `r round(min(loo_cor[, 'Median']), 2)`, inverting the profile. That PSM ranks `r rank(rowMeans(psm_values))[[names(which.min(loo_cor[, 'Median']))]]` of `r nrow(psm_values)` by abundance: `median` reads the middle of the range and nothing else, so whichever PSM sits there controls the result, and here that PSM disagrees with the rest. `sum` gives it only the weight its intensity earns and `robustSummary` discounts it further for sitting away from the consensus, which is why the two of them track each other at `r round(cor(sum_profile, assay(tmt_qf[['protein_robust']])[poi, ]), 2)` for this protein while `median` sits at `r round(cor(sum_profile, assay(tmt_qf[['protein_median']])[poi, ]), 2)` against `sum`.

The median is less sensitive to a bright outlier than the mean, but ignores much of the quantitative data since it's only affected by the 'middle' PSMs. In the most extreme instances, median just simplifies the PSM level abundances down to a single representative 'middle' PSM. That instability is a general property of the estimator rather than a quirk of this protein, and the next section measures it across the dataset.

## Why the median is the least stable summarisation

In the profiles above, the `median` line moves around more than the other two. This is a general property rather than a feature of these particular proteins, and it can be measured. Because `median` and `sum` are both computed here from the same PSMs (`psms_filtered_forSum`), any difference between them is down to the estimator alone.

The proteins in this dataset genuinely differ between `Control` and `Mutant`, so the spread of a profile across all twelve tags mixes real signal with noise. What is noise, and only noise, is how much an estimate moves between replicates of the *same* condition.

```{r}
psm_counts <- table(
  rowData(tmt_qf[['psms_filtered_forSum']])$Master.Protein.Accessions)

condition <- colData(tmt_qf)$Condition

replicate_sd <- function(x) {
  sqrt(mean(tapply(seq_along(condition), condition, function(i) var(x[i]))))
}

methods <- c(Sum = 'protein', Robust = 'protein_robust', Median = 'protein_median')

shared <- Reduce(intersect, lapply(methods, function(a) rownames(tmt_qf[[a]])))
shared <- shared[Reduce(`&`, lapply(methods, function(a)
  complete.cases(assay(tmt_qf[[a]])[shared, ])))]

stability <- data.frame(
  sapply(methods, function(a) apply(assay(tmt_qf[[a]])[shared, ], 1, replicate_sd)),
  n_psm = as.integer(psm_counts[shared]))
```

```{r}
stability %>%
  mutate(psms = cut(n_psm, c(1, 2, 3, 5, 9, Inf),
                    labels = c('2', '3', '4-5', '6-9', '10 or more'))) %>%
  group_by(psms) %>%
  summarise(proteins = n(),
            across(c(Sum, Robust, Median), ~ round(median(.x), 3)))
```

Two things stand out. With exactly 2 PSMs all three methods are equally stable, and `sum` and `median` are in fact identical: the median of two values is their mean, which is the sum divided by two, and dividing by a constant is an offset on the log scale that a profile cannot see. From 3 PSMs upwards `median` is the noisiest, typically by around `r round(100 * (median(with(stability[stability$n_psm >= 3, ], Median / Sum)) - 1))`%, and the gap widens as PSMs accumulate: `sum` and `robustSummary` become steadily more precise, while `median` improves very little.

That is the explanation in full. **`sum` and `robustSummary` combine every PSM; `median` reports one of them.** Averaging several measurements of the same quantity cancels part of the error in each, so the more PSMs a protein has, the better those two estimates get. Taking the middle value does not average anything, so the protein estimate carries a single PSM's worth of noise however many PSMs were available. Even when everything is well behaved this costs about a fifth more variability — the standard penalty for using a median in place of a mean.

The bigger cost is that rank position says nothing about how well a PSM was measured, so `median` has no protection against landing on a badly measured one, and when it does the protein inherits that PSM's noise in full. For `r sum(with(stability[stability$n_psm >= 3, ], Median / Sum) > 2)` of the `r sum(stability$n_psm >= 3)` proteins with 3 or more PSMs, `median` is more than twice as noisy as `sum`. `sum` avoids this without doing anything clever, because weighting PSMs by their absolute intensity is close to weighting them by their precision:

```{r}
psm_quant <- log2(assay(tmt_qf[['psms_filtered_forSum']]))

data.frame(intensity = rowMeans(psm_quant),
           noise = apply(psm_quant, 1, replicate_sd)) %>%
  mutate(quartile = cut(intensity, quantile(intensity), include.lowest = TRUE,
                        labels = paste0('Q', 1:4))) %>%
  group_by(quartile) %>%
  summarise(psms = n(), noise = round(median(noise), 3))
```

The most intense PSMs are the most reproducibly quantified, so `sum` leans on them automatically. `median` weights by rank instead, which is uninformative about quality.

A smaller contribution is that the PSM occupying the middle need not be the same one in every tag. Since the PSMs of a single protein sit at different absolute intensities, every switch moves the estimate to a different baseline:

```{r}
poi_odd <- proteins_of_interest[
  as.integer(psm_counts[proteins_of_interest]) %% 2 == 1][1]

poi_psms <- log2(assay(tmt_qf[['psms_filtered_forSum']])[
  rowData(tmt_qf[['psms_filtered_forSum']])$Master.Protein.Accessions == poi_odd, ])

middle_psm <- apply(poi_psms, 2, function(x) rownames(poi_psms)[order(x)[(length(x) + 1) / 2]])

round(sort(rowMeans(poi_psms)[unique(middle_psm)]), 2)
```

`r poi_odd` has `r nrow(poi_psms)` PSMs, and `r length(unique(middle_psm))` different ones take a turn in the middle across the twelve tags, at mean intensities spanning a `r round(2^diff(range(rowMeans(poi_psms)[unique(middle_psm)])), 1)`-fold range. Each handover puts a step into the profile that is not an abundance change, which is why these profiles look jagged rather than merely noisy.

`robustSummary` avoids all of this while keeping its resistance to outliers, because it does not summarise each sample separately. It fits one linear model across the whole PSM-by-sample matrix, with a term for every PSM and a term for every sample, by iteratively reweighted least squares, and returns the sample terms. Discordant observations are down-weighted rather than discarded, so every PSM informs every sample's estimate; and each PSM's own intensity is absorbed into its own term, so it cannot leak into the between-sample profile. Resistance to outliers comes from the weights rather than from ignoring all but the middle value, which is why it matches `sum` for stability in the table above while remaining, as the disagreeing PSM earlier showed, far less sensitive to any one PSM.

This is why `median` is included here for interpretation rather than recommended: it shows what happens when a summarisation discards quantitative information, and the cost begins as soon as a protein has more than two PSMs.

# Part 2: How the methods handle missing values

Part 2 switches to `tmt_qf_mq`, the MaxQuant IP dataset from the [enrichment designs](interactome_designs.html) vignette. Missing values here are common and potentially informative — a protein absent from the control pulldown is genuinely absent rather than undetected by chance — which is exactly the condition under which the two methods can be expected to diverge.

`tmt_qf_mq[['psms_filtered_interference']]` (the PSMs used for summarisation in that vignette) is stored on the log2 scale, since that is what `robustSummary` requires. A `sum` comparison needs the values back on their original scale.

```{r}
tmt_qf_mq[['psms_filtered_interference_raw']] <- tmt_qf_mq[['psms_filtered_interference']]
assay(tmt_qf_mq[['psms_filtered_interference_raw']]) <- 2^assay(tmt_qf_mq[['psms_filtered_interference_raw']])
```

A fair comparison needs the same rule applied to both approaches, so both keep only proteins with at least 2 retained PSMs. They differ in one respect alone: how much missingness a PSM may carry before it is discarded — none for `sum`, up to 8/12 for `robustSummary`, the same threshold used in the QC vignette.

```{r}
tmt_qf_mq[['psms_filtered_forSum_mq']] <- QFeatures::filterNA(
  tmt_qf_mq[['psms_filtered_interference_raw']], 0)
tmt_qf_mq[['psms_filtered_forSum_mq']] <- biomasslmb::filter_features_per_protein(
  tmt_qf_mq[['psms_filtered_forSum_mq']], min_features = min_psms,
  master_protein_col = 'Leading.razor.protein')

tmt_qf_mq[['psms_filtered_forRobust_mq']] <- QFeatures::filterNA(
  tmt_qf_mq[['psms_filtered_interference']], 8/12)
tmt_qf_mq[['psms_filtered_forRobust_mq']] <- biomasslmb::filter_features_per_protein(
  tmt_qf_mq[['psms_filtered_forRobust_mq']], min_features = min_psms,
  master_protein_col = 'Leading.razor.protein')
```

Both are then summarised to protein level.

```{r, warning=FALSE}
tmt_qf_mq <- QFeatures::aggregateFeatures(
  tmt_qf_mq, i = 'psms_filtered_forSum_mq', fcol = 'Leading.razor.protein',
  name = 'protein_mq_sum', fun = base::colSums)
tmt_qf_mq[['protein_mq_sum']] <- QFeatures::logTransform(tmt_qf_mq[['protein_mq_sum']], base = 2)

tmt_qf_mq <- QFeatures::aggregateFeatures(
  tmt_qf_mq, i = 'psms_filtered_forRobust_mq', fcol = 'Leading.razor.protein',
  name = 'protein_mq_robust', fun = MsCoreUtils::robustSummary, maxit = 1000)
```

```{r, fig.height=4, fig.width=3, out.width='35%'}
n_sum <- nrow(tmt_qf_mq[['protein_mq_sum']])
n_robust <- nrow(tmt_qf_mq[['protein_mq_robust']])

data.frame(method = factor(c('Sum', 'Robust'), levels = c('Sum', 'Robust')),
          n_proteins = c(n_sum, n_robust)) %>%
  ggplot(aes(method, n_proteins, fill = method)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = n_proteins), vjust = -0.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  scale_fill_manual(values = get_cat_palette(2), guide = 'none') +
  theme_biomasslmb(border = FALSE) +
  ylab('Proteins quantified') + xlab('')
```

**Tolerating incomplete PSMs is worth `r n_robust - n_sum` proteins here** — `r n_robust` against `r n_sum` for `sum`. The proteins quantified only via `robustSummary` are the ones to examine, since they show what the completeness requirement actually costs.

```{r}
sum_proteins_mq <- rownames(tmt_qf_mq[['protein_mq_sum']])
robust_proteins_mq <- rownames(QFeatures::filterNA(tmt_qf_mq[['protein_mq_robust']]))

rescued_proteins <- setdiff(robust_proteins_mq, sum_proteins_mq)
length(rescued_proteins)
```

The mechanism is clearest in one of these `r length(rescued_proteins)` proteins with exactly 2 retained PSMs — the minimum allowed — where at least one PSM has missing values.

```{r}
rd_mq <- rowData(tmt_qf_mq[['psms_filtered_forRobust_mq']])
mat_mq <- assay(tmt_qf_mq[['psms_filtered_forRobust_mq']])

example_protein <- rescued_proteins[vapply(rescued_proteins, function(p) {
  psm_rows <- which(rd_mq$Leading.razor.protein == p)
  length(psm_rows) == 2 && any(is.na(mat_mq[psm_rows, ]))
}, logical(1))][2]

retain_psms <- rd_mq$Leading.razor.protein == example_protein
round(2^mat_mq[retain_psms, ], 1)
```

One PSM is missing in most samples, while the other is complete. With `sum`, the incomplete PSM fails the zero-missing-values filter and is discarded, leaving only 1 PSM for this protein — below the minimum of 2, so the protein is dropped entirely. With `robustSummary`, both PSMs are retained (the incomplete one has few enough missing values to pass the 8/12 threshold), giving 2 PSMs to summarise from and allowing the protein to be quantified.

Plotted the same way as Part 1 — PSM-level abundances in grey, summarised protein-level abundance overlaid in colour. Only the `robustSummary` line appears, because `sum` has no value at all for this protein.

```{r}
compare_protein_abundances_mq <- qfeatures_long(
  tmt_qf_mq[,,c('protein_mq_sum', 'protein_mq_robust')]) %>%
  data.frame() %>%
  mutate(method=recode_values(
    assay,
    'protein_mq_sum'~'Sum',
    'protein_mq_robust'~'Robust'))

plot_pep_and_protein_mq <- function(protein_of_interest) {

  to_plot_compare <- compare_protein_abundances_mq %>%
    filter(rowname == protein_of_interest)

  QFeatures::filterFeatures(
    tmt_qf_mq,
    VariableFilter("Leading.razor.protein",
                   protein_of_interest,
                   condition = "=="))[['psms_filtered_forRobust_mq']] %>%
    qfeatures_long() %>%
    ggplot(aes(x = colname, y = value)) +
    geom_line(aes(group = rowname), colour = 'grey') +
    geom_point(colour = 'grey') +
    geom_line(data = to_plot_compare,
              aes(x = colname, y = value, colour = method, group = method)) +
    scale_colour_manual(values = get_cat_palette(4),
                        name = 'Protein summarisation method') +
    theme_biomasslmb(base_size = 15, border = FALSE, base_family = 'sans') +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    labs(
      title = protein_of_interest,
      x = '',
      y = 'PSM/Protein abundance (log2)'
    )
}
```

```{r, fig.width=7, fig.height=6, warning=FALSE, message=FALSE}
plot_pep_and_protein_mq(example_protein)
```

The two PSMs (grey) track each other where both are present, and the `robustSummary` protein-level estimate (blue) follows information presented from both PSMs.

## Where both methods quantify a protein, but disagree

`sum` and `robustSummary` can also disagree on proteins that both methods manage to quantify, simply because they draw on different PSMs: `sum` requires complete PSMs, while `robustSummary` also incorporates partially-missing ones. `P27695` is a good example: it has 3 PSMs retained by `robustSummary`, only 2 of which are complete enough to be used by `sum`.

```{r}
example_protein_3psm <- 'P27695'

sum_mat <- assay(tmt_qf_mq[['protein_mq_sum']])
robust_mat <- assay(tmt_qf_mq[['protein_mq_robust']])
```

```{r}
rd_mq2 <- rowData(tmt_qf_mq[['psms_filtered_forRobust_mq']])
mat_mq2 <- assay(tmt_qf_mq[['psms_filtered_forRobust_mq']])
psm_rows_2 <- rd_mq2$Leading.razor.protein == example_protein_3psm
round(2^mat_mq2[psm_rows_2, ], 1)
```

The third PSM is missing in exactly the two `Control` channels (`Control_4` and `Control_5`) - consistent with genuinely low abundance in the control samples, rather than a random dropout - so `sum` discards it and summarises the protein from the other two, complete, PSMs alone. `robustSummary` retains all three. Since `sum` weights PSMs by their absolute abundance, its profile is dominated by whichever of the two complete PSMs is more abundant, and here those two PSMs disagree on the direction of the IP-vs-Control difference; `robustSummary` also incorporates the partially-missing PSM, which is clearly enriched in the IP samples. The two methods end up describing quite different abundance profiles for the same protein (correlation of `r round(cor(sum_mat[example_protein_3psm, ], robust_mat[example_protein_3psm, ]), 2)` between them).

```{r, fig.width=7, fig.height=6, warning=FALSE, message=FALSE}
plot_pep_and_protein_mq(example_protein_3psm)
```

Comparing the `Control` and `IP` abundances for each method puts the consequence in one plot.

```{r, fig.width=6, fig.height=4.5, warning=FALSE}
condition_lookup <- setNames(tmt_per2_mq_design$Condition, rownames(tmt_per2_mq_design))

compare_protein_abundances_mq %>%
  filter(rowname == example_protein_3psm) %>%
  mutate(condition = condition_lookup[colname]) %>%
  ggplot(aes(x = method, y = value, colour = condition)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.5), size = 2) +
  stat_summary(fun = mean, geom = 'crossbar', width = 0.4,
               position = position_dodge(width = 0.5)) +
  scale_colour_manual(values = get_cat_palette(2),
                      name = '') +
  theme_biomasslmb(base_size = 15, border = FALSE, base_family = 'sans') +
  labs(title = example_protein_3psm, x = '', y = 'Protein abundance (log2)')
```

**`robustSummary` separates `IP` from `Control` here and `sum` largely does not.** `sum` is working from the two complete PSMs alone, which disagree with each other on direction, so its estimate is diluted. `robustSummary` also draws on the missing-value PSM — the one PSM that is unambiguously enriched in the IP samples, and unambiguously enriched for the same reason it goes undetected in two of the `Control` channels. Discarding it as incomplete discards the observation that carries the finding.

# Part 3: The same missing-value trade-off with LFQ peptides

Parts 1 and 2 both used TMT PSMs as the lower-level feature, which leaves open whether the Part 2 conclusion is about missingness or about TMT. Repeating the same comparison on `lfq_qf`, the LFQ-DDA whole-proteome dataset from the [LFQ-DDA QC](LFQ_DDA_Peptide_QC_Summarisation.html) vignette, settles it: there the lower-level feature is a peptide rather than a PSM. Missingness here differs from Part 2 in kind as well as degree: it is not concentrated in one condition, but it is pervasive, because every sample was quantified in a separate run.

`lfq_qf[['peptides_filtered_norm']]` (the peptides used for summarisation in the QC vignette) is stored on the log2 scale, since that is what `robustSummary` requires. A `sum` comparison needs the values back on their original scale.

```{r}
lfq_qf[['peptides_filtered_forSum']] <- lfq_qf[['peptides_filtered_norm']]
assay(lfq_qf[['peptides_filtered_forSum']]) <- 2^assay(lfq_qf[['peptides_filtered_forSum']])
```

As in Part 2, the same rule is applied to both approaches: keep only proteins with at least 2 retained peptides. They differ again in one respect alone — how much missingness a peptide may carry before it is discarded, none for `sum` and up to 4/6 for `robustSummary`, the same threshold used in the QC vignette.

```{r}
min_peps <- 2

lfq_qf[['peptides_filtered_forSum']] <- QFeatures::filterNA(
  lfq_qf[['peptides_filtered_forSum']], 0)
lfq_qf[['peptides_filtered_forSum']] <- biomasslmb::filter_features_per_protein(
  lfq_qf[['peptides_filtered_forSum']], min_features = min_peps)
```

Both are then summarised to protein level. `lfq_qf[['peptides_for_summarisation']]` is the same peptide set (at most 4/6 missing, at least `r min_peps` peptides per protein) used for `robustSummary` in the QC vignette.

```{r, warning=FALSE}
lfq_qf <- QFeatures::aggregateFeatures(
  lfq_qf, i = 'peptides_filtered_forSum', fcol = 'Master.Protein.Accessions',
  name = 'protein_lfq_sum', fun = base::colSums)
lfq_qf[['protein_lfq_sum']] <- QFeatures::logTransform(lfq_qf[['protein_lfq_sum']], base = 2)

set.seed(42)
lfq_qf <- QFeatures::aggregateFeatures(
  lfq_qf, i = 'peptides_for_summarisation', fcol = 'Master.Protein.Accessions',
  name = 'protein_lfq_robust', fun = MsCoreUtils::robustSummary, maxit = 10000)
```

```{r, fig.height=4, fig.width=3, out.width='35%'}
n_sum_lfq <- nrow(lfq_qf[['protein_lfq_sum']])
n_robust_lfq <- nrow(QFeatures::filterNA(lfq_qf[['protein_lfq_robust']]))

data.frame(method = factor(c('Sum', 'Robust'), levels = c('Sum', 'Robust')),
          n_proteins = c(n_sum_lfq, n_robust_lfq)) %>%
  ggplot(aes(method, n_proteins, fill = method)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = n_proteins), vjust = -0.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  scale_fill_manual(values = get_cat_palette(2), guide = 'none') +
  theme_biomasslmb(border = FALSE) +
  ylab('Proteins quantified') + xlab('')
```

**The same trade appears with peptides.** Tolerating incomplete peptides is worth `r n_robust_lfq - n_sum_lfq` proteins here — `r n_robust_lfq` against `r n_sum_lfq` for `sum`. The proteins quantified only via `robustSummary`, and the peptides behind one of them:

```{r}
sum_proteins_lfq <- rownames(lfq_qf[['protein_lfq_sum']])
robust_proteins_lfq <- rownames(QFeatures::filterNA(lfq_qf[['protein_lfq_robust']]))

rescued_proteins_lfq <- setdiff(robust_proteins_lfq, sum_proteins_lfq)
length(rescued_proteins_lfq)
```

```{r}
rd_lfq <- rowData(lfq_qf[['peptides_for_summarisation']])
mat_lfq <- assay(lfq_qf[['peptides_for_summarisation']])

example_protein_lfq <- rescued_proteins_lfq[vapply(rescued_proteins_lfq, function(p) {
  pep_rows <- which(rd_lfq$Master.Protein.Accessions == p)
  length(pep_rows) == 2 && any(is.na(mat_lfq[pep_rows, ]))
}, logical(1))][1]

retain_peps_lfq <- rd_lfq$Master.Protein.Accessions == example_protein_lfq
round(2^mat_lfq[retain_peps_lfq, ], 1)
```

One peptide is missing in every `Control` replicate, while the other is complete. With `sum`, the incomplete peptide fails the zero-missing-values filter and is discarded, leaving only 1 peptide for this protein — below the minimum of `r min_peps`, so the protein is dropped entirely. With `robustSummary`, both peptides are retained (the incomplete one has few enough missing values to pass the 4/6 threshold), giving 2 peptides to summarise from and allowing the protein to be quantified - exactly the same mechanism as the TMT example above, just with peptides in place of PSMs.

```{r}
compare_protein_abundances_lfq <- qfeatures_long(
  lfq_qf[,,c('protein_lfq_sum', 'protein_lfq_robust')]) %>%
  data.frame() %>%
  mutate(method=recode_values(
    assay,
    'protein_lfq_sum'~'Sum',
    'protein_lfq_robust'~'Robust'))

plot_pep_and_protein_lfq <- function(protein_of_interest) {

  to_plot_compare <- compare_protein_abundances_lfq %>%
    filter(rowname == protein_of_interest)

  QFeatures::filterFeatures(
    lfq_qf,
    VariableFilter("Master.Protein.Accessions",
                   protein_of_interest,
                   condition = "=="))[['peptides_for_summarisation']] %>%
    qfeatures_long() %>%
    ggplot(aes(x = colname, y = value)) +
    geom_line(aes(group = rowname), colour = 'grey') +
    geom_point(colour = 'grey') +
    geom_line(data = to_plot_compare,
              aes(x = colname, y = value, colour = method, group = method)) +
    scale_colour_manual(values = get_cat_palette(4),
                        name = 'Protein summarisation method') +
    theme_biomasslmb(base_size = 15, border = FALSE, base_family = 'sans') +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    labs(
      title = protein_of_interest,
      x = '',
      y = 'Peptide/Protein abundance (log2)'
    )
}
```

```{r, fig.width=7, fig.height=6, warning=FALSE, message=FALSE}
plot_pep_and_protein_lfq(example_protein_lfq)
```

One of the two peptides is quantified in every sample; the other is missing from two of the wild type replicates. `sum` needs complete peptides, so discarding the incomplete one leaves the protein with a single peptide, below the two-peptide requirement, and it drops out entirely. `robustSummary` uses both peptides and estimates the protein in all six samples, with the partially-observed peptide contributing where it was measured.

```{r}
both_proteins_lfq <- intersect(sum_proteins_lfq, robust_proteins_lfq)
cors_lfq <- sapply(both_proteins_lfq, function(p) {
  cor(assay(lfq_qf[['protein_lfq_sum']])[p, ], assay(lfq_qf[['protein_lfq_robust']])[p, ])
})
summary(cors_lfq)
```

Where both methods do manage to quantify a protein here, most agree well, but a meaningful minority do not: `r sum(cors_lfq < 0.9)` of the `r length(cors_lfq)` jointly-quantified proteins correlate below 0.9, and the lowest is `r round(min(cors_lfq), 2)`. That is the Part 1 effect - the methods weight the underlying features differently - showing up alongside the Part 2 effect rather than instead of it, and it is more visible here than in the TMT data because LFQ peptide intensities for one protein span a wider range.

Both effects are larger with LFQ peptides than with TMT PSMs, but the rescuing of otherwise-discarded proteins (`r n_robust_lfq` vs `r n_sum_lfq`) remains the dominant one, consistent with LFQ carrying more missing values to begin with.

## Summary

`sum` and `mean` weight the highest intensity features most heavily, which are usually the most accurately quantified — and are also where a high-abundance outlier does the most damage. `robustSummary` treats features more equally regardless of abundance, which costs it some of that automatic weighting towards precision but lets it use features that are partly missing, and so quantify proteins `sum` has to discard. `median` gives up more than either, for the reasons in Part 1.

These are properties of the estimators rather than of any one quantification technology: Part 3 reproduced the Part 2 trade-off with LFQ peptides in place of TMT PSMs, and more strongly, since LFQ carries more missing values to begin with.

**Which one to use follows from how much missingness there is, not from a preference.** Where missingness is very low, as in Part 1, `sum` is the better choice: tolerating missing features buys nothing when there are almost none, and complete data is simpler to reason about downstream. Where missing values are common — informative, as in the Part 2 IP dataset, or merely pervasive, as in the Part 3 LFQ-DDA dataset — `robustSummary` is the better choice, because it recovers proteins `sum` would discard and uses more of the lower-level quantification that was actually measured. Measuring the missingness before choosing is the point; [handling missing values](handling_missing_values.html) covers how.

# Where to go next

- [Handling missing values](handling_missing_values.html) is the article this
  one defers to. Summarisation is one of four stages that make a decision about
  missing values, and choosing `robustSummary` here interacts with what is done
  at the other three.
- The core workflow articles apply the choice made here:
  [TMT](TMT_PSM_QC_Summarisation.html) sums PSMs within a plex,
  [LFQ-DDA](LFQ_DDA_Peptide_QC_Summarisation.html) and
  [LFQ-DIA](LFQ_DIA_Precursor_QC_Summarisation.html) use `robustSummary`, and
  each says what the dataset's missingness rate had to look like for that to be
  right.
- Both estimators assume the features they are given belong to the protein they
  are labelled with. [Peptides are not proteins](gotcha_peptide_to_protein.html)
  measures how often that assumption fails.
- [Data exploration and statistical testing](exploration_and_statistical_testing.html)
  takes the protein-level abundances onward, whichever estimator produced them.

# Getting help

If your experiment does not fit what this article assumes, or a result looks
wrong, please get in touch with Tom Smith (tsmith@mrclmb.ac.uk) rather than
guessing — it is far easier to help while an analysis is in progress than to
unpick a decision afterwards, and easier still before the samples are run.
[Getting started](biomasslmb.html) sets out which article applies to which
experiment.

```{r}
sessionInfo()
```
