---
title: "Comparing protein summarisation approaches"
author: "Tom Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: biomasslmb.json
vignette: >
  %\VignetteIndexEntry{Comparing protein summarisation approaches}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Here, we will consider different approaches for summarising lower-level quantification (PSMs for TMT, peptides for LFQ) to protein-level abundance: `sum`, `median` and `MsCoreUtils::robustSummary`. (`mean` is not considered separately: it's just `sum` divided by the number of contributing features, so it produces the same abundance profile across samples as `sum`, only rescaled.) These are general considerations for bottom-up proteomics, independent of the quantification technology: the same weighting and missing-value behaviours apply whether the lower-level features are PSMs or peptides. The comparison between summarisation approaches is informative in two distinct ways, which we consider separately using three datasets, two from the [TMT QC](TMT_PSM_QC_Summarisation.html) vignette and one from the [LFQ-DDA QC](LFQ_DDA_Peptide_QC_Summarisation.html) vignette:

- **How the methods weight individual features.** Even with complete data, `sum` and `robustSummary`/`median` can produce different protein-level estimates, since they don't treat the underlying features equally. We explore this using `tmt_qf`, the whole-proteome TMT PD dataset, which has very little missingness.
- **How the methods handle missing values.** `sum` requires complete data (or values need to be discarded/ignored), whereas `robustSummary` can make use of features with some missing values. We explore this first using `tmt_qf_mq`, the TMT MaxQuant IP dataset, where missingness is common and potentially informative, and then again using `lfq_qf`, an LFQ-DDA TurboID (proximity biotinylation) dataset, to confirm the same conclusion holds when the lower-level features are peptides from a different quantification technology rather than TMT PSMs.

# Load required packages

```{r, message=FALSE}
library(QFeatures)
library(biomasslmb)
library(ggplot2)
library(tidyr)
library(dplyr)
```

# Part 1: How the methods weight individual PSMs

`tmt_qf` already contains protein-level quantification from a simple `sum` summarisation (`protein`), where missing values were removed and proteins with a single PSM were removed. Here, we compare this against `robustSummary` and `median`.

## Summarisation with robustSummary

With `robustSummary`, we do not need to remove all PSMs with missing values since the summarisation algorithm deals with them appropriately. However, we still don't want to retain PSMs with too many missing values, since these will not be very informative in estimating the protein-level quantification. Here, we will retain PSMs with at most 5/10 missing values.

```{r}
tmt_qf[['psms_filtered_forRobust']] <- QFeatures::filterNA(
  tmt_qf[['psms_filtered_rank']], 5/10)

```

Next, we remove PSMs for proteins with fewer than 2 PSMs.
```{r}

min_psms <- 2

tmt_qf[['psms_filtered_forRobust']] <- biomasslmb::filter_features_per_protein(
  tmt_qf[['psms_filtered_forRobust']], min_features = min_psms)

```

For the `robustSummary` summarisation, we need our quantification values to be approximately Gaussian distributed. Hence, we log transform them.

```{r}
tmt_qf[['psms_filtered_forRobust']] <- QFeatures::logTransform(
  tmt_qf[['psms_filtered_forRobust']], base=2)
```

Now we can summarise with `robustSummary`

```{r, warning=FALSE}
# Aggregate to protein-level abundances (using QFeatures function)
tmt_qf <- QFeatures::aggregateFeatures(tmt_qf,
                                       i = "psms_filtered_forRobust",
                                       fcol = "Master.Protein.Accessions",
                                       name = "protein_robust",
                                       fun = MsCoreUtils::robustSummary,
                                       maxit=1000)


```

## Summarisation with median

We will also include summarisation with median, since it's a typical approach for summarising across observations and will aid the interpretation when we compare the summarisation approaches. We will use the PSMs with missing values filtered, even though `median` can handle missing values with `na.rm=TRUE`, since ignoring missing values is inappropriate for summarisation of PSMs to proteins with this method.

```{r}

tmt_qf <- QFeatures::aggregateFeatures(tmt_qf,
                                       i = "psms_filtered_forSum",
                                       fcol = "Master.Protein.Accessions",
                                       name = "protein_median",
                                       fun = matrixStats::colMedians, na.rm=TRUE)

tmt_qf[['protein_median']] <- QFeatures::logTransform(
  tmt_qf[['protein_median']], base=2)
```

## Comparing the summarisation approaches

As established in the QC vignette, this dataset has very few missing values to begin with. As a result, allowing `robustSummary` to tolerate missing PSMs makes no difference to which proteins are retained here. See Part 2 below, where we use the MaxQuant IP dataset (`tmt_qf_mq`), which has much more missingness, to see that trade-off in action. Here, we can instead ask how much the protein-level abundance estimates from each method actually agree.

```{r}

# Single object with protein inference from both methods
compare_protein_abundances <- biomasslmb:::qfeatures_long(
  tmt_qf[,,c('protein', 'protein_robust', 'protein_median')]) %>%
  data.frame() %>%
  mutate(method=recode_values(
    assay,
    'protein'~'Sum',
    'protein_robust'~'Robust',
    'protein_median'~'Median'))

```

We want to identify the proteins with the lowest correlation for abundances across the samples between summarisation methods so we can inspect the difference between them.

```{r}
proteins_of_interest <- compare_protein_abundances %>%
    select(method, value, rowname,colname) %>%
  pivot_wider(names_from = method, values_from = value) %>%
  group_by(rowname) %>%
  summarise(cor_sum_median=cor(Sum, Median),
            cor_sum_robust=cor(Sum, Robust)) %>%
  pivot_longer(cols=-rowname) %>%
  group_by(rowname) %>%
  summarise(min_cor=min(value)) %>%
  slice_min(min_cor, n=5) %>%
  pull(rowname)
```

Below, we define a function to plot the PSM-level abundances (on their original scale) and the protein-level abundances from each summarisation method (normalised to mean abundance, so the methods are directly comparable) for a single protein, as two separate plots.

```{r}
plot_pep_and_protein <- function(protein_of_interest) {

  to_plot_compare <- compare_protein_abundances %>%
    filter(rowname == protein_of_interest) %>%
    group_by(method) %>%
    mutate(value = value - mean(value)) %>%
    ungroup()

  pep_plot <- QFeatures::filterFeatures(
    tmt_qf,
    VariableFilter("Master.Protein.Accessions",
                   protein_of_interest,
                   condition = "=="))[['psms_filtered_sn']] %>%
    biomasslmb:::qfeatures_long() %>%
    ggplot(aes(x = colname, y = log2(value))) +
    geom_line(aes(group = rowname), colour = 'grey') +
    geom_point(colour = 'grey') +
    theme_biomasslmb(base_size = 15, border = FALSE, base_family = 'sans') +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    labs(
      title = protein_of_interest,
      x = '',
      y = 'PSM abundance (log2)'
    )

  protein_plot <- to_plot_compare %>%
    ggplot(aes(x = colname, y = value, colour = method, group = method)) +
    geom_line() +
    scale_colour_manual(values = get_cat_palette(3),
                        name = 'Protein summarisation method') +
    theme_biomasslmb(base_size = 15, border = FALSE, base_family = 'sans') +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    labs(
      title = protein_of_interest,
      x = '',
      y = 'Protein abundance (log2, mean-normalised)'
    ) +
    guides(color = guide_legend(override.aes = list(linewidth = 2) ) )

  list(peptide = pep_plot, protein = protein_plot)
}
```

In the peptide-level plots below, each grey line is a single PSM's abundance on its original scale. In the protein-level plots, each summarisation method's abundance profile is mean-normalised, so the methods sit on a comparable scale and can be compared directly. The important thing to focus on is the abundance profile across the tags for any one summarisation method or PSM.

```{r, plot.show='hold', out.width='45%', fig.width=7, fig.height=7, warning=FALSE, message=FALSE}
for(poi in proteins_of_interest) {
  plots <- plot_pep_and_protein(poi)
  print(plots$peptide)
  print(plots$protein)
}
```


In all cases, the protein-level quantification values are relatively similar for most tags, with differences between the approaches typically restricted to one or two tags where the PSMs disagree on the relative abundances between the samples.

Since `sum` simply adds together the abundance values across the PSMs for a given tag and the abundance values may span multiple orders of magnitude, the protein-level abundance pattern across the tags is weighted towards the most abundant PSMs. This is likely to be a positive attribute, since the most abundant PSMs will be more accurately quantified. However, the summarisation is also sensitive to high abundance outliers. `robustSummary` and `median`, by contrast, are less sensitive to any single PSM, but treat all PSMs more equally regardless of abundance.

We can see this directly for one of the proteins above. For each PSM of a given protein, we can correlate its abundance profile across the tags with the `sum`-summarised protein profile; a PSM that disagrees with the majority of the other PSMs will have low, or negative, correlation.

```{r}
poi <- proteins_of_interest[grepl("^Q9C", proteins_of_interest)][1]

rd <- rowData(tmt_qf[['psms_filtered_sn']])
psm_mask <- rd$Master.Protein.Accessions == poi
psm_values <- log2(assay(tmt_qf[['psms_filtered_sn']])[psm_mask, , drop = FALSE])
sum_profile <- assay(tmt_qf[['protein']])[poi, ]

sort(apply(psm_values, 1, function(x) cor(x, sum_profile, use = "complete.obs")))
```

The PSM with the lowest (most negative) correlation is in the same broad abundance range as several of the other PSMs for this protein, so it isn't simply down-weighted by `sum` for being low intensity: because this method combines all PSMs, its disagreement with the majority still shows up in the summarised profile, whereas `robustSummary` and `median` are less influenced by a single disagreeing PSM.

The median is less sensitive to outliers than the mean, but ignores much of the quantitative data since it's only affected by the 'middle' PSMs. In the most extreme instances, median just simplifies the PSM level abundances down to a single representative 'middle' PSM.

# Part 2: How the methods handle missing values

For Part 2, we switch to `tmt_qf_mq`, the MaxQuant IP dataset from the QC vignette. Unlike `tmt_qf`, missing values here are common and potentially informative (a protein absent from the control pulldown is genuinely missing, not just undetected by chance), so this is a better dataset for comparing how `sum` and `robustSummary` cope with missingness.

`tmt_qf_mq[['psms_filtered_interference']]` (the PSMs used for summarisation in the QC vignette) is stored on the log2 scale, since that's what `robustSummary` requires. For a `sum` comparison we need the values back on their original scale.

```{r}
tmt_qf_mq[['psms_filtered_interference_raw']] <- tmt_qf_mq[['psms_filtered_interference']]
assay(tmt_qf_mq[['psms_filtered_interference_raw']]) <- 2^assay(tmt_qf_mq[['psms_filtered_interference_raw']])
```

For a fair comparison, we apply the same rule used in Part 1 to both approaches: keep only proteins with at least 2 retained PSMs. The two approaches differ only in how much missingness a PSM is allowed to have before it's discarded: none, for `sum`; up to 8/12, for `robustSummary` (the same threshold used in the QC vignette).

```{r}
tmt_qf_mq[['psms_filtered_forSum_mq']] <- QFeatures::filterNA(
  tmt_qf_mq[['psms_filtered_interference_raw']], 0)
tmt_qf_mq[['psms_filtered_forSum_mq']] <- biomasslmb::filter_features_per_protein(
  tmt_qf_mq[['psms_filtered_forSum_mq']], min_features = min_psms,
  master_protein_col = 'Leading.razor.protein')

tmt_qf_mq[['psms_filtered_forRobust_mq']] <- QFeatures::filterNA(
  tmt_qf_mq[['psms_filtered_interference']], 8/12)
tmt_qf_mq[['psms_filtered_forRobust_mq']] <- biomasslmb::filter_features_per_protein(
  tmt_qf_mq[['psms_filtered_forRobust_mq']], min_features = min_psms,
  master_protein_col = 'Leading.razor.protein')
```

Now we summarise both to protein level.

```{r, warning=FALSE}
tmt_qf_mq <- QFeatures::aggregateFeatures(
  tmt_qf_mq, i = 'psms_filtered_forSum_mq', fcol = 'Leading.razor.protein',
  name = 'protein_mq_sum', fun = base::colSums)
tmt_qf_mq[['protein_mq_sum']] <- QFeatures::logTransform(tmt_qf_mq[['protein_mq_sum']], base = 2)

tmt_qf_mq <- QFeatures::aggregateFeatures(
  tmt_qf_mq, i = 'psms_filtered_forRobust_mq', fcol = 'Leading.razor.protein',
  name = 'protein_mq_robust', fun = MsCoreUtils::robustSummary, maxit = 1000)
```

```{r, fig.height=4, fig.width=3, out.width='35%'}
n_sum <- nrow(tmt_qf_mq[['protein_mq_sum']])
n_robust <- nrow(tmt_qf_mq[['protein_mq_robust']])

data.frame(method = factor(c('Sum', 'Robust'), levels = c('Sum', 'Robust')),
          n_proteins = c(n_sum, n_robust)) %>%
  ggplot(aes(method, n_proteins, fill = method)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = n_proteins), vjust = -0.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  scale_fill_manual(values = get_cat_palette(2), guide = 'none') +
  theme_biomasslmb(border = FALSE) +
  ylab('Proteins quantified') + xlab('')
```

Tolerating PSMs with some missing values lets `robustSummary` retain more proteins (`r n_robust`) than requiring complete PSMs for `sum` (`r n_sum`). We can identify the proteins that are only quantified via `robustSummary`.

```{r}
sum_proteins_mq <- rownames(tmt_qf_mq[['protein_mq_sum']])
robust_proteins_mq <- rownames(QFeatures::filterNA(tmt_qf_mq[['protein_mq_robust']]))

rescued_proteins <- setdiff(robust_proteins_mq, sum_proteins_mq)
length(rescued_proteins)
```

To understand why, we inspect the PSMs for one of these `r length(rescued_proteins)` proteins directly: one with exactly 2 retained PSMs (the minimum allowed), where at least one PSM has missing values.

```{r}
rd_mq <- rowData(tmt_qf_mq[['psms_filtered_forRobust_mq']])
mat_mq <- assay(tmt_qf_mq[['psms_filtered_forRobust_mq']])

example_protein <- rescued_proteins[vapply(rescued_proteins, function(p) {
  psm_rows <- which(rd_mq$Leading.razor.protein == p)
  length(psm_rows) == 2 && any(is.na(mat_mq[psm_rows, ]))
}, logical(1))][2]

retain_psms <- rd_mq$Leading.razor.protein == example_protein
round(2^mat_mq[retain_psms, ], 1)
```

One PSM is missing in most samples, while the other is complete. With `sum`, the incomplete PSM fails the zero-missing-values filter and is discarded, leaving only 1 PSM for this protein - below our minimum of 2, so the protein is dropped entirely. With `robustSummary`, both PSMs are retained (the incomplete one has few enough missing values to pass the 8/12 threshold), giving 2 PSMs to summarise from and allowing the protein to be quantified.

We can visualise this the same way as Part 1: PSM-level abundances in grey, with the summarised protein-level abundance overlaid in colour. Since `sum` has no value at all for this protein, only the `robustSummary` line appears.

```{r}
compare_protein_abundances_mq <- biomasslmb:::qfeatures_long(
  tmt_qf_mq[,,c('protein_mq_sum', 'protein_mq_robust')]) %>%
  data.frame() %>%
  mutate(method=recode_values(
    assay,
    'protein_mq_sum'~'Sum',
    'protein_mq_robust'~'Robust'))

plot_pep_and_protein_mq <- function(protein_of_interest) {

  to_plot_compare <- compare_protein_abundances_mq %>%
    filter(rowname == protein_of_interest)

  QFeatures::filterFeatures(
    tmt_qf_mq,
    VariableFilter("Leading.razor.protein",
                   protein_of_interest,
                   condition = "=="))[['psms_filtered_forRobust_mq']] %>%
    biomasslmb:::qfeatures_long() %>%
    ggplot(aes(x = colname, y = value)) +
    geom_line(aes(group = rowname), colour = 'grey') +
    geom_point(colour = 'grey') +
    geom_line(data = to_plot_compare,
              aes(x = colname, y = value, colour = method, group = method)) +
    scale_colour_manual(values = get_cat_palette(4),
                        name = 'Protein summarisation method') +
    theme_biomasslmb(base_size = 15, border = FALSE, base_family = 'sans') +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    labs(
      title = protein_of_interest,
      x = '',
      y = 'PSM/Protein abundance (log2)'
    )
}
```

```{r, fig.width=7, fig.height=6, warning=FALSE, message=FALSE}
plot_pep_and_protein_mq(example_protein)
```

The two PSMs (grey) track each other where both are present, and the `robustSummary` protein-level estimate (blue) follows information presented from both PSMs.

## Where both methods quantify a protein, but disagree

`sum` and `robustSummary` can also disagree on proteins that both methods manage to quantify, simply because they draw on different PSMs: `sum` requires complete PSMs, while `robustSummary` also incorporates partially-missing ones. `P27695` is a good example: it has 3 PSMs retained by `robustSummary`, only 2 of which are complete enough to be used by `sum`.

```{r}
example_protein_3psm <- 'P27695'

sum_mat <- assay(tmt_qf_mq[['protein_mq_sum']])
robust_mat <- assay(tmt_qf_mq[['protein_mq_robust']])
```

```{r}
rd_mq2 <- rowData(tmt_qf_mq[['psms_filtered_forRobust_mq']])
mat_mq2 <- assay(tmt_qf_mq[['psms_filtered_forRobust_mq']])
psm_rows_2 <- rd_mq2$Leading.razor.protein == example_protein_3psm
round(2^mat_mq2[psm_rows_2, ], 1)
```

The third PSM is missing in exactly the two `Control` channels (`Control_4` and `Control_5`) - consistent with genuinely low abundance in the control samples, rather than a random dropout - so `sum` discards it and summarises the protein from the other two, complete, PSMs alone. `robustSummary` retains all three. Since `sum` weights PSMs by their absolute abundance, its profile is dominated by whichever of the two complete PSMs is more abundant, and here those two PSMs disagree on the direction of the IP-vs-Control difference; `robustSummary` also incorporates the partially-missing PSM, which is clearly enriched in the IP samples. The two methods end up describing quite different abundance profiles for the same protein (correlation of `r round(cor(sum_mat[example_protein_3psm, ], robust_mat[example_protein_3psm, ]), 2)` between them).

```{r, fig.width=7, fig.height=6, warning=FALSE, message=FALSE}
plot_pep_and_protein_mq(example_protein_3psm)
```

We can summarise this more directly by comparing the `Control` and `IP` abundances for each method.

```{r, fig.width=6, fig.height=4.5, warning=FALSE}
condition_lookup <- setNames(tmt_per2_mq_design$Condition, rownames(tmt_per2_mq_design))

compare_protein_abundances_mq %>%
  filter(rowname == example_protein_3psm) %>%
  mutate(condition = condition_lookup[colname]) %>%
  ggplot(aes(x = method, y = value, colour = condition)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.5), size = 2) +
  stat_summary(fun = mean, geom = 'crossbar', width = 0.4,
               position = position_dodge(width = 0.5)) +
  scale_colour_manual(values = get_cat_palette(2),
                      name = '') +
  theme_biomasslmb(base_size = 15, border = FALSE, base_family = 'sans') +
  labs(title = example_protein_3psm, x = '', y = 'Protein abundance (log2)')
```

`robustSummary` shows a much clearer separation between `IP` and `Control` than `sum` does. This is because `sum`'s estimate is diluted by the two complete PSMs alone, which disagree with each other on direction, whereas `robustSummary` also draws on the missing-value PSM - the one PSM that is unambiguously enriched in the IP samples, and is precisely why it goes undetected in two of the `Control` channels.

# Part 3: The same missing-value trade-off with LFQ peptides

Parts 1 and 2 both used TMT PSMs as the lower-level feature. To confirm the missing-value conclusion from Part 2 isn't specific to TMT, we repeat the same comparison using `lfq_qf`, the LFQ-DDA TurboID (proximity biotinylation) dataset from the [LFQ-DDA QC](LFQ_DDA_Peptide_QC_Summarisation.html) vignette, where the lower-level feature is a peptide rather than a PSM. As in that vignette, missingness here is common and condition-informative: a genuine proximity partner of the bait is expected to be reliably detected in the biotin-labelled replicates but frequently missing from the (mostly background) unlabelled control replicates.

`lfq_qf[['peptides_filtered']]` (the peptides used for summarisation in the QC vignette) is stored on the log2 scale, since that's what `robustSummary` requires. For a `sum` comparison we need the values back on their original scale.

```{r}
lfq_qf[['peptides_filtered_forSum']] <- lfq_qf[['peptides_filtered']]
assay(lfq_qf[['peptides_filtered_forSum']]) <- 2^assay(lfq_qf[['peptides_filtered_forSum']])
```

As in Part 2, we apply the same rule to both approaches: keep only proteins with at least 2 retained peptides. The two approaches again differ only in how much missingness a peptide is allowed to have before it's discarded: none, for `sum`; up to 4/6, for `robustSummary` (the same threshold used in the QC vignette).

```{r}
min_peps <- 2

lfq_qf[['peptides_filtered_forSum']] <- QFeatures::filterNA(
  lfq_qf[['peptides_filtered_forSum']], 0)
lfq_qf[['peptides_filtered_forSum']] <- biomasslmb::filter_features_per_protein(
  lfq_qf[['peptides_filtered_forSum']], min_features = min_peps)
```

Now we summarise both to protein level. `lfq_qf[['peptides_for_summarisation']]` is the same peptide set (at most 4/6 missing, at least `r min_peps` peptides per protein) used for `robustSummary` in the QC vignette.

```{r, warning=FALSE}
lfq_qf <- QFeatures::aggregateFeatures(
  lfq_qf, i = 'peptides_filtered_forSum', fcol = 'Master.Protein.Accessions',
  name = 'protein_lfq_sum', fun = base::colSums)
lfq_qf[['protein_lfq_sum']] <- QFeatures::logTransform(lfq_qf[['protein_lfq_sum']], base = 2)

set.seed(42)
lfq_qf <- QFeatures::aggregateFeatures(
  lfq_qf, i = 'peptides_for_summarisation', fcol = 'Master.Protein.Accessions',
  name = 'protein_lfq_robust', fun = MsCoreUtils::robustSummary, maxit = 10000)
```

```{r, fig.height=4, fig.width=3, out.width='35%'}
n_sum_lfq <- nrow(lfq_qf[['protein_lfq_sum']])
n_robust_lfq <- nrow(QFeatures::filterNA(lfq_qf[['protein_lfq_robust']]))

data.frame(method = factor(c('Sum', 'Robust'), levels = c('Sum', 'Robust')),
          n_proteins = c(n_sum_lfq, n_robust_lfq)) %>%
  ggplot(aes(method, n_proteins, fill = method)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = n_proteins), vjust = -0.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  scale_fill_manual(values = get_cat_palette(2), guide = 'none') +
  theme_biomasslmb(border = FALSE) +
  ylab('Proteins quantified') + xlab('')
```

As in the TMT MaxQuant IP dataset, tolerating peptides with some missing values lets `robustSummary` retain more proteins (`r n_robust_lfq`) than requiring complete peptides for `sum` (`r n_sum_lfq`). We can again identify the proteins that are only quantified via `robustSummary`, and inspect the peptides for one of them directly.

```{r}
sum_proteins_lfq <- rownames(lfq_qf[['protein_lfq_sum']])
robust_proteins_lfq <- rownames(QFeatures::filterNA(lfq_qf[['protein_lfq_robust']]))

rescued_proteins_lfq <- setdiff(robust_proteins_lfq, sum_proteins_lfq)
length(rescued_proteins_lfq)
```

```{r}
rd_lfq <- rowData(lfq_qf[['peptides_for_summarisation']])
mat_lfq <- assay(lfq_qf[['peptides_for_summarisation']])

example_protein_lfq <- rescued_proteins_lfq[vapply(rescued_proteins_lfq, function(p) {
  pep_rows <- which(rd_lfq$Master.Protein.Accessions == p)
  length(pep_rows) == 2 && any(is.na(mat_lfq[pep_rows, ]))
}, logical(1))][1]

retain_peps_lfq <- rd_lfq$Master.Protein.Accessions == example_protein_lfq
round(2^mat_lfq[retain_peps_lfq, ], 1)
```

One peptide is missing in every `Control` replicate, while the other is complete. With `sum`, the incomplete peptide fails the zero-missing-values filter and is discarded, leaving only 1 peptide for this protein - below our minimum of `r min_peps`, so the protein is dropped entirely. With `robustSummary`, both peptides are retained (the incomplete one has few enough missing values to pass the 4/6 threshold), giving 2 peptides to summarise from and allowing the protein to be quantified - exactly the same mechanism as the TMT example above, just with peptides in place of PSMs.

```{r}
compare_protein_abundances_lfq <- biomasslmb:::qfeatures_long(
  lfq_qf[,,c('protein_lfq_sum', 'protein_lfq_robust')]) %>%
  data.frame() %>%
  mutate(method=recode_values(
    assay,
    'protein_lfq_sum'~'Sum',
    'protein_lfq_robust'~'Robust'))

plot_pep_and_protein_lfq <- function(protein_of_interest) {

  to_plot_compare <- compare_protein_abundances_lfq %>%
    filter(rowname == protein_of_interest)

  QFeatures::filterFeatures(
    lfq_qf,
    VariableFilter("Master.Protein.Accessions",
                   protein_of_interest,
                   condition = "=="))[['peptides_for_summarisation']] %>%
    biomasslmb:::qfeatures_long() %>%
    ggplot(aes(x = colname, y = value)) +
    geom_line(aes(group = rowname), colour = 'grey') +
    geom_point(colour = 'grey') +
    geom_line(data = to_plot_compare,
              aes(x = colname, y = value, colour = method, group = method)) +
    scale_colour_manual(values = get_cat_palette(4),
                        name = 'Protein summarisation method') +
    theme_biomasslmb(base_size = 15, border = FALSE, base_family = 'sans') +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    labs(
      title = protein_of_interest,
      x = '',
      y = 'Peptide/Protein abundance (log2)'
    )
}
```

```{r, fig.width=7, fig.height=6, warning=FALSE, message=FALSE}
plot_pep_and_protein_lfq(example_protein_lfq)
```

The complete peptide (grey) is present throughout the biotin replicates and absent from the control replicates, and the `robustSummary` protein-level estimate (blue) follows it into the control replicates too, drawing on the partially-missing peptide rather than discarding the protein outright as `sum` must.

```{r}
both_proteins_lfq <- intersect(sum_proteins_lfq, robust_proteins_lfq)
cors_lfq <- sapply(both_proteins_lfq, function(p) {
  cor(assay(lfq_qf[['protein_lfq_sum']])[p, ], assay(lfq_qf[['protein_lfq_robust']])[p, ])
})
sort(cors_lfq)
```

Where both methods do manage to quantify a protein here, they agree closely (lowest correlation `r round(min(cors_lfq), 2)`) - unlike the TMT IP example above, this dataset doesn't happen to contain a case where an outlier peptide flips the direction of the summarised profile. The rescuing of otherwise-discarded proteins, rather than disagreement on jointly-quantified ones, is the dominant effect here, and its size (`r n_robust_lfq` vs `r n_sum_lfq` proteins) is if anything larger than in the TMT IP dataset - consistent with LFQ peptides typically carrying more missing values than TMT PSMs to begin with.

## Summary

`sum`/`mean` are more sensitive to the abundances for the highest intensity features, which are likely to be the most accurately quantified. However, they are therefore also more sensitive to outlier values with very high abundance. `robustSummary` handles missing values and therefore enables more proteins to be quantified, but it is more sensitive to low intensity features, which may be less accurately quantified, and treats all features more equally regardless of abundance.

These are general properties of the summarisation methods, not artefacts of any one quantification technology: Part 3 showed the same missing-value trade-off from Part 2 reproduced with LFQ peptides in place of TMT PSMs, and if anything more pronounced, since LFQ typically has more missing values to begin with. For datasets with very low missing values (Part 1), `sum`/`mean` are likely to be the best approach: retaining more proteins isn't a meaningful benefit when there's little missingness to tolerate in the first place, so it's simpler to work with complete data throughout. Where missing values are more common, and especially where they may be informative rather than incidental - such as the IP dataset in Part 2 and the TurboID dataset in Part 3 - `robustSummary` is the better choice, since it recovers proteins that `sum` would otherwise have to discard and uses more information from the lower-level quantification, regardless of whether that quantification comes from TMT PSMs or LFQ peptides.

```{r}
sessionInfo()
```
