---
title: "TMT workflow: PSM QC and protein summarisation"
author: "Tom Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: biomasslmb.json
vignette: >
  %\VignetteIndexEntry{TMT workflow: PSM QC and protein summarisation}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Quantitative proteomics using isobaric tagging such as Tandem Mass Tags (TMT) has a considerable benefit over Label-Free Quantification (LFQ) in that many samples can be quantified for each Peptide Spectrum Match (PSM). The standard TMTpro reagent set provides 18 channels [@li2021tmtpro18]. Higher multiplexing is now commercially available, up to 35 channels, achieved by adding a set of deuterated reagents [@zuniga2024tmt35]. The trade-off is that deuterium substitution shifts retention time and degrades coelution between the deuterated and non-deuterated channels, so the two sub-plexes have to be designed and normalised separately using bridge channels. This has multiple benefits over analysing samples in separate runs (LFQ):

1.  TMT reduces protein quantification variance since PSM-level quantification is derived from the same MS1 ion for all samples
2.  LFQ suffers from much higher missing values when comparing across samples due to the limited number of ions that can be fragmented in each run and the associated issue of peptides being identified in only a subset of runs [@http://zotero.org/users/5634351/items/ME322P2A]. This is ameliorated to a significant degree by Data-Independent Acquisition (DIA) LFQ. However, DIA still involves quantifying each sample separately, so missing values are not entirely removed, and proteins in each sample may be quantified from different sets of peptides.

Because TMT quantifies from the same MS1 ion for all samples, this standardises the features quantified in each sample, which simplifies the comparison between samples and increases quantification accuracy of summarised features such as proteins.

However, TMT does suffer from ratio compression, caused by co-isolation of contaminating precursor ions that get fragmented alongside the target peptide and inflate the reporter ion signal. This should be avoided by performing quantification with SPS MS3 [@http://zotero.org/users/5634351/items/ZK69WYZ2].

This vignette works through one typical experiment from end to end: a whole-proteome comparison between two conditions, labelled in a single TMT plex and searched with Proteome Discoverer (PD). That combination covers the majority of TMT experiments, and the choices made below are the conventional ones for it.

Experiments that depart from it are covered separately, since the departures change the reasoning rather than just the code:

- **Several plexes.** A design too large for one plex needs each plex processed on its own and then brought onto a common scale — see [multi-plex TMT](TMT_multiplex.html).
- **An enrichment design.** An IP, BioID or TurboID experiment breaks the assumptions behind normalisation and behind discarding missing values — see [enrichment designs](interactome_designs.html).
- **An enriched PTM fraction.** Modified peptides have to be localised to a residue, summarised to sites rather than proteins, and normalised against a matched total fraction rather than against themselves — see [PTM site quantification](PTM_site_quantification.html).
- **MaxQuant instead of PD.** The principles are the same, but the column names differ and two of the QC metrics used below — reporter signal:noise and SPS mass matches — are not reported at all, so the PSM quality filter has to be rebuilt around the one that is. [Reading MaxQuant output](#reading-maxquant-output) at the end of this vignette covers what transfers and what does not.

# Load required packages

```{r, message=FALSE}
library(QFeatures)
library(biomasslmb)
library(ggplot2)
library(tidyr)
library(dplyr)
```

# The experimental design

`psm_tmt_clock` and `tmt_clock_design` are datasets available from the `biomasslmb` package, derived from a real TMT12plex experiment comparing `Control` and `Mutant` samples (6 replicates each). `psm_tmt_clock` is the PD PSM-level output, truncated to a subset of proteins for a manageable vignette. `tmt_clock_design` gives the experimental design: which sample each TMT tag corresponds to, and its `Condition` and `Replicate`.

```{r}
knitr::kable(tmt_clock_design)
```

Unlike a `data.frame` read from a single PD output file, real experiments nearly always have an associated design table like this (often as a spreadsheet from whoever ran the samples), which needs to be read in and used both to make sense of the TMT tags and to define experimental details for exploration and statistical testing.

# Read in input data

The data go into a `QFeatures` object, the standard Bioconductor container for quantitative proteomics data. See [here](https://www.bioconductor.org/packages/release/bioc/html/QFeatures.html) for documentation about the `QFeatures` object.

`psm_tmt_clock` contains the PSM-level output from PD for this experiment, with the abundance columns already renamed to match the sample names used in `tmt_clock_design`. Passing `tmt_clock_design` as the `colData` keeps the experimental design travelling with the quantification data.

`readQFeatures()` attaches the design at the object level only, so `sync_coldata()` copies it onto the assay as well — functions that plot or model a single assay, such as `plot_quant()` below, read it from there. [Working with QFeatures objects](qfeatures_objects.html) covers this and the rest of the object's vocabulary.

```{r}

tmt_qf <- QFeatures::readQFeatures(assayData = psm_tmt_clock,
                                   colData = tmt_clock_design,
                                   quantCols = rownames(tmt_clock_design),
                                   name = "psms_raw")

tmt_qf <- sync_coldata(tmt_qf, "psms_raw")

```


# Defining the contaminant proteins

Contaminant proteins have to be removed. This search was performed against the '0602_Universal Contaminants' database [@Frankenfield2022], so parsing that same fasta gives the accessions to match against, in the `Cont_<UniProt accession>` format PD reports them in. PD sometimes assigns the bare UniProt accession as the master protein even for a contaminant hit — matching on both forms costs nothing and catches those.

```{r}

contaminant_fasta_inf <- system.file(
  "extdata", "0602_Universal_Contaminants.fasta.gz",
  package = "biomasslmb"
)

# Extract the protein IDs associated with each contaminant protein
contaminant_accessions <- biomasslmb::get_contaminant_fasta_accessions(contaminant_fasta_inf)
contaminant_accessions <- c(contaminant_accessions, sub('^Cont_', '', contaminant_accessions))

print(head(contaminant_accessions))
```

Routine filtering removes PSMs that:

-   Could originate from contaminants. See `?filter_features_pd_dda` for further details, including the removal of 'associated' contaminants.
-   Have more than one accession in the master protein column, meaning the search engine could not resolve which protein the PSM came from.

The accession list, the `Cont_` prefix and PD's own `Contaminant` column are three separate defences, and they do not always agree with one another. [Contaminants and protein FDR](gotcha_contaminants_and_FDR.html) measures what each one catches and what is left behind when the database naming does not match what the filter expects.

`unique_master` asks whether the search engine settled on a single protein identity. The separate `proteotypic` argument asks whether the peptide sequence occurs in only one protein at all — a feature can have one unambiguous master protein and still be shared with the other members of that protein group. See [peptides are not proteins](gotcha_peptide_to_protein.html).

```{r}
# Perform routine raw data filtering.
# - Remove PSMs from contaminant proteins
# - Remove PSMs with no master protein, or with more than one
tmt_qf[['psms_filtered']] <- filter_features_pd_dda(tmt_qf[['psms_raw']],
                                             contaminant_proteins=contaminant_accessions,
                                             filter_contaminant=TRUE,
                                             filter_associated_contaminant=TRUE,
                                             unique_master=TRUE)

tmt_qf <- sync_coldata(tmt_qf, 'psms_filtered')
```

# Normalise

`biomasslmb::plot_quant` shows whether the PSM intensity distributions are approximately the same across channels. Colouring by `Condition` answers a second question at the same time — whether the two conditions already separate before any normalisation or filtering, which would point at a labelling or loading problem rather than at biology.

```{r, warning=FALSE, fig.show='hold', fig.width=7, fig.height=5, out.width="75%"}

# Plot the peptide-level quantification distributions per sample
biomasslmb::plot_quant(tmt_qf[['psms_filtered']],
                       log2transform=TRUE,
                       method='density') +
  theme_biomasslmb() +
  aes(colour=Condition, group=sample) +
  xlab('PSM abundance (log2)')

```

The same amount of material was labelled in every channel, and this is a whole-proteome comparison in which most proteins are expected to be unchanged, so any systematic offset between those distributions is technical. `diff.median` normalisation with `QFeatures::normalize` shifts each channel to a common median and removes it. The abundances are log-transformed first, so that the shift is multiplicative on the original scale, then exponentiated back afterwards — `sum` summarisation below needs untransformed values.

That assumption is worth stating explicitly, because it is the step most easily carried into an experiment where it does not hold: an [enrichment design](interactome_designs.html) expects the conditions to differ in overall composition, and normalising to a common median there erodes the signal being measured.

```{r, warning=FALSE, fig.show='hold', fig.width=7, fig.height=5, out.width="75%"}

# Normalise the log2-transformed abundances using diff.median
tmt_qf[['psms_filtered_norm']] <- QFeatures::normalize(
  logTransform(tmt_qf[['psms_filtered']], base=2), method='diff.median')

# Exponentiate the quantification values back to the initial scale.
assay(tmt_qf[['psms_filtered_norm']]) <- 2^assay(tmt_qf[['psms_filtered_norm']])

tmt_qf <- sync_coldata(tmt_qf, 'psms_filtered_norm')

# Plot the peptide-level quantification distributions per sample
biomasslmb::plot_quant(tmt_qf[['psms_filtered_norm']],
                       log2transform=TRUE,
                       method='density') +
  theme_biomasslmb() +
  aes(colour=Condition, group=sample) +
  xlab('PSM abundance (log2)')

```

# Removing low quality PSMs

Low Signal:Noise (S:N) PSMs are worth removing on two counts at once: their quantification is less accurate, and they account for most of the missing values. `plot_missing_SN` shows the relationship between the two.

**Missing channels are concentrated below S:N 10 and scarce above it.** That coincidence is what makes a single S:N threshold the right tool here — it removes the PSMs that generate missing values and the PSMs whose measured values are least trustworthy, in one filter, without any modelling.

```{r, fig.height=6, fig.width=8, fig.fullwidth=TRUE, fig.cap="Missing values per PSM, in relation to the signal:noise ratio", out.width='75%'}
# Add a more accurate average S:N ratio value.
# The one calculated by PD doesn't treat NA values appropriately!
tmt_qf[['psms_filtered_norm']] <- update_average_sn(tmt_qf[['psms_filtered_norm']])

plot_missing_SN(tmt_qf[['psms_filtered_norm']], bins = 20)
```

`plot_missing_SN_per_sample` breaks the same relationship down by tag, which is how a labelling failure in one channel shows up: that tag would keep dropping values where the others do not. No tag behaves that way here above S:N 10.

```{r, fig.height=6, fig.width=8, fig.fullwidth=TRUE, fig.cap="Missing values per tag, in relation to the signal:noise ratio", out.width='75%'}
plot_missing_SN_per_sample(tmt_qf[['psms_filtered_norm']], bins = 20)
```

`filter_TMT_PSMs` applies both thresholds the plots above justify: S:N \> 10, and interference/co-isolation \< 50%. The second targets a different problem — a PSM whose isolation window contained more than one precursor reports a blend of their reporter ions, so its ratios are compressed towards no change rather than merely noisy. [Enrichment designs](interactome_designs.html) shows what the interference distribution looks like when the threshold has to be argued rather than assumed.

```{r}
# Then filter PSMs to remove low S:N and/or high interference
tmt_qf[['psms_filtered_sn']] <- filter_TMT_PSMs(tmt_qf[['psms_filtered_norm']],
                                                inter_thresh=50, sn_thresh=10)
tmt_qf <- sync_coldata(tmt_qf, 'psms_filtered_sn')
```
Peptides that are not rank 1 according to the search engine are removed next.

```{r}

tmt_qf[['psms_filtered_rank']] <- tmt_qf[['psms_filtered_sn']]

tmt_qf <- tmt_qf %>%
  filterFeatures(~ Rank == 1, i = 'psms_filtered_rank')

message_parse(rowData(tmt_qf[['psms_filtered_rank']]),
                         'Master.Protein.Accessions',
                         "Removing peptides that are not rank 1")

```

# Summarising to protein-level abundances

With the PSM-level quantification inspected and filtered, the PSMs can be summarised to protein-level abundances.

Summarisation takes the master protein assignment at face value: every PSM contributes to exactly one protein, the one the search engine assigned it to. That assignment is an inference rather than a measurement, and for proteins whose peptides are largely shared with their homologues it is not a reliable one. [Peptides are not proteins](gotcha_peptide_to_protein.html) covers how the assignment is made and how to tell which of your proteins it can support.

For TMT, summing the PSM-level abundances gives accurate protein estimates so long as there are no missing values. Where there are many, `MsCoreUtils::robustSummary` is the better estimator, since it summarises accurately without complete data [@http://zotero.org/users/5634351/items/FZN3QTTZ]. Which of those applies is a property of the dataset rather than a preference, so it is settled below by measurement. [Comparing protein summarisation approaches](summarisation_methods.html) runs both on this dataset and on one where the answer goes the other way.

## Exploring missing values

Before deciding how to handle missing values, it's worth checking how much missingness there is, and whether it's random or structured by experimental condition. Condition-structured missingness (e.g. a PSM detected in every `Control` sample but no `Mutant` sample) can indicate a real biological or technical difference between conditions, and should make you wary of simply discarding those PSMs. Missingness scattered independently of condition, on the other hand, more likely reflects individual TMT channels whose intensity falls around the limit of detection: the S:N filter above is applied per-PSM, so a PSM can pass that threshold overall while one or two of its channels still drop out.

`plot_missing_upset` visualises the most common patterns of missingness across samples. Samples with no missing PSMs at all are omitted from the plot, since they don't contribute to any missingness pattern.

```{r, fig.height=6, fig.width=8, warning=FALSE, message=FALSE}
plot_missing_upset(tmt_qf, i = 'psms_filtered_rank')
```

`condition_miss_score` complements this by fitting, for each PSM, a logistic regression of missingness against a `colData` grouping column, and returning Tjur's R² as a per-feature score: values near 1 mean a PSM's missingness pattern is well explained by `Condition`, values near 0 mean it looks unrelated to `Condition`. `condition_miss_index` aggregates these per-feature scores into a single dataset-level value between 0 and 1.

```{r}
miss_res <- condition_miss_score(tmt_qf, i = 'psms_filtered_rank', group_cols = 'Condition')
condition_miss_index(miss_res$summary)$index
```

**Missing values are both rare and unstructured here**, which is what settles the summarisation choice. Only `r miss_res$global['n_features_informative']` of `r miss_res$global['n_features_total']` PSMs have any missing value at all, and the condition missingness index is close to zero, so there is no evidence that the missingness there is tracks `Condition`.

That makes the simplest route the defensible one: discard the PSMs with any missing value and sum the rest. The cost is those PSMs, and it is small precisely because they are few — in a dataset where they were many, or where they clustered in one condition, the same step would discard a large fraction of the data or the signal itself. The [LFQ-DDA workflow](LFQ_DDA_Peptide_QC_Summarisation.html) works through a dataset where the trade comes out the other way.

Complete data is a requirement rather than a convenience for `sum`. Summing over whatever PSMs happen to be present would build a protein's abundance from different PSMs in different samples, so the resulting values would not be comparable across the very samples they exist to be compared between.

```{r}
tmt_qf[['psms_filtered_missing']] <- QFeatures::filterNA(
  tmt_qf[['psms_filtered_rank']], 0)

```

PSMs belonging to proteins with fewer than 2 PSMs are then removed, so that every protein quantification rests on at least two independent observations. The filter is not always right: in phosphoproteomics a site is often supported by one PSM and discarding it discards the measurement, which is why [PTM site quantification](PTM_site_quantification.html) does not apply it.

```{r}
min_psms <- 2
tmt_qf[['psms_filtered_forSum']] <- biomasslmb::filter_features_per_protein(
  tmt_qf[['psms_filtered_missing']], min_features = min_psms)

```

Then the summarisation itself.

```{r, message=FALSE}
# Aggregate to protein-level abundances (using QFeatures function)
tmt_qf <- QFeatures::aggregateFeatures(tmt_qf,
                                       i = "psms_filtered_forSum",
                                       fcol = "Master.Protein.Accessions",
                                       name = "protein",
                                       fun = base::colSums)

tmt_qf <- sync_coldata(tmt_qf, 'protein')
```

```{r}
tmt_qf[['protein']] <- QFeatures::logTransform(
  tmt_qf[['protein']], base=2)
```

## Inspecting the number of PSMs and proteins through the processing steps

Each filtering step above removed something, and the totals are easier to judge together than one message at a time. `get_samples_present()` and `plot_samples_present()` show how many PSMs and proteins survived each stage, and in how many samples. Both take a named character vector selecting the assays to plot, since the `QFeatures` names are not self-explanatory.

The assay names to choose from:

```{r}
names(tmt_qf)
```

`psms_filtered_norm` is left out: it holds the normalised quantification of `psms_filtered` and no PSMs were removed at that step, so it would plot as a duplicate.

### Samples per PSM

PSMs first. The row variables are `Sequence`, `Modifications` and `RT.in.min`, so that the count is of unique PSMs rather than of rows.

```{r, fig.height=7, , fig.width=7, fig.fullwidth=TRUE, fig.cap="Samples quantified for each PSM at each level of processing", out.width='75%'}


rename_cols <- c('All PSMs' = 'psms_raw' ,
                 'Quantified, contaminants removed' = 'psms_filtered',
                 'Signal:Noise > 10' = 'psms_filtered_sn',
                 'PSMs - filtered (Rank1)' = 'psms_filtered_rank',
                 'No missing values' = 'psms_filtered_missing',
                 '>1 PSM per protein' = 'psms_filtered_forSum')

rowvars <- c('Sequence', 'Modifications', 'RT.in.min')

samples_present <- get_samples_present(tmt_qf[,,unname(rename_cols)], rowvars, rename_cols)
plot_samples_present(samples_present, rowvars, breaks=seq(2,12,2)) + ylab('PSM')

```

### Samples per Protein

Then proteins, from the same functions: the named vector gains the `protein` assay, and the row variable becomes `Master.Protein.Accessions` alone.

```{r, fig.height=7, , fig.width=7, fig.fullwidth=TRUE, fig.cap="Samples quantified for each protein at each level of processing", out.width='75%'}


rename_cols_prot <- c(rename_cols, 'Protein'='protein')

rowvars_prot <- c('Master.Protein.Accessions')

samples_present <- get_samples_present(tmt_qf, rowvars_prot, rename_cols_prot)
plot_samples_present(samples_present, rowvars_prot, breaks=seq(2,12,2))
```

**The two-PSM rule is cheap in PSMs and expensive in proteins.** It removes few PSMs, because most PSMs belong to proteins that have several; it removes many more proteins, because a protein identified by one PSM loses its only evidence. Those proteins are the ones whose quantification would have been least reliable, so the trade is usually worth making — but it is a trade, and an experiment targeting a low-abundance protein family may not want it.

```{r, eval = FALSE}
# Save file to package as data so it can be read back in in other vignettes
usethis::use_data(tmt_qf, overwrite=TRUE)
```

# Summary

The PSMs were QCed and filtered in the following steps:

-   Remove PSMs that are likely from contaminants
-   Remove PSMs without a unique 'master protein'
-   Remove PSMs without any quantification values
-   Normalise the abundance values so they have the same median value in all samples, which is reasonable here because the same amount of material was labelled in every channel
-   Remove PSMs with very low signal:noise and high interference

The PSM-level abundances were then summarised to protein level with `sum`, having first removed the small number of PSMs with any missing value. See the [Comparing protein summarisation approaches](summarisation_methods.html) vignette for a direct comparison against `robustSummary`, and the [Data exploration and statistical testing](exploration_and_statistical_testing.html) vignette for how to explore and statistically test the resulting protein-level data using the experimental design attached here.

Two of those steps are specific to this being a straightforward whole-proteome comparison in a single plex. Normalising to a common median assumes most proteins are unchanged, which an [enrichment design](interactome_designs.html) violates, and summing complete PSMs is only cheap because within-plex missingness is low — between plexes it is not, which is what [multi-plex TMT](TMT_multiplex.html) addresses.

# Reading MaxQuant output {#reading-maxquant-output}

Everything above is Proteome Discoverer specific in its column names, and one
section of it more deeply than that. This part covers what transfers to
MaxQuant and what does not, using `psm_tmt_factorial`: the `evidence.txt`
PSM-level output from a TMT18plex whole-proteome experiment, with
`tmt_factorial_design`. It is subsetted to a random sample of proteins, so the
counts below are smaller than a full experiment's, but every filtering step is
doing real work on real PSMs.

| Quantity | Proteome Discoverer | MaxQuant |
|---|---|---|
| Protein assignment | `Master.Protein.Accessions` | `Leading.razor.protein` |
| All candidate proteins | `Protein.Accessions` | `Proteins` |
| Contaminant accessions | search FASTA, `Cont_` prefix | `get_maxquant_cont_accessions()`, `CON__` prefix |
| Decoy flag | removed before export | `Reverse` |
| Search engine rank | `Rank` | not reported per PSM |
| Co-isolation | `Isolation.Interference.in.Percent` | `PIF` |
| Reporter signal:noise | `Average.Reporter.SN` | **not reported** |
| SPS mass matches | `SPS.Mass.Matches.in.Percent` | **not reported** |
| Missing quantification | empty | `0` |

The design is crossed rather than a single comparison: three cell lines, each
treated with a vehicle or a compound, in three replicates.

```{r}
knitr::kable(tmt_factorial_design)
```

Nothing in the reading or filtering below depends on that structure. It starts
to matter at the point of testing, where a second factor changes both what the
model should contain and what its coefficients mean, and
[data exploration and statistical testing](exploration_and_statistical_testing.html)
picks this dataset up there.

```{r, message=FALSE, warning=FALSE}
tmt_mq <- QFeatures::readQFeatures(assayData = psm_tmt_factorial,
                                   colData = tmt_factorial_design,
                                   quantCols = rownames(tmt_factorial_design),
                                   name = 'psms_raw')

tmt_mq <- sync_coldata(tmt_mq, 'psms_raw')
```

One step has no counterpart in the PD workflow and has to come first. MaxQuant
writes an intensity of exactly zero where it has no measurement, whereas PD
leaves the cell empty. Mass spectrometry cannot assert that a peptide was
absent, only that it was not detected, so those zeros have to become `NA` —
left as they are they are treated as real measurements at the bottom of the
scale, and they pull every median, every summarisation and every fold change
towards it.

```{r}
tmt_mq[['psms_raw']] <- QFeatures::zeroIsNA(tmt_mq[['psms_raw']])
```

`filter_features_mq_dda()` replaces `filter_features_pd_dda()`, and takes its
contaminant accessions from `get_maxquant_cont_accessions()` rather than from a
search FASTA, because MaxQuant searches against its own contaminants database
with `CON__` prefixed accessions. It also removes hits to the decoy database,
which PD has already removed before the PSM export exists.

```{r, message=FALSE, warning=FALSE}
tmt_mq[['psms_filtered']] <- filter_features_mq_dda(
  tmt_mq[['psms_raw']],
  contaminant_proteins = get_maxquant_cont_accessions(),
  filter_contaminant = TRUE,
  filter_associated_contaminant = TRUE)

tmt_mq <- sync_coldata(tmt_mq, 'psms_filtered')
```

Normalisation is unchanged, and rests on the same assumption: equal loading per
channel, and most proteins unchanged between conditions.

```{r, warning=FALSE}
tmt_mq[['psms_norm']] <- QFeatures::normalize(
  logTransform(tmt_mq[['psms_filtered']], base = 2), method = 'diff.median')

assay(tmt_mq[['psms_norm']]) <- 2^assay(tmt_mq[['psms_norm']])

tmt_mq <- sync_coldata(tmt_mq, 'psms_norm')
```

**The low-quality PSM section is the part that does not transfer.** MaxQuant
does not report reporter ion signal:noise at all, so the S:N threshold that
does most of the quality filtering above has no counterpart, and neither has
the SPS mass match percentage. What remains is `PIF`, the precursor intensity
fraction, which measures the same thing as PD's co-isolation percentage from
the other direction: the fraction of the isolation window accounted for by the
intended precursor, so a high `PIF` is good where a high interference
percentage is bad.

```{r}
summary(rowData(tmt_mq[['psms_norm']])$PIF)
```

Two consequences follow, and both are worth taking seriously rather than
working around.

The first is that filtering has to lean on `PIF` alone, which controls for
co-isolation but says nothing about how close a reporter ion was to the noise.
The PD filter above discarded PSMs on both counts, and only one of them is
available here.

The second is that `PIF` is itself frequently `NA`, reported only where
MaxQuant could assess the window — `r sum(is.na(rowData(tmt_mq[['psms_norm']])$PIF))`
of `r nrow(tmt_mq[['psms_norm']])` PSMs here — so a threshold on it carries a
second decision about what to do with the PSMs it cannot assess. Discarding
them is the conservative choice and the one taken below.

The threshold itself is a judgement rather than a standard. The direct
counterpart of the PD filter used above, interference under 50%, is `PIF` above
0.5, which would remove
`r sum(rowData(tmt_mq[['psms_norm']])$PIF <= 0.5, na.rm = TRUE)` PSMs; the
stricter `PIF` above 0.75 used here removes
`r sum(rowData(tmt_mq[['psms_norm']])$PIF <= 0.75, na.rm = TRUE)`. The stricter
threshold buys cleaner ratios at the price of those PSMs, and is worth it in a
whole-proteome experiment where PSMs are plentiful. In an experiment where they
are not — a low-input sample, or a protein of interest identified by two
peptides — the same threshold can remove the measurement being sought.

```{r}
pif <- rowData(tmt_mq[['psms_norm']])$PIF

tmt_mq[['psms_pif']] <- tmt_mq[['psms_norm']][!is.na(pif) & pif > 0.75, ]

tmt_mq <- sync_coldata(tmt_mq, 'psms_pif')

nrow(tmt_mq[['psms_pif']])
```

The rank filter has no MaxQuant equivalent and is simply omitted: `evidence.txt`
reports the accepted match rather than every candidate, and the confidence
filtering has already been applied through MaxQuant's posterior error
probabilities.

Missing values are the one thing to re-check rather than carry over. The
argument above for discarding PSMs with any missing value rested on there being
very few, and that in turn rested on the S:N filter, which is not available
here — so it has to be re-made on this dataset.

```{r}
miss_mq <- condition_miss_score(tmt_mq, i = 'psms_pif', group_cols = 'Treatment')

condition_miss_index(miss_mq$summary)$index
```

**The argument survives.** Only
`r as.integer(miss_mq$global['n_features_informative'])` of
`r as.integer(miss_mq$global['n_features_total'])` PSMs have any missing value, and the
condition missingness index is near zero, so discarding them costs little and
loses nothing that tracks `Treatment`.

```{r}
tmt_mq[['psms_complete']] <- QFeatures::filterNA(tmt_mq[['psms_pif']], 0)

tmt_mq[['psms_forSum']] <- filter_features_per_protein(
  tmt_mq[['psms_complete']],
  master_protein_col = 'Leading.razor.protein',
  min_features = 2)

tmt_mq <- sync_coldata(tmt_mq, 'psms_forSum')
```

Summarisation is then the same operation with a different column name.
`Leading.razor.protein` replaces `Master.Protein.Accessions` everywhere it
appears as a `master_protein_col` or an `fcol`, which is the single most common
cause of a MaxQuant analysis failing partway through a workflow written for PD.

```{r, message=FALSE}
tmt_mq <- QFeatures::aggregateFeatures(tmt_mq,
                                       i = 'psms_forSum',
                                       fcol = 'Leading.razor.protein',
                                       name = 'protein',
                                       fun = base::colSums)

tmt_mq <- sync_coldata(tmt_mq, 'protein')

tmt_mq[['protein']] <- QFeatures::logTransform(tmt_mq[['protein']], base = 2)

nrow(tmt_mq[['protein']])
```

```{r, eval = FALSE}
# Save to the package so the testing vignette can read it back in. Only the
# protein assay is kept: the PSM-level assays are already available as
# `psm_tmt_factorial`, and carrying them here would make the saved object an
# order of magnitude larger.
tmt_qf_factorial <- tmt_mq[, , 'protein']

usethis::use_data(tmt_qf_factorial, overwrite = TRUE)
```

# Where to go next

Both routes above end in the same place: protein-level abundances with the
experimental design attached, summarised with `sum`.

- [Data exploration and statistical testing](exploration_and_statistical_testing.html)
  picks up from exactly that object — checking the experiment worked, deciding
  which proteins can be tested, and testing them.
- [Multiplexed TMT](TMT_multiplex.html) covers what changes once a design runs
  to more than one plex. Summing complete PSMs is cheap here only because
  within-plex missingness is low; between plexes it is not, and a bridge
  channel rather than a summarisation choice is what makes the plexes
  comparable.
- [Comparing protein summarisation approaches](summarisation_methods.html)
  measures `sum` against `robustSummary` on this dataset and on one where the
  answer goes the other way.
- The contaminant filtering above rests on three separate defences that do not
  always agree; [contaminants and protein FDR](gotcha_contaminants_and_FDR.html)
  shows what each one catches and how a silent failure looks.

# Getting help

If your experiment does not fit what this article assumes, or a result looks
wrong, please get in touch with Tom Smith (tsmith@mrclmb.ac.uk) rather than
guessing — it is far easier to help while an analysis is in progress than to
unpick a decision afterwards, and easier still before the samples are run.
[Getting started](biomasslmb.html) sets out which article applies to which
experiment.

```{r}
sessionInfo()
```
