---
title: "TMT - comparing protein summarisation approaches"
author: "Tom Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: biomasslmb.json
vignette: >
  %\VignetteIndexEntry{TMT QC PSM-level quantification and summarisation to protein-level abundance}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Here, we will consider different protein summarisation approaches for TMT data, depending on whether missing values need to be retained. The comparison between summarisation approaches is also informative when considering what it means to summarise from PSMs to protein-level abundances.


For PSM to protein summarisation with TMT quantification, simply summing together the PSM-level abundances provides accurate estimates, so long as there are no missing values.  If many missing values are present, the `MsCoreUtils::robustSummary` method may be preferred, since it is able 
to summarise accurately, even in the presence of missing values [@http://zotero.org/users/5634351/items/FZN3QTTZ]. 

# Load required packages

To clarify which functionality is provided by which package, we will use `package::function`. For your own code, there is no need to specify the package unless you want to maintain this clarity.

```{r, message=FALSE}
library(QFeatures)
library(biomasslmb)
library(ggplot2)
library(tidyr)
library(dplyr)
```

`tmt_qf` is a `QFeatures` object available from the `biomasslmb` package which was generated in the `TMT QC PSM-level quantification and summarisation to protein-level abundance` vignette. It already contains protein-level quantification from a simple 'sum' summarisation, where missing values were removed and proteins with a single PSM were removed. Here, we will compare this with the 'robust' summaristion, which requires some additional steps to ensure it is performed appropriately.

## Summarisation with robustSummary

With `robustSummary`, we do not need to remove all PSMs with missing values since the summarisation algorithm deals with them appropriately. However, we still don't want to retain PSMs with too many missing values, since these will not be very informative in estimating the protein-level quantification. Here, we will retain PSMs with at most 5/10 missing values 

```{r}
tmt_qf[['psms_filtered_forRobust']] <- QFeatures::filterNA(
  tmt_qf[['psms_filtered_sn']], 5/10)

```

Next, we remove PSMs for proteins with fewer than 2 PSMs.
```{r}

min_psms <- 2
 
tmt_qf[['psms_filtered_forRobust']] <- biomasslmb::filter_features_per_protein(
  tmt_qf[['psms_filtered_forRobust']], min_features = min_psms)

```

For the `robustSummary` summarisation, we need our quantification values to be approximately Gaussian distributed. Hence, we log transform them.

```{r}
tmt_qf[['psms_filtered_forRobust']] <- QFeatures::logTransform(
  tmt_qf[['psms_filtered_forRobust']], base=2)
```

Now we can summarise with `robustSummary`

```{r}
# Aggregate to protein-level abundances (using QFeatures function)
tmt_qf <- QFeatures::aggregateFeatures(tmt_qf, 
                                       i = "psms_filtered_forRobust", 
                                       fcol = "Master.Protein.Accessions",
                                       name = "protein_robust",
                                       fun = MsCoreUtils::robustSummary,
                                       maxit=1000)


```

Prior to summaristaion, we removed PSMs from proteins with fewer than 2 PSMs. However, since we left in PSMs with missing values, it's possible for some protein-level abundances to be derived from just a single PSM. We can use the `get_protein_no_quant_mask` from `biomasslmb` to identify where the protein abundances will be derived from fewer than `n` features (PSMs). We can then give this mask to `mask_protein_level_quant` to replace these quantification values with NA.
 
```{r, fig.height=5, fig.width=5, out.width='50%'}
# plot = TRUE means we will also get a plot of the number of proteins quantified in each sample
protein_retain_mask <- biomasslmb::get_protein_no_quant_mask(
  tmt_qf[['psms_filtered_forRobust']], min_features=2, plot=TRUE) 

tmt_qf[['protein_robust']] <- biomasslmb::mask_protein_level_quant(
  tmt_qf[['protein_robust']], protein_retain_mask)
```

## Summarisation with median or mean

We will also include summarisation with mean and median, since these are typical approaches for summarising across observations and will aid the interpretation when we compare the summasiation approaches. We will use the PSMs with missing values filtered, even though `mean` and `median` can handle missing values with `na.rm=TRUE`, since ignoring missing values is inappropriate for summarisation of PSMs to proteins with these methods.

```{r}

tmt_qf <- QFeatures::aggregateFeatures(tmt_qf, 
                                       i = "psms_filtered_forSum", 
                                       fcol = "Master.Protein.Accessions",
                                       name = "protein_mean",
                                       fun = base::colMeans, na.rm=TRUE)

tmt_qf <- QFeatures::aggregateFeatures(tmt_qf, 
                                       i = "psms_filtered_forSum", 
                                       fcol = "Master.Protein.Accessions",
                                       name = "protein_median",
                                       fun = matrixStats::colMedians, na.rm=TRUE)

tmt_qf[['protein_mean']] <- QFeatures::logTransform(
  tmt_qf[['protein_mean']], base=2)

tmt_qf[['protein_median']] <- QFeatures::logTransform(
  tmt_qf[['protein_median']], base=2)
```

## Comparing the summarisation approaches

We can see that performing the robust summarisation allows us to retain more proteins, since we can still use the PSMs with a low number of missing values.

```{r}
nrow(tmt_qf[['protein']])
nrow(tmt_qf[['protein_robust']])
```

However, with sum summarisation, we have complete quantification data (no missing values), whereas with `robustSummary`, we will proteins with missing values in some samples. In this case, the protein-level data is `r round(100*(1-mean(is.na(assay(tmt_qf[['protein_robust']])))), 2)` % complete, so the impact of the missing values on the downstream analysis will be minimal. We can inspect missing values using the `QFeatures::nNA` function, or just `is.na` on the quantitative data matrix.

```{r}
QFeatures::nNA(tmt_qf[['protein']])$nNAcols
QFeatures::nNA(tmt_qf[['protein_robust']])$nNAcols


message(sprintf('%.2f %% missing values', 100*mean(
  is.na(assay(tmt_qf[['protein_robust']])))))

```

To better understand the difference between how the two summarisation methods use the available data, we can inspect the single protein which has no missing values with `robustSummary`, but is not summarised with `sum`. 

```{r}
# Identify the proteins with no missing values present only following robustSummary
protein_robust_proteins <- rownames(filterNA(tmt_qf[['protein_robust']]))
protein_sum_proteins <- rownames(filterNA(tmt_qf[['protein']]))

no_missing_robustSummary_only <- setdiff(protein_robust_proteins, protein_sum_proteins)
print(no_missing_robustSummary_only)
```

If we inspect the PSM level abundances for this protein, we see that there are 3 PSMs, of which the first two contain missing values in a single sample. Thus, we have only one complete PSM, which is insufficient for confident summarisation using `sum`. However, all 3 PSMs have sufficiently few missing values to be used with the `robustSummary` method and all samples have at least two non missing values across the 3 PSMs, so this protein is quantified in all samples with `robustSummary`.

```{r}

retain_psms <- rowData(
  tmt_qf[['psms_filtered_sn']])$Master.Protein.Accessions==no_missing_robustSummary_only

assay(tmt_qf[['psms_filtered_sn']][retain_psms,])


```


We can directly compare the protein-level abundance estimates to explore where these two summarisation methods differ in their estimates.

```{r}
# Define a function to extract the protein abundances in long form and
# add a column annotating the method
get_long_form_prot_exp <- function(obj, method_name) {
  assay(obj) %>%
    data.frame() %>%
    tibble::rownames_to_column('protein') %>%
    pivot_longer(cols=-protein) %>%
    mutate(method = method_name)
}

# Single object with protein inference from both methods 
compare_protein_abundances <- rbind(
  mutate(longFormat(tmt_qf[['protein']]), method='Sum'),
  mutate(longFormat(tmt_qf[['protein_robust']]), method='Robust'),
  mutate(longFormat(tmt_qf[['protein_median']]), method='Median'),
  mutate(longFormat(tmt_qf[['protein_mean']]), method='Mean')
)


print(head(compare_protein_abundances))
```
We want to identify the proteins with the lowest correlation for abundances across the samples between summarisation methods so we can inspect the difference between them.

```{r}
proteins_of_interest <- compare_protein_abundances %>%
  pivot_wider(names_from = method, values_from = value) %>%
  group_by(rowname) %>%
  summarise(cor_sum_median=cor(Sum, Median),
            cor_sum_robust=cor(Sum, Robust)) %>%
  pivot_longer(cols=-rowname) %>%
  group_by(rowname) %>%
  summarise(min_cor=min(value)) %>%
  slice_min(min_cor, n=5) %>%
  pull(rowname)
```

Below, we define a function to plot the PSM and protein level abundances for a single protein.

```{r}
plot_pep_and_protein <- function(protein_of_interest) {
  
  to_plot_compare <- compare_protein_abundances %>% 
    filter(rowname == protein_of_interest)
  
  QFeatures::filterFeatures(
    tmt_qf,
    VariableFilter("Master.Protein.Accessions",
                   protein_of_interest,
                   condition = "=="))[['psms_filtered_sn']] %>%
    longFormat() %>%
    ggplot(aes(x = colname, y = log2(value))) +
    geom_line(aes(group = rowname), colour = 'grey') +
    geom_point(colour = 'grey') +
    geom_line(data = to_plot_compare,
              aes(x = colname, y = value, colour = method, group = method)) +
    #geom_point(data = to_plot_compare,  position=position_dodge(width=0.5),
    #           aes(x = colname, y = value, colour = method)) +
    scale_colour_manual(values = get_cat_palette(4),
                        name = 'Protein summarisation method') +
    theme_biomasslmb(base_size = 15, border = FALSE, base_family = 'sans') +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    labs(
      title = protein_of_interest,
      x = '',
      y = 'PSM/Protein abundance (log2)'
    ) +
    guides(color = guide_legend(override.aes = list(linewidth = 2) ) )
}
```


In the plots below, PSM level quantifications are in grey and the summarised protein-level abundances are in colours. You can disregard the y-axis scale when comparing between the methods, or between the protein and peptide level abundances. The important thing to focus on is the abundance profile between the tags for any one summarisation method or PSM.

```{r, plot.show='hold', out.width='45%', fig.width=7, fig.height=7, warning=FALSE, message=FALSE}
for(poi in proteins_of_interest) print(plot_pep_and_protein(poi))
```


In all cases, the protein-level quantification values are relatively similar for most tags, the differences between the approach typically restricted to one or two tags where the PSMs disagree on the relative abundances between the samples. 

Since `sum` simply adds together the abundance values across the PSMs for a given tag and the abundance values may be across multiple orders of magnitude, the protein-level abundance pattern across the tags is weighted towards the most abundant PSMs. This is likely to be a positive attribute, since the most abundant PSMs will be more accurately quantified. However, the summarisation is also sensitive to high abundance outliers. 

On the other hand, `robustSummary` is less sensitive to outliers, but it treats all PSMs equally. In the forth example (Q9NUQ9) we can see how this can lead to potential inaccuracies when we have only 2 PSMs and one is very low intensity. For example, tag 129C is mid-range intensity compared to the other tags for the abundant PSM, but very low for the less abundant PSM. This has almost no impact on the `sum` summarise protein abundance, but a large impact when we use `robustSummary`.

Note that mean is identical to sum with respect to the relative abundance across the samples, just on a difference scale, since mean is sum/n(PSMs). 

The median is less sensitive to outliers than the mean, but ignores much of the quantitative data since it's only affected by the 'middle' PSMs. In the most extreme instances, median just simplifies the PSM level abundances down to a single representative 'middle' PSM.

## Summary

`sum`/`mean` are more sensitive to the abundances for the highest intensity PSMs, which are likely to be the most accurately quantified. However, they are therefore also more sensitive to outlier values with very high abundance. `robustSummary` handles missing values and therefore enables more proteins to be quantified but it is more sensitive to low intensity PSMs, which may be less accurately quantified. For typical TMT datasets with very low missing values, `sum`/`mean` are likely to be the best approach. Where there are more missing values, consider using `robust` instead.

```{r}
sessionInfo()
```

