---
title: "TMT QC PSM-level quantification and summarisation to protein-level abundance"
author: "Tom Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: biomasslmb.json
vignette: >
  %\VignetteIndexEntry{TMT QC PSM-level quantification and summarisation to protein-level abundance}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Quantitative proteomics using isobaric tagging such as Tandem Mass Tags (TMT) has a considerable benefit over Label-Free Quantification (LFQ) in that up to 35 samples can be quantified for each Peptide Spectrum Match (PSM). This has multiple benefits over analysing samples in separate runs (LFQ):

1.  TMT reduces protein quantification variance since PSM-level quantification is derived from the same MS1 ion for all samples
2.  LFQ suffers from much higher missing values when comparing across samples due to the limited number of ions that can be fragmented in each run and the associated issue of peptides being identified in only a subset of runs [@http://zotero.org/users/5634351/items/ME322P2A]. This is ameliorated to a significant degree by Data-Independent Aquisition (DIA) LFQ. However, DIA still involves quantifying each sample separately, so missing values are not entirely removed, and proteins in each sample may be quantified from different sets of peptides.

Because TMT quantifies from the same MS1 ion for all samples, this standardises the features quantified in each sample, which simplifies the comparison between samples and increases quantification accuracy of summarised features such as proteins.

However, TMT does suffer from ratio compression, which should be avoiding by performing quantification with SPS MS3 [@http://zotero.org/users/5634351/items/ZK69WYZ2].

This vignette works through one typical experiment from end to end: a whole-proteome comparison between two conditions, labelled in a single TMT plex and searched with Proteome Discoverer (PD). That combination covers the majority of TMT experiments, and the choices made below are the conventional ones for it.

Experiments that depart from it are covered separately, since the departures change the reasoning rather than just the code:

- **Several plexes.** A design too large for one plex needs each plex processed on its own and then brought onto a common scale — see [multi-plex TMT](TMT_multiplex.html).
- **An enrichment design.** An IP, BioID or TurboID experiment breaks the assumptions behind normalisation and behind discarding missing values — see [enrichment designs](interactome_designs.html).
- **MaxQuant instead of PD.** The principles are the same, but the column names differ and the PD-specific QC metrics (signal:noise, interference, SPS-MM) are unavailable. The [enrichment designs](interactome_designs.html) vignette works through MaxQuant input.

# Load required packages

```{r, message=FALSE}
library(QFeatures)
library(biomasslmb)
library(ggplot2)
library(tidyr)
library(dplyr)
```

# The experimental design

`psm_tmt_clock` and `tmt_clock_design` are datasets available from the `biomasslmb` package, derived from a real TMT12plex experiment comparing `Control` and `Mutant` samples (6 replicates each). `psm_tmt_clock` is the PD PSM-level output, truncated to a subset of proteins for a manageable vignette. `tmt_clock_design` gives the experimental design: which sample each TMT tag corresponds to, and its `Condition` and `Replicate`.

```{r}
knitr::kable(tmt_clock_design)
```

Unlike a `data.frame` read from a single PD output file, real experiments nearly always have an associated design table like this (often as a spreadsheet from whoever ran the samples), which needs to be read in and used both to make sense of the TMT tags and to define experimental details for exploration and statistical testing.

# Read in input data

We start by reading in quantitative proteomics data into a `QFeatures` object, which is the standard Bioconductor object for holding quantitative proteomics data. See [here](https://www.bioconductor.org/packages/release/bioc/html/QFeatures.html) for documentation about the `QFeatures` object.

`psm_tmt_clock` contains the PSM-level output from PD for this experiment, with the abundance columns already renamed to match the sample names used in `tmt_clock_design`. We pass `tmt_clock_design` as the `colData`, so the experimental design travels with the quantification data.

```{r}

tmt_qf <- QFeatures::readQFeatures(assayData = psm_tmt_clock,
                                   colData = tmt_clock_design,
                                   quantCols = rownames(tmt_clock_design),
                                   name = "psms_raw")

# QFeatures::readQFeatures only attaches colData at the QFeatures level;
# we also want it directly on the PSM-level SummarizedExperiment so that
# functions which plot/model a single assay (e.g. plot_quant) can use it.
colData(tmt_qf[["psms_raw"]]) <- colData(tmt_qf)

```


# Defining the contaminant proteins

We need to remove contaminant proteins. This search was performed against the '0602_Universal Contaminants' database [@Frankenfield2022]. Below, we parse the contaminants fasta to extract the protein IDs, in the `Cont_<UniProt accession>` format PD reports them in. Since PD sometimes assigns the bare UniProt accession (without the `Cont_` prefix) as the master protein even for a contaminant hit, we match on both forms.

```{r}

contaminant_fasta_inf <- system.file(
  "extdata", "0602_Universal_Contaminants.fasta.gz",
  package = "biomasslmb"
)

# Extract the protein IDs associated with each contaminant protein
contaminant_accessions <- biomasslmb::get_contaminant_fasta_accessions(contaminant_fasta_inf)
contaminant_accessions <- c(contaminant_accessions, sub('^Cont_', '', contaminant_accessions))

print(head(contaminant_accessions))
```

We first perform routine filtering to remove PSMs that:

-   Could originate from contaminants. See `?filter_features_pd_dda` for further details, including the removal of 'associated' contaminants.
-   Don't have a unique master protein.

```{r}
# Perform routine raw data filtering.
# - Remove PSMs from contaminant proteins
# - Remove PSMs where protein ID is empty or not unique
tmt_qf[['psms_filtered']] <- filter_features_pd_dda(tmt_qf[['psms_raw']],
                                             contaminant_proteins=contaminant_accessions,
                                             filter_contaminant=TRUE,
                                             filter_associated_contaminant=TRUE,
                                             unique_master=TRUE)

colData(tmt_qf[['psms_filtered']]) <- colData(tmt_qf)
```

# Normalise

Next, we plot the peptide intensities to check they are approximately the same using `biomasslmb::plot_quant`. Colouring by `Condition` also lets us check the two conditions are not obviously different at this stage, before any normalisation or filtering.

```{r, warning=FALSE, fig.show='hold', fig.width=7, fig.height=5, out.width="75%"}

# Plot the peptide-level quantification distributions per sample
biomasslmb::plot_quant(tmt_qf[['psms_filtered']],
                       log2transform=TRUE,
                       method='density') +
  theme_biomasslmb() +
  aes(colour=Condition, group=sample) +
  xlab('PSM abundance (log2)')

```

Since the same amount of sample was labelled in each case, it's reasonable to use 'diff.median' normalisation with `QFeatures::normalize`. First though, we need to log-transform the abundances so the normalisation works as expected. We exponentiate the quantification values back to the original scale (reverse the log-transformation) afterwards, since the downstream steps including summarisation to protein require non-transformed values.

```{r, warning=FALSE, fig.show='hold', fig.width=7, fig.height=5, out.width="75%"}

# Normalise the log2-transformed abundances using diff.median
tmt_qf[['psms_filtered_norm']] <- QFeatures::normalize(
  logTransform(tmt_qf[['psms_filtered']], base=2), method='diff.median')

# Exponentiate the quantification values back to the initial scale.
assay(tmt_qf[['psms_filtered_norm']]) <- 2^assay(tmt_qf[['psms_filtered_norm']])

colData(tmt_qf[['psms_filtered_norm']]) <- colData(tmt_qf)

# Plot the peptide-level quantification distributions per sample
biomasslmb::plot_quant(tmt_qf[['psms_filtered_norm']],
                       log2transform=TRUE,
                       method='density') +
  theme_biomasslmb() +
  aes(colour=Condition, group=sample) +
  xlab('PSM abundance (log2)')

```

# Removing low quality PSMs

We want to remove low Signal:Noise (S:N) PSMs, since the quantification values will be less accurate and there will be more missing values. We can inspect the relationship between S:N and missing values using the `plot_missing_SN` function.

Note that where the signal:noise \> 10, there are far fewer missing values.

```{r, fig.height=6, fig.width=8, fig.fullwidth=TRUE, fig.cap="Missing values per PSM, in relation to the signal:noise ratio", out.width='75%'}
# Add a more accurate average S:N ratio value.
# The one calculated by PD doesn't treat NA values appropriately!
tmt_qf[['psms_filtered_norm']] <- update_average_sn(tmt_qf[['psms_filtered_norm']])

plot_missing_SN(tmt_qf[['psms_filtered_norm']], bins = 20)
```

We can also look into this relationship at the tag level using `plot_missing_SN_per_sample` to inspect whether there was a labeling issue with any particular tag. In this case, there is no tag which appears to have a high proportion of missing values when signal:noise \> 10.

```{r, fig.height=6, fig.width=8, fig.fullwidth=TRUE, fig.cap="Missing values per tag, in relation to the signal:noise ratio", out.width='75%'}
plot_missing_SN_per_sample(tmt_qf[['psms_filtered_norm']], bins = 20)
```

Based on the above, we will filter the PSMs to only retain those with S:N \> 10 using `filter_TMT_PSMs` since these PSMs are clearly much closer to the limit of detection for the MS and therefore likely to contain less accurate quantification data. Using the same function, we will also remove PSMs with interference/co-isolation \>50% since these are also likely to contain less accurate quantification data.

```{r}
# Then filter PSMs to remove low S:N and/or high interference
tmt_qf[['psms_filtered_sn']] <- filter_TMT_PSMs(tmt_qf[['psms_filtered_norm']],
                                                inter_thresh=50, sn_thresh=10)
colData(tmt_qf[['psms_filtered_sn']]) <- colData(tmt_qf)
```
We will also remove peptides that are not rank 1 according to the search engine.

```{r}

tmt_qf[['psms_filtered_rank']] <- tmt_qf[['psms_filtered_sn']]

tmt_qf <- tmt_qf %>%
  filterFeatures(~ Rank == 1, i = 'psms_filtered_rank')

biomasslmb:::message_parse(rowData(tmt_qf[['psms_filtered_rank']]),
                         'Master.Protein.Accessions',
                         "Removing peptides that are not rank 1")

```

# Summarising to protein-level abundances

Now that we have inspected the PSM-level quantification and filtered the PSMs, we can summarise the PSMs to protein-level abundances.

For PSM to protein summarisation with TMT quantification, simply summing together the PSM-level abundances provides accurate estimates, so long as there are no missing values. If many missing values are present, the `MsCoreUtils::robustSummary` method may be preferred, since it is able to summarise accurately, even in the presence of missing values [@http://zotero.org/users/5634351/items/FZN3QTTZ]. For a comparison of the two summarisation approaches, see the [Comparing protein summarisation approaches](summarisation_methods.html) vignette.

## Exploring missing values

Before deciding how to handle missing values, it's worth checking how much missingness there is, and whether it's random or structured by experimental condition. Condition-structured missingness (e.g. a PSM detected in every `Control` sample but no `Mutant` sample) can indicate a real biological or technical difference between conditions, and should make you wary of simply discarding those PSMs. Missingness scattered independently of condition, on the other hand, more likely reflects individual TMT channels whose intensity falls around the limit of detection: the S:N filter above is applied per-PSM, so a PSM can pass that threshold overall while one or two of its channels still drop out.

`plot_missing_upset` visualises the most common patterns of missingness across samples. Samples with no missing PSMs at all are omitted from the plot, since they don't contribute to any missingness pattern.

```{r, fig.height=6, fig.width=8, warning=FALSE, message=FALSE}
plot_missing_upset(tmt_qf, i = 'psms_filtered_rank')
```

`condition_miss_score` complements this by fitting, for each PSM, a logistic regression of missingness against a `colData` grouping column, and returning Tjur's R² as a per-feature score: values near 1 mean a PSM's missingness pattern is well explained by `Condition`, values near 0 mean it looks unrelated to `Condition`. `condition_miss_index` aggregates these per-feature scores into a single dataset-level value between 0 and 1.

```{r}
miss_res <- condition_miss_score(tmt_qf, i = 'psms_filtered_rank', group_cols = 'Condition')
condition_miss_index(miss_res$summary)$index
```

Only a small fraction of PSMs have any missing values at all (`r miss_res$global['n_features_informative']` out of `r miss_res$global['n_features_total']`), and the condition missingness index is close to zero, so we have no evidence that the missingness we do have is structured by `Condition`. On that basis, we'll proceed with the simplest approach: remove PSMs with any missing value, rather than reaching for `robustSummary` or imputation strategies designed to handle structured missingness (see Part B below for a dataset where that trade-off comes out differently).

We'll use the simple `sum` summarisation, which requires complete data: leaving missing values in would mean either allowing some protein quantifications to be `NA`, or summing over whatever PSMs happen to be present, so that a given protein's abundance is built from different PSMs in different samples and becomes less accurately comparable across them. We therefore remove all PSMs with missing values before summarising.

```{r}
tmt_qf[['psms_filtered_missing']] <- QFeatures::filterNA(
  tmt_qf[['psms_filtered_rank']], 0)

```

Next, we will remove PSMs for proteins with fewer than 2 PSMs. This is a common filter in proteomics to ensure the protein-level quantifications are derived from at least two independent observations. In some cases, for example phosphoproteomics, or where this filter appears to be too stringent, it may be appropriate to skip it.

```{r}
min_psms <- 2
tmt_qf[['psms_filtered_forSum']] <- biomasslmb::filter_features_per_protein(
  tmt_qf[['psms_filtered_missing']], min_features = min_psms)

```

Finally, we perform the summarisation.

```{r, message=FALSE}
# Aggregate to protein-level abundances (using QFeatures function)
tmt_qf <- QFeatures::aggregateFeatures(tmt_qf,
                                       i = "psms_filtered_forSum",
                                       fcol = "Master.Protein.Accessions",
                                       name = "protein",
                                       fun = base::colSums)

colData(tmt_qf[['protein']]) <- colData(tmt_qf)
```

```{r}
tmt_qf[['protein']] <- QFeatures::logTransform(
  tmt_qf[['protein']], base=2)
```

## Inspecting the number of PSMs and proteins through the processing steps

Now that we have performed all the filtering steps and summarisation to protein-level abundances, it's helpful to visualise how many PSMs/proteins were retained at each level of the processing. We can use the `biomasslmb::get_samples_present` and `biomasslmb::plot_samples_present` functions for this. First though, we need to decide which 'experiments' we want to plot and define a named character vector since the QFeatures names are not sufficiently clear by themselves.

Below, we inspect the experiment names.

```{r}
names(tmt_qf)
```

In this case, we don't want to plot the `psms_filtered_norm`, since that's just the normalised quantification present in `psms_filtered` and no PSMs were removed in that step.

### Samples per PSM

We'll start by inspecting the number of PSMs in each experiment. We therefore define a named character vector all the PSM-level experiments, excluding `psms_filtered_norm`. We set the row variables to be `Sequence`, `Modifications` and `RT.in.min` so that we count the number of unique PSMs.

```{r, fig.height=7, , fig.width=7, fig.fullwidth=TRUE, fig.cap="Samples quantified for each PSM at each level of processing", out.width='75%'}


rename_cols <- c('All PSMs' = 'psms_raw' ,
                 'Quantified, contaminants removed' = 'psms_filtered',
                 'Signal:Noise > 10' = 'psms_filtered_sn',
                 'PSMs - filtered (Rank1)' = 'psms_filtered_rank',
                 'No missing values' = 'psms_filtered_missing',
                 '>1 PSM per protein' = 'psms_filtered_forSum')

rowvars <- c('Sequence', 'Modifications', 'RT.in.min')

samples_present <- get_samples_present(tmt_qf[,,unname(rename_cols)], rowvars, rename_cols)
plot_samples_present(samples_present, rowvars, breaks=seq(2,12,2)) + ylab('PSM')

```

### Samples per Protein

Next, we'll use the same functions to inspect the number of proteins at each level of processing. We need to supply an updated named character vector to include the `protein` experiment and set the row variables to be just the `Master.Protein.Accesions` column.

```{r, fig.height=7, , fig.width=7, fig.fullwidth=TRUE, fig.cap="Samples quantified for each protein at each level of processing", out.width='75%'}


rename_cols_prot <- c(rename_cols, 'Protein'='protein')

rowvars_prot <- c('Master.Protein.Accessions')

samples_present <- get_samples_present(tmt_qf, rowvars_prot, rename_cols_prot)
plot_samples_present(samples_present, rowvars_prot, breaks=seq(2,12,2))
```

From these two plots, we can see that the filtering to ensure that all proteins have \>1 PSM removed only a few PSMs, but more proteins. Whether this is appropriate will depend on your data in hand.

```{r, eval = FALSE}
# Save file to package as data so it can be read back in in other vignettes
usethis::use_data(tmt_qf, overwrite=TRUE)
```

# Summary

Here, we have QCed the PSMs and filtered them in the following steps:

-   Remove PSMs that are likely from contaminants
-   Remove PSMs without a unique 'master protein'
-   Remove PSMs without any quantification values
-   Normalise the abundance values so they have the same median value in all samples, which is reasonable here because the same amount of material was labelled in every channel
-   Remove PSMs with very low signal:noise and high interference

We then summarised the PSM level abundances to protein level with `sum`, having first removed the small number of PSMs with any missing value. See the [Comparing protein summarisation approaches](summarisation_methods.html) vignette for a direct comparison against `robustSummary`, and the [Data exploration and statistical testing](exploration_and_statistical_testing.html) vignette for how to explore and statistically test the resulting protein-level data using the experimental design attached here.

Two of those steps are specific to this being a straightforward whole-proteome comparison in a single plex. Normalising to a common median assumes most proteins are unchanged, which an [enrichment design](interactome_designs.html) violates, and summing complete PSMs is only cheap because within-plex missingness is low — between plexes it is not, which is what [multi-plex TMT](TMT_multiplex.html) addresses.

```{r}
sessionInfo()
```
