---
title: "TMT QC PSM-level quantification and summarisation to protein-level abundance"
author: "Tom Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: biomasslmb.json
vignette: >
  %\VignetteIndexEntry{TMT QC PSM-level quantification and summarisation to protein-level abundance}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Quantitative proteomics using isobaric tagging such as Tandem Mass Tags (TMT) has a considerable benefit over Label-Free Quantification (LFQ) in that up to 35 samples can be quantified for each Peptide Spectrum Match (PSM). This has multiple benefits over analysing samples in separate runs (LFQ):

1.  TMT reduces protein quantification variance since PSM-level quantification is derived from the same MS1 ion for all samples
2.  LFQ suffers from much higher missing values when comparing across samples due to the limited number of ions that can be fragmented in each run and the associated issue of peptides being identified in only a subset of runs [@http://zotero.org/users/5634351/items/ME322P2A]. This is ameliorated to a significant degree by Data-Independent Aquisition (DIA) LFQ. However, DIA still involves quantifying each sample separately, so missing values are not entirely removed, and proteins in each sample may be quantified from different sets of peptides.

Because TMT quantifies from the same MS1 ion for all samples, this standardises the features quantified in each sample, which simplifies the comparison between samples and increases quantification accuracy of summarised features such as proteins.

However, TMT does suffer from ratio compression, which should be avoiding by performing quantification with SPS MS3 [@http://zotero.org/users/5634351/items/ZK69WYZ2].

Here, we will QC and filter the PSM level abundances from PD before summarising them to protein-level abundances.

# Load required packages

To clarify which functionality is provided by which package, we will use `package::function`. For your own code, there is no need to specify the package unless you want to maintain this clarity.

```{r, message=FALSE}
library(QFeatures)
library(biomasslmb)
library(ggplot2)
library(tidyr)
library(dplyr)
```

# Defining the contaminant proteins

We need to remove contaminant proteins. These were defined here using the cRAP database. Below, we parse the contaminants fasta to extract the IDs for the proteins in both 'cRAP' format and Uniprot IDs.

```{r}

crap_fasta_inf <- system.file(
  "extdata", "cRAP_20190401.fasta.gz", 
  package = "biomasslmb"
)

# Extract the protein IDs associated with each cRAP protein
crap_accessions <- biomasslmb::get_crap_fasta_accessions(crap_fasta_inf)

print(head(crap_accessions))
```

# Read in input data

We start by reading in quantitative proteomics data into a `QFeatures` object, which is the standard Bioconductor object for holding quantitative proteomics data. See [here](https://www.bioconductor.org/packages/release/bioc/html/QFeatures.html) for documentation about the `QFeatures` object.In this case, we are not adding any experimental details to the `QFeatures` object, so the `ColData` will be empty.

`psm_tmt_total` is a data set available from the `biomasslmb` package containing the PSM-level output from Proteome Discoverer (PD) for an experiment with 10 samples (each one being a separate TMT tags). It is a truncated file containing the first 5,000 PSMs only. Here, we will not include any details about the experimental design. Usually, these would be included by providing a `data.frame` to the `colData` argument. See `?readQFeatures` for full details on how to read the quantification data in a `QFeatures` object.

```{r}

tmt_qf <- QFeatures::readQFeatures(assayData = psm_tmt_total,
                                   quantCols = 36:45, 
                                   name = "psms_raw")



```

We have `r sum(assay(tmt_qf[['psms_raw']])==0, na.rm=TRUE)` quantification values which are exactly zero. These should be replaced with NA, since mass spectrometry is not capable of asserting that the protein had zero abundance in the sample.

```{r}
tmt_qf[['psms_raw']] <- QFeatures::zeroIsNA(tmt_qf[['psms_raw']])

```

We first perform routine filtering to remove PSMs that:

-   Could originate from contaminants. See `?filter_features_pd_dda` for further details, including the removal of 'associated cRAP'.
-   Don't have a unique master protein.

```{r}
# Perform routine raw data filtering.
# - Remove PSMs from contaminant proteins
# - Remove PSMs where protein ID is empty or not unique
tmt_qf[['psms_filtered']] <- filter_features_pd_dda(tmt_qf[['psms_raw']],
                                             contaminant_proteins=crap_accessions,
                                             filter_contaminant=TRUE,
                                             filter_associated_contaminant=TRUE,
                                             unique_master=TRUE, )
```


# Normalise

Next, we plot the peptide intensities to check they are approximately the same using `biomasslmb::plot_quant`.

```{r, warning=FALSE, fig.show='hold', fig.width=7, fig.height=5, out.width="75%"}

# Plot the peptide-level quantification distributions per sample
biomasslmb::plot_quant(tmt_qf[['psms_filtered']],
                       log2transform=TRUE,
                       method='density') +
  theme_biomasslmb() +
  xlab('PSM abundance (log2)')

```

Since the same amount of sample was labelled in each case, it's reasonable to use 'diff.median' normalisation with `QFeatures::normalize`. First though, we need to log-transform the abundances so the normalisation works as expected. We exponentiate the quantification values back to the original scale (reverse the log-transformation) afterwards, since the downstream steps including summarisation to protein require non-transformed values

```{r, warning=FALSE, fig.show='hold', fig.width=7, fig.height=5, out.width="75%"}

# Normalise the log2-transformed abundances using diff.median
tmt_qf[['psms_filtered_norm']] <- QFeatures::normalize(
  logTransform(tmt_qf[['psms_filtered']], base=2), method='diff.median') 

# Exponentiate the quantification values back to the initial scale.
assay(tmt_qf[['psms_filtered_norm']]) <- 2^assay(tmt_qf[['psms_filtered_norm']])

# Plot the peptide-level quantification distributions per sample
biomasslmb::plot_quant(tmt_qf[['psms_filtered_norm']],
                       log2transform=TRUE,
                       method='density') +
  theme_biomasslmb() +
  xlab('PSM abundance (log2)')

```

# Removing low quality PSMs

We want to remove low Signal:Noise (S:N) PSMs, since the quantification values will be less accurate and there will be more missing values. We can inspect the relationship between S:N and missing values using the `plot_missing_SN` function.

Note that where the signal:noise \> 10, there are far fewer missing values.

```{r, fig.height=6, fig.width=8, fig.fullwidth=TRUE, fig.cap="Missing values per PSM, in relation to the signal:noise ratio", out.width='75%'}
# Add a more accurate average S:N ratio value.
# The one calculated by PD doesn't treat NA values appropriately!
tmt_qf[['psms_filtered_norm']] <- update_average_sn(tmt_qf[['psms_filtered_norm']])

plot_missing_SN(tmt_qf[['psms_filtered_norm']], bins = 20)
```

We can also look into this relationship at the tag level using `plot_missing_SN_per_sample` to inspect whether there was a labeling issue with any particular tag. In this case, there is no tag which appears to have a high proportion of missing values when signal:noise \> 10.

```{r, fig.height=6, fig.width=8, fig.fullwidth=TRUE, fig.cap="Missing values per tag, in relation to the signal:noise ratio", out.width='75%'}
plot_missing_SN_per_sample(tmt_qf[['psms_filtered_norm']], bins = 20)
```

Based on the above, we will filter the PSMs to only retain those with S:N \> 10 using `filter_TMT_PSMs` since these PSMs are clearly much closer to the limit of detection for the MS and therefore likely to contain less accurate quantification data. Note that this only necessitates removing \~ 1/20 of the PSMs in this case. Using the same function, we will also remove PSMs with interference/co-isolation \>50% since these are also likely to contain less accurate quantification data.

```{r}
# Then filter PSMs to remove low S:N and/or high interference
tmt_qf[['psms_filtered_sn']] <- filter_TMT_PSMs(tmt_qf[['psms_filtered_norm']],
                                                inter_thresh=50, sn_thresh=10)
```
We will also remove peptides that are not rank 1 according to the search engine

```{r}

tmt_qf[['psms_filtered_rank']] <- tmt_qf[['psms_filtered_sn']]

tmt_qf <- tmt_qf %>%
  filterFeatures(~ Rank == 1, i = 'psms_filtered_rank')

biomasslmb:::message_parse(rowData(tmt_qf[['psms_filtered_rank']]),
                         'Master.Protein.Accessions',
                         "Removing peptides that are not rank 1")




```

# Summarising to protein-level abundances

Now that we have inspected the PSM-level quantification and filtered the PSMs, we can summarise the PSMs to protein-level abundances.

For PSM to protein summarisation with TMT quantification, simply summing together the PSM-level abundances provides accurate estimates, so long as there are no missing values. If many missing values are present, the `MsCoreUtils::robustSummary` method may be preferred, since it is able to summarise accurately, even in the presence of missing values [@http://zotero.org/users/5634351/items/FZN3QTTZ]. For a comparison of the two summarisation approaches, see the `TMT - comparing protein summarisation approaches` vignette.

Here, we will use the simple sum summarisation. First, we need to consider the missing values. If we leave the missing values in, we either need to allow some protein quantifications to be `NA`, or else ignore the missing values in the summarisation. The latter would potentially mean different PSMs are summed for the a given protein across the samples, making a given protein's summarised abundances across the samples less accurately estimated. Here, we will therefore remove all PSMs with missing values.

```{r}
tmt_qf[['psms_filtered_missing']] <- QFeatures::filterNA(
  tmt_qf[['psms_filtered_rank']], 0)

```

Next, we will remove PSMs for proteins with fewer than 2 PSMs. This is a common filter in proteomics to ensure the protein-level quantifications are derived from at least two independent observations. In some cases, for example phosphoproteomics, or where this filter appears to be too stringent, it may be appropriate to skip it.

```{r}
min_psms <- 2
tmt_qf[['psms_filtered_forSum']] <- biomasslmb::filter_features_per_protein(
  tmt_qf[['psms_filtered_missing']], min_features = min_psms)

```

Finally, we perform the summarisation.

```{r}
# Aggregate to protein-level abundances (using QFeatures function)
tmt_qf <- QFeatures::aggregateFeatures(tmt_qf, 
                                       i = "psms_filtered_forSum", 
                                       fcol = "Master.Protein.Accessions",
                                       name = "protein",
                                       fun = base::colSums)

```

```{r}
tmt_qf[['protein']] <- QFeatures::logTransform(
  tmt_qf[['protein']], base=2)
```

## Inspecting the number of PSMs and proteins through the processing steps

Now that we have performed all the filtering steps and summarisation to protein-level abundances, it's helpful to visualise how many PSMs/proteins were retained at each level of the processing. We can use the `biomasslmb::get_samples_present` and `biomasslmb::plot_samples_present` functions for this. First though, we need to decide which 'experiments' we want to plot and define a named character vector since the QFeatures names are not sufficiently clear by themselves

Below, we inspect the experiment names.

```{r}
names(tmt_qf)
```

In this case, we don't want to plot the `psms_filterered_norm`, since that's just the normalised quantification present in `psms_filtered` and no PSMs were removed in that step.

### Samples per PSM

We'll start by inspecting the number of PSMs in each experiment. We therefore define a named character vector all the PSM-level experiments, excluding `psms_filtered_norm`. We set the row variables to be `Sequence`, `Modifications` and `RT.in.min` so that we count the number of unique PSMs.

```{r, fig.height=7, , fig.width=7, fig.fullwidth=TRUE, fig.cap="Samples quantified for each PSM at each level of processing", out.width='75%'}


rename_cols <- c('All PSMs' = 'psms_raw' ,
                 'Quantified, contaminants removed' = 'psms_filtered',
                 'Signal:Noise > 10' = 'psms_filtered_sn',
                 'PSMs - filtered (Rank1)' = 'psms_filtered_rank',
                 'No missing values' = 'psms_filtered_missing',
                 '>1 PSM per protein' = 'psms_filtered_forSum')

rowvars <- c('Sequence', 'Modifications', 'RT.in.min')

samples_present <- get_samples_present(tmt_qf[,,unname(rename_cols)], rowvars, rename_cols)
plot_samples_present(samples_present, rowvars, breaks=seq(2,10,2)) + ylab('PSM')


   
```

### Samples per Protein

Next, we'll use the same functions to inspect the number of proteins at each level of processing. We need to supply an updated named character vector to include the `protein` experiment and set the row variables to be just the `Master.Protein.Accesions` column.

```{r, fig.height=7, , fig.width=7, fig.fullwidth=TRUE, fig.cap="Samples quantified for each protein at each level of processing", out.width='75%'}


rename_cols_prot <- c(rename_cols, 'Protein'='protein')

rowvars_prot <- c('Master.Protein.Accessions')

samples_present <- get_samples_present(tmt_qf, rowvars_prot, rename_cols_prot)
plot_samples_present(samples_present, rowvars_prot, breaks=seq(2,10,2))
```

In this case, because we have removed all missing values, the last two experiments have identical counts for the number of samples for each protein. From these two plots, we can see that the filtering to ensure that all proteins have \>1 PSM removed only a few PSMs, but many proteins. Whether this is appropriate will depend on your data in hand.

```{r, include = FALSE}
# Save file to package as data so it can be read back in in other vignettes
usethis::use_data(tmt_qf, overwrite=TRUE)
```

## Summary

Here, we have QCed the PSMs and filtered them in the following steps:

-   Remove PSMs that are likely from contaminants
-   Remove PSMs without a unique 'master protein'
-   Remove PSMs without any quantification values
-   Normalise the abundance values so they have the same median value in all samples
-   Remove PSMs with very low signal:noise
-   Remove PSMs with high interference

We then summarised the PSM level abundances with two approaches, `sum` and `robustSummary`, before comparing the protein-level abundances obtained with each of them. `robustSummary` handles missing values and therefore enables more proteins to be quantified. `sum` is sensitive to outliers with very high abundance, while `robust` is more sensitive to outliers with very low abundance.

```{r}
sessionInfo()
```
