---
title: "TMT data exploration and statistical testing"
author: "Tom Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: biomasslmb.json
vignette: >
  %\VignetteIndexEntry{TMT data exploration and statistical testing}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Once protein-level abundances have been QCed, filtered and summarised (see the `TMT QC PSM-level quantification and summarisation to protein-level abundance` vignette), the next step is to explore the data with respect to the experimental design, then test for differential abundance between conditions. Below, we work through that process twice, using the same two datasets as the QC vignette: first with `tmt_qf` (Part A), a whole-proteome comparison of `Control` and `Mutant` samples; then with `tmt_qf_mq` (Part B), an immunoprecipitation comparing a bait pulldown (`IP`) against a `Control` pulldown. The two designs call for different choices at the statistical testing stage, which we highlight in Part B.

# Load required packages

```{r, message=FALSE}
library(QFeatures)
library(biomasslmb)
library(ggplot2)
library(tidyr)
library(dplyr)
library(limma)
```

# Part A: Proteome Discoverer (PD) input

```{r}
names(tmt_qf)
knitr::kable(data.frame(colData(tmt_qf)))
```

## Sample correlation

A pairwise correlation heatmap is a quick way to check that replicates agree with each other more than they agree with samples from the other condition. `plot_cor_samples` computes the Spearman correlation between samples for a given assay and plots it as a heatmap.

```{r, fig.height=5, fig.width=5}
plot_cor_samples(tmt_qf, 'protein')
```

## Principal component analysis

`plot_pca` performs a PCA on a given assay and plots two of the components, optionally colouring points by a `colData` column. Since we removed all missing values in the QC vignette, we can use the default `allowing_missing=FALSE`, which uses `stats::prcomp` directly.

```{r, fig.height=4, fig.width=5, out.width='60%'}
plot_pca(tmt_qf, 'protein', colour_by='Condition', scale=TRUE, center=TRUE) +
  theme_biomasslmb(base_size=10)
```

Control and Mutant samples separate clearly along a principal component, consistent with the correlation heatmap above.

## Missingness and experimental condition

`sum` summarisation (used for `tmt_qf` in the QC vignette) requires complete PSMs, so the final `protein` assay has no missing values. It's still useful to check whether missingness *earlier* in the pipeline, before that filtering, was random or structured by condition — this can flag labelling or handling problems with specific samples, and is essential context when choosing whether `robustSummary`-style summarisation (which tolerates missing values) is appropriate for a new dataset.

`condition_miss_score` fits a per-feature logistic regression of missingness against condition and returns Tjur's R² per feature: values near 1 mean a feature's missingness is well explained by condition, values near 0 mean it looks unrelated to condition. `condition_miss_index` summarises this into a single dataset-level value.

```{r}
miss_res <- condition_miss_score(tmt_qf, i = 'psms_filtered_rank', group_cols = 'Condition')
miss_idx <- condition_miss_index(miss_res$summary)

miss_idx$index
```

The index is close to zero, and only a small fraction of PSMs had any missing values at all (`r miss_res$global['n_features_informative']` out of `r miss_res$global['n_features_total']`), so we have no evidence that missingness in this experiment is structured by condition. That is consistent with our choice to simply remove PSMs with any missing value, rather than reaching for `robustSummary` or imputation.

## Differential abundance testing with limma

We use `limma` for differential abundance testing between `Control` and `Mutant`. `limma` fits a linear model per protein, then applies an empirical Bayes step which borrows information across all proteins to stabilise the variance estimate — important here since 6 replicates per condition gives fairly noisy per-protein variance estimates on their own.

We use `treat` rather than plain `eBayes`. `treat` tests the null hypothesis that the absolute fold-change is *below* a threshold (here 1.1-fold), rather than the standard null of exactly no difference — this avoids flagging changes that are statistically detectable but too small to be biologically interesting. `trend=TRUE` lets the variance prior depend on mean abundance; `robust=TRUE` down-weights proteins with unusually extreme variance so they don't distort the shared prior.

```{r, fig.show='hide'}
condition <- factor(tmt_qf[['protein']]$Condition, levels = c('Control', 'Mutant'))
design <- model.matrix(~condition)

fit <- lmFit(assay(tmt_qf[['protein']]), design)
final_model <- treat(fit, fc = 1.1, trend = TRUE, robust = TRUE)

limma_results <- topTreat(final_model, coef = 'conditionMutant', number = Inf) %>%
  tibble::rownames_to_column('Protein') %>%
  arrange(P.Value)

table(direction = ifelse(limma_results$logFC < 0, 'Dec.', 'Inc.'),
      sig = limma_results$adj.P.Val < 0.05)
```

Note that, since this vignette works with a small subset of the proteins detected in the full experiment (see the QC vignette), the multiple-testing burden — and hence the FDR-adjusted p-values — will differ from an analysis of the complete dataset.

## Volcano plot

`plot_volcano` plots log-fold-change against -log10(p-value) directly from a statistical testing results `data.frame`; by default it expects `logFC` and `adj.P.Val` columns, which is exactly what `topTreat` returns.

```{r, fig.width=6, fig.height=6}
sig_hits <- limma_results %>% filter(adj.P.Val < 0.05) %>% pull(Protein)

plot_volcano(limma_results, sig_col = NULL) +
  geom_point(data = filter(limma_results, Protein %in% sig_hits),
             aes(fill = adj.P.Val < 0.05), pch = 21, size = 3) +
  scale_fill_manual(values = get_cat_palette(2)[2], guide = 'none') +
  ggrepel::geom_text_repel(
    data = filter(limma_results, Protein %in% sig_hits),
    aes(label = Protein), min.segment.length = 0, size = 3)
```

## Inspecting proteins of interest

Before trusting a hit, it's worth looking at the abundance values behind it across the processing pipeline — from PSM level through to the final protein-level estimate — to confirm the result isn't being driven by a single outlying PSM. `plot_protein_assays` plots one or more proteins across any set of assays in the `QFeatures` object.

```{r, fig.height=6, fig.width=8, warning=FALSE, message=FALSE}
top_hit <- head(sig_hits, 1)

plot_protein_assays(
  tmt_qf, top_hit,
  experiments_to_plot = c('psms_raw', 'psms_filtered_norm', 'protein'),
  protein_id_col = 'Master.Protein.Accessions',
  label_col = 'Master.Protein.Accessions',
  log2transform_cols = c('psms_raw')) +
  aes(colour = colData(tmt_qf)[colname, 'Condition']) +
  scale_colour_manual(values = get_cat_palette(2), name = 'Condition') +
  theme(aspect.ratio = 1/3)
```

The PSM-level trend is consistent across the processing steps and across the individual PSMs contributing to the protein, giving confidence that this is a genuine change rather than an artefact of summarisation.

# Part B: MaxQuant (MQ) input

`tmt_qf_mq` is the immunoprecipitation dataset from the QC vignette: a bait pulldown (`IP`) against a control pulldown (`Control`), 6 replicates each, summarised to protein level with `robustSummary`.

```{r}
names(tmt_qf_mq)
knitr::kable(data.frame(colData(tmt_qf_mq)))
```

## Sample correlation

```{r, fig.height=5, fig.width=5}
plot_cor_samples(tmt_qf_mq, 'protein')
```

## Principal component analysis

Unlike Part A, `robustSummary` doesn't require complete PSMs, so `tmt_qf_mq[['protein']]` retains a small number of missing values (`r round(100*mean(is.na(assay(tmt_qf_mq[['protein']]))), 2)` % of the matrix). `stats::prcomp` can't handle missing values, so instead of the default `allowing_missing=FALSE` (which would silently drop every protein with any missing value via `filterNA`), we set `allowing_missing=TRUE`, which uses `pcaMethods::pca` instead. Note that `scale`/`center` are passed on to whichever underlying PCA function is used, so with `pcaMethods::pca` `scale` takes a scaling method name (`'uv'`, unit-variance, is the equivalent of `scale=TRUE` in `stats::prcomp`) rather than `TRUE`/`FALSE`.

```{r, fig.height=4, fig.width=5, out.width='60%'}
plot_pca(tmt_qf_mq, 'protein', colour_by='Condition', scale='uv', center=TRUE, allowing_missing=TRUE) +
  theme_biomasslmb(base_size=10)
```

IP and Control samples separate clearly, as expected for a specific enrichment.

## Differential abundance testing with limma

The statistical model itself is the same as Part A, but the biological question is different, and that changes what fold-change threshold is appropriate. In Part A, we were looking for subtle shifts in a whole-proteome comparison, so a modest 1.1-fold `treat` threshold made sense. Here, we're trying to identify bona fide interactors of the bait protein: a real interactor should be substantially enriched in the IP relative to the control, not just detectably different. We therefore use a stricter 2-fold threshold.

```{r, fig.show='hide'}
condition_mq <- factor(tmt_qf_mq[['protein']]$Condition, levels = c('Control', 'IP'))
design_mq <- model.matrix(~condition_mq)

fit_mq <- lmFit(assay(tmt_qf_mq[['protein']]), design_mq)
final_model_mq <- treat(fit_mq, fc = 2, trend = TRUE, robust = TRUE)

limma_results_mq <- topTreat(final_model_mq, coef = 'condition_mqIP', number = Inf) %>%
  tibble::rownames_to_column('Protein') %>%
  arrange(P.Value)

table(direction = ifelse(limma_results_mq$logFC < 0, 'Dec.', 'Inc.'),
      sig = limma_results_mq$adj.P.Val < 0.05)
```

Both enriched and depleted proteins can be statistically significant here, but for identifying interactors we care specifically about enrichment in the IP: a depleted protein doesn't indicate an interaction with the bait, so below we restrict to the significantly *enriched* proteins as our candidate interactors.

```{r}
sig_hits_mq <- limma_results_mq %>%
  filter(adj.P.Val < 0.05, logFC > 0) %>%
  pull(Protein)

length(sig_hits_mq)
```

## Volcano plot

With `r length(sig_hits_mq)` candidate interactors, labelling every point would make the plot unreadable, so we label only the 10 most significant.

```{r, fig.width=6, fig.height=6}
top_hits_mq <- head(sig_hits_mq, 10)

plot_volcano(limma_results_mq, sig_col = NULL) +
  geom_point(data = filter(limma_results_mq, Protein %in% sig_hits_mq),
             aes(fill = adj.P.Val < 0.05), pch = 21, size = 3) +
  scale_fill_manual(values = get_cat_palette(2)[2], guide = 'none') +
  ggrepel::geom_text_repel(
    data = filter(limma_results_mq, Protein %in% top_hits_mq),
    aes(label = Protein), min.segment.length = 0, size = 3)
```

## Inspecting proteins of interest

As in Part A, we inspect the PSM-to-protein trend behind the top candidate interactor, using `Leading.razor.protein` in place of `Master.Protein.Accessions` and `psms_filtered_interference` in place of `psms_filtered_norm`.

```{r, fig.height=6, fig.width=8, warning=FALSE, message=FALSE}
top_hit_mq <- head(sig_hits_mq, 1)

plot_protein_assays(
  tmt_qf_mq, top_hit_mq,
  experiments_to_plot = c('psms_raw', 'psms_filtered_interference', 'protein'),
  protein_id_col = 'Leading.razor.protein',
  label_col = 'Leading.razor.protein',
  log2transform_cols = c('psms_raw')) +
  aes(colour = colData(tmt_qf_mq)[colname, 'Condition']) +
  scale_colour_manual(values = get_cat_palette(2), name = 'Condition') +
  theme(aspect.ratio = 1/3)
```

The PSMs consistently show higher abundance in the IP samples than the Control samples, supporting this as a genuine interactor rather than an artefact of a single PSM or a missing-value pattern.

# Summary

Here, we have, for both a whole-proteome comparison (Part A) and an immunoprecipitation (Part B):

-   Checked that replicate samples correlate more strongly with each other than with the other condition, and that PCA separates the conditions — using `allowing_missing=TRUE` in Part B, since `robustSummary` retains some missing values
-   Tested for differential protein abundance with `limma::treat`, using a fold-change threshold appropriate to the biological question: a modest threshold to detect subtle whole-proteome changes in Part A, and a strict threshold to identify substantially-enriched candidate interactors in Part B
-   Visualised the results with a volcano plot and inspected the PSM-to-protein trend for a top hit

Part A also checked whether missingness earlier in the pipeline was structured by experimental condition; see the QC vignette for the equivalent check on the Part B dataset, where missingness turned out to be much more prevalent and informative.

```{r}
sessionInfo()
```
