---
title: "TMT data exploration and statistical testing"
author: "Tom Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: biomasslmb.json
vignette: >
  %\VignetteIndexEntry{TMT data exploration and statistical testing}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Once protein-level abundances have been QCed, filtered and summarised (see the `TMT QC PSM-level quantification and summarisation to protein-level abundance` vignette), the next step is to explore the data with respect to the experimental design, then test for differential abundance between conditions. Here, we work through that process using `tmt_qf`, the `QFeatures` object produced by that vignette: a TMT12plex experiment comparing `Control` and `Mutant` samples (6 replicates each).

# Load required packages

```{r, message=FALSE}
library(QFeatures)
library(biomasslmb)
library(ggplot2)
library(tidyr)
library(dplyr)
library(limma)
```

```{r}
names(tmt_qf)
knitr::kable(data.frame(colData(tmt_qf)))
```

# Sample correlation

A pairwise correlation heatmap is a quick way to check that replicates agree with each other more than they agree with samples from the other condition. `plot_cor_samples` computes the Spearman correlation between samples for a given assay and plots it as a heatmap.

```{r, fig.height=5, fig.width=5}
plot_cor_samples(tmt_qf, 'protein')
```

# Principal component analysis

`plot_pca` performs a PCA on a given assay and plots two of the components, optionally colouring points by a `colData` column. Since we removed all missing values in the QC vignette, we can use the default `allowing_missing=FALSE`, which uses `stats::prcomp` directly.

```{r, fig.height=4, fig.width=5, out.width='60%'}
plot_pca(tmt_qf, 'protein', colour_by='Condition', scale=TRUE, center=TRUE) +
  theme_biomasslmb(base_size=10)
```

Control and Mutant samples separate clearly along a principal component, consistent with the correlation heatmap above.

# Missingness and experimental condition

`sum` summarisation (used for `tmt_qf` in the QC vignette) requires complete PSMs, so the final `protein` assay has no missing values. It's still useful to check whether missingness *earlier* in the pipeline, before that filtering, was random or structured by condition — this can flag labelling or handling problems with specific samples, and is essential context when choosing whether `robustSummary`-style summarisation (which tolerates missing values) is appropriate for a new dataset.

`condition_miss_score` fits a per-feature logistic regression of missingness against condition and returns Tjur's R² per feature: values near 1 mean a feature's missingness is well explained by condition, values near 0 mean it looks unrelated to condition. `condition_miss_index` summarises this into a single dataset-level value.

```{r}
miss_res <- condition_miss_score(tmt_qf, i = 'psms_filtered_rank', group_cols = 'Condition')
miss_idx <- condition_miss_index(miss_res$summary)

miss_idx$index
```

The index is close to zero, and only a small fraction of PSMs had any missing values at all (`r miss_res$global['n_features_informative']` out of `r miss_res$global['n_features_total']`), so we have no evidence that missingness in this experiment is structured by condition. That is consistent with our choice to simply remove PSMs with any missing value, rather than reaching for `robustSummary` or imputation.

# Differential abundance testing with limma

We use `limma` for differential abundance testing between `Control` and `Mutant`. `limma` fits a linear model per protein, then applies an empirical Bayes step which borrows information across all proteins to stabilise the variance estimate — important here since 6 replicates per condition gives fairly noisy per-protein variance estimates on their own.

We use `treat` rather than plain `eBayes`. `treat` tests the null hypothesis that the absolute fold-change is *below* a threshold (here 1.1-fold), rather than the standard null of exactly no difference — this avoids flagging changes that are statistically detectable but too small to be biologically interesting. `trend=TRUE` lets the variance prior depend on mean abundance; `robust=TRUE` down-weights proteins with unusually extreme variance so they don't distort the shared prior.

```{r, fig.show='hide'}
condition <- factor(tmt_qf[['protein']]$Condition, levels = c('Control', 'Mutant'))
design <- model.matrix(~condition)

fit <- lmFit(assay(tmt_qf[['protein']]), design)
final_model <- treat(fit, fc = 1.1, trend = TRUE, robust = TRUE)

limma_results <- topTreat(final_model, coef = 'conditionMutant', number = Inf) %>%
  tibble::rownames_to_column('Protein') %>%
  arrange(P.Value)

table(direction = ifelse(limma_results$logFC < 0, 'Dec.', 'Inc.'),
      sig = limma_results$adj.P.Val < 0.05)
```

Note that, since this vignette works with a small subset of the proteins detected in the full experiment (see the QC vignette), the multiple-testing burden — and hence the FDR-adjusted p-values — will differ from an analysis of the complete dataset.

# Volcano plot

`plot_volcano` plots log-fold-change against -log10(p-value) directly from a statistical testing results `data.frame`; by default it expects `logFC` and `adj.P.Val` columns, which is exactly what `topTreat` returns.

```{r, fig.width=6, fig.height=6}
sig_hits <- limma_results %>% filter(adj.P.Val < 0.05) %>% pull(Protein)

plot_volcano(limma_results, sig_col = NULL) +
  geom_point(data = filter(limma_results, Protein %in% sig_hits),
             aes(fill = adj.P.Val < 0.05), pch = 21, size = 3) +
  scale_fill_manual(values = get_cat_palette(2)[2], guide = 'none') +
  ggrepel::geom_text_repel(
    data = filter(limma_results, Protein %in% sig_hits),
    aes(label = Protein), min.segment.length = 0, size = 3)
```

# Inspecting proteins of interest

Before trusting a hit, it's worth looking at the abundance values behind it across the processing pipeline — from PSM level through to the final protein-level estimate — to confirm the result isn't being driven by a single outlying PSM. `plot_protein_assays` plots one or more proteins across any set of assays in the `QFeatures` object.

```{r, fig.height=6, fig.width=8, warning=FALSE, message=FALSE}
top_hit <- head(sig_hits, 1)

plot_protein_assays(
  tmt_qf, top_hit,
  experiments_to_plot = c('psms_raw', 'psms_filtered_norm', 'protein'),
  protein_id_col = 'Master.Protein.Accessions',
  label_col = 'Master.Protein.Accessions',
  log2transform_cols = c('psms_raw')) +
  aes(colour = colData(tmt_qf)[colname, 'Condition']) +
  scale_colour_manual(values = get_cat_palette(2), name = 'Condition') +
  theme(aspect.ratio = 1/3)
```

The PSM-level trend is consistent across the processing steps and across the individual PSMs contributing to the protein, giving confidence that this is a genuine change rather than an artefact of summarisation.

## Summary

Here, we have:

-   Checked that replicate samples correlate more strongly with each other than with the other condition, and that PCA separates the conditions
-   Checked whether missingness earlier in the pipeline was structured by experimental condition
-   Tested for differential protein abundance between conditions with `limma::treat`
-   Visualised the results with a volcano plot and inspected the PSM-to-protein trend for a top hit

```{r}
sessionInfo()
```
